---
title: "Central Limit Theorem" 
author:
  name: "Carsten Lange"
  email: "clange@cpp.edu"
  affiliation: "Cal Poly, Pomona"

execute:
  message: false
  warning: false

format: 
  live-revealjs:
    theme: [moon,../../CustomCL.scss]
    warning: false
    controls: true
    chalkboard:
      theme: whiteboard
      boardmarker-width: 2.7
    incremental: false
    scrollable: true
    footer: "<a href='https://ai.lange-analytics.com/htmlbook/RAndRStudio.html'>Textbook </a> "
webr:
  packages:
    - tidyverse
    - rio
    - plotly
    - TeachHist
    - ggdist
    

engine: knitr    
---

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}


## What Will You Learn {.scrollable .smaller}

```{r}
#| echo: false
library(TeachHist)

```

- You will learn how to work with sample measurements (called: "statistics")

- You learned already to calculate these *statistics*, such as the sample mean

- You will learn that *statistics* vary with different samples and that you can estimate the *Standard Error* even when you can see only one sample

-   You will distinguish between *Standard Deviation* and *Standard Error* and learn how to calculate the *Standard Error*

-   You will distinguish between *Normal Distribution* and *t-Distribution*

-   You will calculate a *Confidence Interval* for a mean based on its *Standard Error*

-   You will learn that means from multiple samples are *normal distributed* even when the original data are not *normal distributed*



## Histogram Cholesterol with Density as Column Height and Rel. Freq./Probability as Column Area plus Normal Curve  {.smaller}

```{r}
#| echo: false
TeachHistDens(Mean=230.5, Sd=50, SeedValue=123, NOfSimData=10000, VLine1 = 360, PlotNormCurv = TRUE)
```

`pnorm(x, mean=?, sd=?)` calculates the probability of getting a value smaller than x. <br> **Try:** Probability of a value of smaller than 360 when the mean is 230.5 and the standard deviation is 50


```{webr}
pnorm(@@@,@@@,@@@)
```


## Research about Groups (Samples) is More Important than Individuals {.smaller}

A survey of 50 customers from a branch leads to average sentiment score of 72 (scale 0 - 100). It is known that the Mean for all customers is $70$ and the Standard Deviation is $3$.

. . .

**The mean of the sample is different than the mean of all customers** This should not come as surprise. If we had used another sample, we would have measured a **different** mean.

. . .

**So, by how much would the *Sample Mean* spread, if we considered several surveys each with $N=50$?**
<br>
(Note, the spread of a sample mean is called *Standard Error*)

. . .

-   **Standard Error is less than the Standard Deviation of the population**. <br> 
**Reason:** In a survey (sample), customers with a high and low sentiment score cancel out.

. . .



. . .

-   The *Standard Error* is smaller when the sample size ($N$) is bigger.






. . .

-   The *Standard Error* is bigger when the *Standard Deviation* of the population is bigger.



. . .

$$SE=\frac{StdDev}{\sqrt{N}}$$

## Experiment: Heigths in Inches in this Course (CL: backup student heights!)

**Provide Data for 3 Samples (A, B, C):**

```{webr}
#| autorun: true
DataCourseHeight=tribble(~Sample, ~Height,
                           "A", 71,
                           "A", 68,
                           "A", 68,
                           "A", 66,
                           "A", 77,
                           "B", 69,
                           "B", 72,
                           "B", 74, 
                           "B", 70,
                           "B", 72,
                           "C", 70,
                           "C", 69,
                           "C", 72,
                           "C", 75,
                           "C", 68)
```

## Standard Deviation for All

```{webr}
SdHeightAll=sd(DataCourseHeight$Height)
MeanHeightAll=mean(DataCourseHeight$Height)
cat("Mean Height for All", MeanHeightAll)
cat("Std. Dev.: Height for All", SdHeightAll)
```

```{webr}
#| max-lines: 0
PlotHeight=ggplot(DataCourseHeight, aes(x=Height, y=5))+
  scale_y_continuous(limits = c(0,10))+
  geom_jitter(height = 1)+
  geom_vline(xintercept = MeanHeightAll)+
  labs(x = "Height", y = "Observation (jittered)")+
  theme_minimal() +
  theme(
    axis.text.y  = element_blank(), 
    axis.ticks.y = element_blank(),  
    panel.grid.major.y = element_blank(),   
    panel.grid.minor.y = element_blank())
ggplotly(PlotHeight)
```
## Mean and Standard Error for the 3 Samples

```{webr}
GroupMeans=group_by(DataCourseHeight, Sample) |> 
  summarise(MeanHeight = mean(Height))
print(GroupMeans)
cat("Standard Error (Group Means)", sd(GroupMeans$MeanHeight))
```

```{webr}
#| max-lines: 0
library(plotly)
ggplotly(ggplot(DataCourseHeight, aes(x=Height, y=5, color=Sample))+
  scale_y_continuous(limits = c(0,10))+
  geom_jitter(size=2, height = 1)+
  geom_vline(xintercept = MeanHeightAll)+
  geom_vline(xintercept = GroupMeans[[1,2]], color="red")+
  geom_vline(xintercept = GroupMeans[[2,2]], color="green")+
  geom_vline(xintercept = GroupMeans[[3,2]], color="blue")+
  labs(x = "Height", y = "Observation (jittered)")+
  theme_minimal() +
  theme(
    axis.text.y  = element_blank(), 
    axis.ticks.y = element_blank(),  
    panel.grid.major.y = element_blank(),   
    panel.grid.minor.y = element_blank(),
    legend.position = c(0.95, 0.95)))
```


## Central Limit Theorem and Why It Is Usefull

> If the original variable is normal distributed (e.g., persons'height) **or** the sample size is greater than 30 (rule of thumb). **Then** *statistics* such as mean and proportion are also normal distributed.

The rule that statistics are normal distribute when the sample size is big enough $\approx > 30$ is called the **Central Limit Theorem**

>Given normal-distributed statistics, we can apply confidence intervals, hypothesis test, and others procedures to statistics (such as means and proportions) although they original variables are not normal-disttributed.

## Estimated Standard Error for Means and Proportions
**Means**
$$SE_{Means}=\frac{StdDev}{\sqrt{n}}$$
**Proportions:**

$$SE_{Prop.}=
\sqrt{\frac{p(1-p)}{N}}$$

## Application: Estimating Mean Height from Sample A 
### (the only sample we can see, but we know the overall Std.Dev.)

```{webr}
#| max-lines: 0
library(plotly)
DataRedSample=filter(DataCourseHeight, Sample=="A")
PlotRedOnly=ggplot(DataRedSample, aes(x=Height, y=5))+
  scale_y_continuous(limits = c(0,10))+
  geom_jitter(size=2, height = 1, color="red")+
  geom_vline(xintercept = GroupMeans[[1,2]], color="red")+
  labs(x = "Height", y = "Observation (jittered)")+
  theme_minimal() +
  theme(
    axis.text.y  = element_blank(), 
    axis.ticks.y = element_blank(),  
    panel.grid.major.y = element_blank(),   
    panel.grid.minor.y = element_blank(),
    legend.position = c(0.95, 0.95))
ggplotly(PlotRedOnly)
```

```{webr}
#| max-lines: 0
cat("Known Standard Deviation:", SdHeightAll)
```

$$SE=\frac{StdDev}{\sqrt{N}}$$
```{webr}
StdError=SdHeightAll/sqrt(5)
print(StdError)
```



## Application: Estimating Confidence Intervall for Mean Height from Sample A 

```{webr}
#| max-lines: 0
PlotConf=invisible(TeachHistConfInterv(SampleMean = mean(DataRedSample$Height) , StandardError=StdError))
```
```{webr}
ggplotly(PlotConf)
```

`qnorm(ProbNotExceeded,Mean,Std/SE)`

```{webr}
qnorm(@@@,@@@,@@@)
```




## Another Application: Mean and Standard Deviation not Known {.smaller}
### (but we have sample data to estimate both)

Our daily sales data for a new product are normally distributed. We **do not know** the *Mean* nor the *Standard Deviation*.

. . .

However, we have data from the last 36 days to estimate the *Mean* and *Standard Deviation* based on our sample data.

. . .

**Sample Size:** $N=36$

**Estimated Mean (in \$ 1,000):** $\bar x= 90$

**Estimated Standard Deviation (in \$ 1,000):** $StdDev= 12$

**Estimated Standard Error:** $SE= \frac{12}{\sqrt{36}}=2$

## t-Distribution vs. Normal Distribution

**If the Standard Deviation is not known and is estimated from the sample:** <br> Normal Distribution changes to a t-Distribution

**However, Normal Distribution and t-Distribution are very similar with the t-Distribution having fatter tails.**



## t-Distribution vs. Normal Distribution ($N=10$)

```{r}
#| echo: false
DataPlot=data.frame(Sales=seq(90-5*2, 90+5*2, 0.1)) |> 
             mutate(SalesStd=(Sales-90)/2) |> 
             mutate(DensityNorm=dnorm(SalesStd)) |> 
             mutate(DensityT=dt(SalesStd, df=9))

ggplot(DataPlot, aes(x=Sales, y=DensityNorm))+
       geom_line(color="red", linewidth=2)+
       geom_line(aes(x=Sales, y=DensityT), color="blue", linewidth=1)
```

## t-Distribution vs. Normal Distribution ($N=36$)

```{r}
#| echo: false
DataPlot=data.frame(Sales=seq(90-5*2, 90+5*2, 0.1)) |> 
             mutate(SalesStd=(Sales-90)/2) |> 
             mutate(DensityNorm=dnorm(SalesStd)) |> 
             mutate(DensityT=dt(SalesStd, df=35))

ggplot(DataPlot, aes(x=Sales, y=DensityNorm))+
       geom_line(color="red", linewidth=2)+
       geom_line(aes(x=Sales, y=DensityT), color="blue", linewidth=1)
```

## Calculating the 95% Confidence Interval for Mean {.smaller}




**Sample Size:** $N=36$
**Estimated Mean in (\$ 1,000):** $\bar x= 90$

**Estimated Standard Deviation:** $StdDev= 12$

**Estimated Standard Error:** $SE= \frac{12}{\sqrt{36}}=2$

**Not correct:** Confidence Interval based on `qnorm()`:

```{webr}
LowerBound=qnorm(0.025, 90, 2)
UpperBound=qnorm(0.975, 90, 2)
cat("The confidence interval based on normal-distribution:", LowerBound, UpperBound)
```

. . .

Confidence Interval based on *t-distribution* (our data are in:


```{webr}
#| echo: false
Sales <- seq(from = -3.5, to = 3.5, length.out = 36)
Sales <- (Sales - mean(Sales)) / sd(Sales) * 12 + 90
```

```{webr}
mean(Sales)
sd(Sales)
SESales=Sales/6
```

```{webr}
t.test(Sales, conf.level=0.95)
```

## Missing Link: Indication that Central Limit Theorem Works

DieDrawVsSample


## Hypothesis Test

**H0:** Before the sample of 36 days was taken, it was believed that the *Mean* for sales is 85?

**A (Research):** Research of the 36 days suggests that the *Mean* is different $(\bar x=90)$.

. . .

Is the mean from the sample just by accident different from 80 or is there more behind it? The latter would mean the H0 is likely wrong.

. . .

Let us work with 95% *Confidence*.

```{r}
TeachHistHypTest(NullHyp=85, StandardError=2, SampleMean=90, SeedValue=123)
```

