---
title: "Hirachical Aglomerative Clustering"

format: 
  revealjs:
    code-fold: true
    scrollable: true
    echo: true
    incremental: false
---


## What Will You Learn/Review {.scrollable .smaller}

-   How a

## Loading the Libraries and the Data

```{r}
LinkageMethod="single"
```


```{r}
library(tidymodels)
library(tidyclust)
library(rio)
library(janitor)
library(modeldata)

set.seed(888)

N=10 #N=100

DataCustSeg=import("Mall_Customers.xlsx") |>  
            select(SpendingScore1To100, AnnualIncome)|> 
            sample_n(N)
```

https://econ.lange-analytics.com/RData/Datasets/CustSegData.xlsx

Info and source for data: 

https://www.kaggle.com/code/himanshikawade04/customersegmentation



## How Hierachicle Aglomaritive Clustering Works
### Each Point is its own Cluster

```{r}
#| echo: false
ggplot(DataCustSeg, aes(x=SpendingScore1To100, y=AnnualIncome))+
geom_point(size=3)+
theme(aspect.ratio = 1)

RecipeCustSeg=recipe(~SpendingScore1To100+AnnualIncome, data=DataCustSeg) |> 
              step_range(min = 0, max = 1)
```

## How Hierachicle Aglomaritive Clustering Works
### 9 Clusters



```{r}
#| echo: false
NumOfClust=9

ModelDesignClust=hier_clust(num_clusters = NumOfClust,
                            linkage_method = LinkageMethod)

WFModelClust=workflow() |> 
             add_model(ModelDesignClust) |> 
             add_recipe(RecipeCustSeg) |> 
             fit(DataCustSeg)
ModelTrainedClust=extract_fit_engine(WFModelClust)
DataCustSegWithAssignm=extract_cluster_assignment(WFModelClust) |> 
                       cbind(DataCustSeg)

DataCentroids=extract_centroids(WFModelClust)


ggplot(DataCustSegWithAssignm, aes(x=SpendingScore1To100, y=AnnualIncome, color=.cluster))+
    geom_point(data=DataCentroids, aes(x=SpendingScore1To100, y=AnnualIncome, color="Centroids"), 
               size=6, color="grey")+
    geom_point(size=3)+
theme(aspect.ratio = 1)
```

## How Hierachicle Aglomaritive Clustering Works
### 8 Clusters



```{r}

#| echo: false
NumOfClust=8



ModelDesignClust=hier_clust(num_clusters = NumOfClust,
                            linkage_method = LinkageMethod)

WFModelClust=workflow() |> 
             add_model(ModelDesignClust) |> 
             add_recipe(RecipeCustSeg) |> 
             fit(DataCustSeg)
ModelTrainedClust=extract_fit_engine(WFModelClust)
DataCustSegWithAssignm=extract_cluster_assignment(WFModelClust) |>                         cbind(DataCustSeg) 

DataCentroids=extract_centroids(WFModelClust)


ggplot(DataCustSegWithAssignm, aes(x=SpendingScore1To100, y=AnnualIncome, color=.cluster))+
    geom_point(data=DataCentroids, aes(x=SpendingScore1To100, y=AnnualIncome, color="Centroids"), 
               size=8, color="grey")+
    geom_point(size=3)+
    theme(aspect.ratio = 1)
```

## How Hierachicle Aglomaritive Clustering Works
### 7 Clusters


```{r}
#| echo: false
NumOfClust=7



ModelDesignClust=hier_clust(num_clusters = NumOfClust,
                            linkage_method = LinkageMethod)

WFModelClust=workflow() |> 
             add_model(ModelDesignClust) |> 
             add_recipe(RecipeCustSeg) |> 
             fit(DataCustSeg)
ModelTrainedClust=extract_fit_engine(WFModelClust) 
DataCustSegWithAssignm=extract_cluster_assignment(WFModelClust) |>                         cbind(DataCustSeg) 

DataCentroids=extract_centroids(WFModelClust)


ggplot(DataCustSegWithAssignm, aes(x=SpendingScore1To100, y=AnnualIncome, color=.cluster))+
    geom_point(data=DataCentroids, aes(x=SpendingScore1To100, y=AnnualIncome, color="Centroids"), 
               size=8, color="grey")+
    geom_point(size=3)+
    theme(aspect.ratio = 1)
```

## How Hierachicle Aglomaritive Clustering Works
### 6 Clusters


```{r}
#| echo: false
NumOfClust=6

ModelDesignClust=hier_clust(num_clusters = NumOfClust,
                            linkage_method = LinkageMethod)

WFModelClust=workflow() |> 
             add_model(ModelDesignClust) |> 
             add_recipe(RecipeCustSeg) |> 
             fit(DataCustSeg)
ModelTrainedClust=extract_fit_engine(WFModelClust) 
DataCustSegWithAssignm=extract_cluster_assignment(WFModelClust) |>                         cbind(DataCustSeg) 

DataCentroids=extract_centroids(WFModelClust)


ggplot(DataCustSegWithAssignm, aes(x=SpendingScore1To100, y=AnnualIncome, color=.cluster))+
    geom_point(data=DataCentroids, aes(x=SpendingScore1To100, y=AnnualIncome, color="Centroids"), 
               size=8, color="grey")+
    geom_point(size=3)
```

## How Hierachicle Aglomaritive Clustering Works
### 5 Clusters


```{r}
#| echo: false
NumOfClust=5

ModelDesignClust=hier_clust(num_clusters = NumOfClust,
                            linkage_method = LinkageMethod)

WFModelClust=workflow() |> 
             add_model(ModelDesignClust) |> 
             add_recipe(RecipeCustSeg) |> 
             fit(DataCustSeg)
ModelTrainedClust=extract_fit_engine(WFModelClust) 
DataCustSegWithAssignm=extract_cluster_assignment(WFModelClust) |>                         cbind(DataCustSeg) 

DataCentroids=extract_centroids(WFModelClust)


ggplot(DataCustSegWithAssignm, aes(x=SpendingScore1To100, y=AnnualIncome, color=.cluster))+
    geom_point(data=DataCentroids, aes(x=SpendingScore1To100, y=AnnualIncome, color="Centroids"), 
               size=8, color="grey")+
    geom_point(size=3)+
    theme(aspect.ratio = 1)
```

## How Hierachicle Aglomaritive Clustering Works
### 4 Clusters


```{r}
#| echo: false
NumOfClust=4

ModelDesignClust=hier_clust(num_clusters = NumOfClust,
                            linkage_method = LinkageMethod)

WFModelClust=workflow() |> 
             add_model(ModelDesignClust) |> 
             add_recipe(RecipeCustSeg) |> 
             fit(DataCustSeg)
ModelTrainedClust=extract_fit_engine(WFModelClust) 
DataCustSegWithAssignm=extract_cluster_assignment(WFModelClust) |>                         cbind(DataCustSeg) 

DataCentroids=extract_centroids(WFModelClust)


ggplot(DataCustSegWithAssignm, aes(x=SpendingScore1To100, y=AnnualIncome, color=.cluster))+
    geom_point(data=DataCentroids, aes(x=SpendingScore1To100, y=AnnualIncome, color="Centroids"), 
               size=8, color="grey")+
    geom_point(size=3)+
    theme(aspect.ratio = 1)
```

## How Hierachicle Aglomaritive Clustering Works
### 3 Clusters


```{r}
#| echo: false
NumOfClust=3

ModelDesignClust=hier_clust(num_clusters = NumOfClust,
                            linkage_method = LinkageMethod)

WFModelClust=workflow() |> 
             add_model(ModelDesignClust) |> 
             add_recipe(RecipeCustSeg) |> 
             fit(DataCustSeg)
ModelTrainedClust=extract_fit_engine(WFModelClust) 
DataCustSegWithAssignm=extract_cluster_assignment(WFModelClust) |>                         cbind(DataCustSeg) 

DataCentroids=extract_centroids(WFModelClust)


ggplot(DataCustSegWithAssignm, aes(x=SpendingScore1To100, y=AnnualIncome, color=.cluster))+
    geom_point(data=DataCentroids, aes(x=SpendingScore1To100, y=AnnualIncome, color="Centroids"), 
               size=8, color="grey")+
    geom_point(size=3)+
    theme(aspect.ratio = 1)
```


## How Hierachicle Aglomaritive Clustering Works
### 2 Clusters


```{r}
#| echo: false
NumOfClust=2

ModelDesignClust=hier_clust(num_clusters = NumOfClust,
                            linkage_method = LinkageMethod)

WFModelClust=workflow() |> 
             add_model(ModelDesignClust) |> 
             add_recipe(RecipeCustSeg) |> 
             fit(DataCustSeg)
ModelTrainedClust=extract_fit_engine(WFModelClust) 
DataCustSegWithAssignm=extract_cluster_assignment(WFModelClust) |>                         cbind(DataCustSeg) 

DataCentroids=extract_centroids(WFModelClust)


ggplot(DataCustSegWithAssignm, aes(x=SpendingScore1To100, y=AnnualIncome, color=.cluster))+
    geom_point(data=DataCentroids, aes(x=SpendingScore1To100, y=AnnualIncome, color="Centroids"), 
               size=8, color="grey")+
    geom_point(size=3)+
    theme(aspect.ratio = 1)
```

## How Hierachicle Aglomaritive Clustering Works
### 1 Clusters


```{r}
#| echo: false
NumOfClust=1

ModelDesignClust=hier_clust(num_clusters = NumOfClust,
                            linkage_method = LinkageMethod)

WFModelClust=workflow() |> 
             add_model(ModelDesignClust) |> 
             add_recipe(RecipeCustSeg) |> 
             fit(DataCustSeg)
ModelTrainedClust=extract_fit_engine(WFModelClust) 
DataCustSegWithAssignm=extract_cluster_assignment(WFModelClust) |>                         cbind(DataCustSeg) 

DataCentroids=extract_centroids(WFModelClust)


ggplot(DataCustSegWithAssignm, aes(x=SpendingScore1To100, y=AnnualIncome, color=.cluster))+
    geom_point(data=DataCentroids, aes(x=SpendingScore1To100, y=AnnualIncome, color="Centroids"), 
               size=8, color="grey")+
    geom_point(size=3)+
    theme(aspect.ratio = 1)
```



## Dendogram 
### Clustering Pricess in a Graph

```{r}
#| echo: false
plot(ModelTrainedClust)
```


## Elbow Method to Determine an Appropiate Number of Clusters


```{r}
#| echo: false
ggplot(aes(y=Y,x=X), data=tibble(Y=ModelTrainedClust$height, X=c((N-1):1)))+
       geom_line()+
        geom_line()+
        scale_x_continuous(breaks = c(1:(N-1)))
```
## How Hierachicle Aglomaritive Clustering Works
### Appropiate Cluster Number (=4)


```{r}
RecipeCustSeg=recipe(~SpendingScore1To100+AnnualIncome, data=DataCustSeg) |> 
              step_range(min = 0, max = 1)


ModelDesignClust=hier_clust(num_clusters = 4,
                            linkage_method = "ward.D") |> 
                 set_engine("stats") |> 
                 set_mode("partition")

WFModelClust=workflow() |> 
             add_model(ModelDesignClust) |> 
             add_recipe(RecipeCustSeg) |> 
             fit(DataCustSeg)
ModelTrainedClust=extract_fit_engine(WFModelClust) 
DataCustSegWithAssignm=extract_cluster_assignment(WFModelClust) |>                         cbind(DataCustSeg) 

DataCentroids=extract_centroids(WFModelClust)
```


```{r}
ggplot(DataCustSegWithAssignm, aes(x=SpendingScore1To100, y=AnnualIncome, color=.cluster))+
    geom_point(data=DataCentroids, aes(x=SpendingScore1To100, y=AnnualIncome, color="Centroids"), 
               size=8, color="grey")+
    geom_point(size=3)+
    theme(aspect.ratio = 1)
```
 
 ## Muti Variable