---
title: "Tree Based Models"
subtitle: "Random Forest"
author:
  name: "Carsten Lange"
  email: "clange@cpp.edu"
  affiliation: "Cal Poly, Pomona"

format: 
  live-revealjs:
    theme: [moon, CustomCL.scss]
    smaller: true
    controls: true
    chalkboard:
      theme: whiteboard
      boardmarker-width: 2.7
    incremental: false
    scrollable: true
    footer: "<a href='https://ai.lange-analytics.com/htmlbook/RidgeLasso.html' target='_blank'>Textbook </a> "
   
webr:
  packages: ['dplyr', 'rsample', 'janitor',  'recipes', 'broom',
             'yardstick','parsnip','workflows',ranger] # Auto-install packages in the browser

engine: knitr
---

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}



## Introduction

```{r}
library(tidymodels);library(rio);library(janitor);library(kableExtra)
```




-   *Random Forest* can be used for

    -   Classification
    -   Regression

-   A *Random Forest* model is an ensample model consisting of many (hundreds or thousands) of slightly different *Decision Trees*.

- The majority predictions (majority vote) from all *Decision Trees* becomes the prediction of the *Random Forest* for classification» 

- The average predictions from all *Decision Trees* becomes the prediction of the *Random Forest* for regression» 

## The Idea Behind Random Forest
### The Wisdom of the Crowd

The idea that a combination of weak predictors can lead to a strong prediction is analogous to the wisdom of crowds
phenomenon described in Galton 1907:

*Visitors at a stock and poultry exhibition in England submitted
guesses about the weight of an ox. Although most of the visitors were off with their predictions, surprisingly
the mean of all predictions was very close to the real weight of the ox.*

## The Idea Behind Random Forest
### Using Many (Slightly) Different Decision Trees

1. Train many (hundreds or even more) different Decision Trees

2. To predict an outcome for an observation use all Desicion Trees, which results in hundrets (or more predictions)

3. Aggregate the predictions<br><br>
  **Regression:** Calculate the average prediction<br><br>
  **Classification:** Treat each prediction as a vote for a class --- the class with most of the votes is the predicted class

## How to ensure that Decision Trees are (slightly) Different {.smaller}

- **Random Subspace Method:** Each decision tree uses only a random selection of the predictors.
     - As a rule of thumb: If we have $m$ predictors, each decision tree uses only $\sqrt(M)$ predictors (rounded down if needed).
     - In the case here, where we have $7$ predictors, each decision tree uses only $2$ randomly chosen predictors ($\sqrt(7)=2.65$ rounded down to $2$)<br> <br>  

- **Bagging:** Each decision tree is presented with slightly different training data: We draw from the original training data with replacement to step-wise generate a new training dataset (**Bootstrapping**).
     - The new training dataset has the same size as the original dataset.
     - Generate a bootstrap dataset for every decision tree. 
     
## Example for Bagging with Bootstrapping {.scrollable .smaller}

```{webr}
#| context: setup
#| echo: false
DataTrain= tibble(ID=seq(1:7),
                  name=c("Adam","Berta","Carlos","Dora","Ernst","Fiola","Gert"),
                  Sex=c("male","female","male","female","male","female","male"),
                  Age=c(22,54,32,21,32,87,18),
                  Weight=c(145,180,167,197,221,146,230))
```

Creating 5 Bootstrap Training Datasets `DataTrain1`, `DataTrain2`, `DataTrain3`, `DataTrain4`, `DataTrain5` from original dataset `DataTrain`.

::::: columns
:::: {.column width="49%"}

**Original Training Data:**


```{webr}
DataTrain= data.frame(ID=seq(1:7),
                  name=c("Adam","Berta","Carlos","Dora","Ernst","Fiola","Gert"),
                  Sex=c("male","female","male","female","male","female","male"),
                  Age=c(22,54,32,21,32,87,18),
                  Weight=c(145,180,167,197,221,146,230))
kable(DataTrain)
```

::::

:::: {.column width="1%"}
.
::::

:::: {.column width="49%"}

**1st Bootstrap Sample:**

```{webr}
#| exercise: exc1
set.seed(771)
DataTrain1=sample_n(______, ______, replace = ______)
kable(DataTrain1)
```

::: {.solution exercise="exc1"}
::: { .callout-tip collapse="false"}

## Solution: you can copy and paste

```r
set.seed(771)
DataTrain1=sample_n(DataTrain, 7, replace = TRUE)
kable(DataTrain1)
```
:::
:::

::::
:::::

## Example for Bagging with Bootstrapping {.scrollable .smaller}

Creating 5 Bootstrap Training Datasets `DataTrain1`, `DataTrain2`, `DataTrain3`, `DataTrain4`, `DataTrain5` from original dataset `DataTrain`.

::::: columns
:::: {.column width="40%"}

**Original Training Data:**


```{webr}
#| autorun: true
#| echo: false
DataTrain= data.frame(ID=seq(1:7),
                  name=c("Adam","Berta","Carlos","Dora","Ernst","Fiola","Gert"),
                  Sex=c("male","female","male","female","male","female","male"),
                  Age=c(22,54,32,21,32,87,18),
                  Weight=c(145,180,167,197,221,146,230))
kable(DataTrain)
```

::::



:::: {.column width="60%"}

**2nd Bootstrap Sample:**

```{webr}
set.seed(772)
DataTrain2=sample_n(DataTrain, 7, replace = TRUE)
kable(DataTrain2)
```



::::
:::::

## Example for Bagging with Bootstrapping {.scrollable .smaller}

Creating 5 Bootstrap Training Datasets `DataTrain1`, `DataTrain2`, `DataTrain3`, `DataTrain4`, `DataTrain5` from original dataset `DataTrain`.

::::: columns
:::: {.column width="40%"}

**Original Training Data:**


```{webr}
#| autorun: true
#| echo: false
DataTrain= data.frame(ID=seq(1:7),
                  name=c("Adam","Berta","Carlos","Dora","Ernst","Fiola","Gert"),
                  Sex=c("male","female","male","female","male","female","male"),
                  Age=c(22,54,32,21,32,87,18),
                  Weight=c(145,180,167,197,221,146,230))
kable(DataTrain)
```

::::



:::: {.column width="60%"}

**3rd Bootstrap Sample:**

```{webr}
set.seed(773)
DataTrain3=sample_n(DataTrain, 7, replace = TRUE)
kable(DataTrain3)
```

::::
:::::

## Example for Bagging with Bootstrapping {.scrollable .smaller}

Creating 5 Bootstrap Training Datasets `DataTrain1`, `DataTrain2`, `DataTrain3`, `DataTrain4`, `DataTrain5` from original dataset `DataTrain`.

::::: columns
:::: {.column width="40%"}

**Original Training Data:**


```{webr}
#| autorun: true
#| echo: false
DataTrain= data.frame(ID=seq(1:7),
                  name=c("Adam","Berta","Carlos","Dora","Ernst","Fiola","Gert"),
                  Sex=c("male","female","male","female","male","female","male"),
                  Age=c(22,54,32,21,32,87,18),
                  Weight=c(145,180,167,197,221,146,230))
kable(DataTrain)
```

::::



:::: {.column width="60%"}

**4th Bootstrap Sample:**

```{webr}
set.seed(774)
DataTrain4=sample_n(DataTrain, 7, replace = TRUE)
kable(DataTrain4)
```



::::
:::::

## Example for Bagging with Bootstrapping {.scrollable .smaller}

Creating 5 Bootstrap Training Datasets `DataTrain1`, `DataTrain2`, `DataTrain3`, `DataTrain4`, `DataTrain5` from original dataset `DataTrain`.

::::: columns
:::: {.column width="40%"}

**Original Training Data:**


```{webr}
#| autorun: true
#| echo: false
DataTrain= data.frame(ID=seq(1:7),
                  name=c("Adam","Berta","Carlos","Dora","Ernst","Fiola","Gert"),
                  Sex=c("male","female","male","female","male","female","male"),
                  Age=c(22,54,32,21,32,87,18),
                  Weight=c(145,180,167,197,221,146,230))
kable(DataTrain)
```

::::



:::: {.column width="60%"}

**5th Bootstrap Sample:**

```{webr}
set.seed(775)
DataTrain5=sample_n(DataTrain, 7, replace = TRUE)
kable(DataTrain5)
```



::::
:::::

## Example for Bagging with Bootstrapping {.scrollable .smaller}
### Summary

:::: columns
::: {.column width="50%"}

**Data Train**
```{webr}
#| autorun: true
#| echo: false
DataTrain= data.frame(ID=seq(1:7),
                  name=c("Adam","Berta","Carlos","Dora","Ernst","Fiola","Gert"),
                  Sex=c("male","female","male","female","male","female","male"),
                  Age=c(22,54,32,21,32,87,18),
                  Weight=c(145,180,167,197,221,146,230))
kable(DataTrain)
```

**2nd Bootstrap Sample**
```{webr}
#| autorun: true
#| echo: false
set.seed(772)
DataTrain2=sample_n(DataTrain, 7, replace = TRUE)
kable(DataTrain2)
```

**4th Bootstrap Sample**
```{webr}
#| autorun: true
#| echo: false
set.seed(774)
DataTrain4=sample_n(DataTrain, 7, replace = TRUE)
kable(DataTrain4)
```
:::

::: {.column width="50%"}
**1st Bootstrap Sample**
```{webr}
#| autorun: true
#| echo: false
set.seed(771)
DataTrain1=sample_n(DataTrain, 7, replace = TRUE)
kable(DataTrain1)
```

**3rd Bootstrap Sample**
```{webr}
#| autorun: true
#| echo: false
set.seed(773)
DataTrain3=sample_n(DataTrain, 7, replace = TRUE)
kable(DataTrain3)
```

**5th Bootstrap Sample**
```{webr}
#| autorun: true
#| echo: false
set.seed(775)
DataTrain5=sample_n(DataTrain, 7, replace = TRUE)
kable(DataTrain5)
```

:::
::::

 

## Real World Data with a Decision Tree

Predicting vaccination rates in the U.S. based on data from September 2021:

- Outcome variable: Percentage of fully vaccinated (two shots) people ($PercVacFull$).

- Data from 2,630 continental U.S. counties.»

## Real World Data with a Decision Tree --- Predictor Variables {.smaller}

- Race/Ethnicity:

  - Counties' proportion African Americans ($PercBlack$), 
  - Counties' proportion Asian Americans ($PercAsian$), and 
  - Counties' proportion Hispanics ($PercHisp$)

- Political Affiliation (Presidential election 2020): 
  
  - Counties' proportion Republican votes ($PercRep$)

- Age Groups in Counties: 
  
  - Counties' proportion young adults (20-25 years);  $PercYoung25$
  
  - Counties' proportion older adults (65 years and older); $PercOld65$

- Income related: 
  - Proportion of households receiving food stamps ($PercFoodSt$)»

## Loading the Data and Assigning Training and Testing Data{.smaller}
kable|>
```{r}
DataVax=import("https://ai.lange-analytics.com/data/DataVax.rds") |>   
        select(County, State, PercVacFull, PercRep,
              PercAsian, PercBlack, PercHisp,
              PercYoung25, PercOld65, PercFoodSt)

set.seed(2021)
Split85=DataVax |> initial_split(prop = 0.85,
                                 strata = PercVacFull,
                                 breaks = 3)

DataTrain=training(Split85) |> select(-County, -State)
DataTest=testing(Split85) |> select(-County, -State)
kable(head(DataVax))
```

## Creating Model Design, Recipe, and Fitted Workflow (default hyper-parameters)

```{r }
#| echo: true

DefaultMtry=floor(sqrt(7)) # floor() rounds down

ModelDesignRandFor=rand_forest(trees=500, min_n=5, mtry = DefaultMtry) |> 
                    set_engine("ranger") |> 
                    set_mode("regression")
  
RecipeVax=recipe(PercVacFull~., data=DataTrain)

WfModelVax=workflow() |> 
               add_model(ModelDesignRandFor) |> 
               add_recipe(RecipeVax) |> 
               fit(DataTrain)
```
```{r}
cat("Default value for mtry is:", DefaultMtry)
print(WfModelVax)
```

## Metrics for Default Random Forest Model

```{r}
#| echo: true
DataTestWithPred=augment(WfModelVax, new_data=DataTest)
metrics(DataTestWithPred, truth=PercVacFull, estimate=.pred)
```

## Tuning the Random Forest Model

On the next slide you find a link to an R script that you can use to tune the Random Forest model. 

Try to beat the predictive quality of the *default* model we used here.

## Tuning the Random Forest Model {.smaller}

```{r}
#| echo: true
#| code-fold: false
ModelDesignRandForTree=rand_forest(trees=500, min_n=tune(), mtry = tune()) |> 
                    set_engine("ranger") |> 
                    set_mode("regression")
```

What to consider for the tuning grid?

- `trees=500`: No need to tune because a high number of trees cannot lead to overlearning. This is because the Random Forest prediction is calculated from the average of the trees. However, a high number of trees slows the training (`trees=500' was set; feel free to change it).

- `mtry=tune()` is a hyper-parameter that should be tuned. It determines the number of predictors randomly considered for each decision tree. Recommended is $\sqrt(7)=2.65$. Maybe you try 2, 3, 5.

- `min_n=tune()` is a hyper-parameter that should be tuned. It determines the minimum  number of observations inside a node required for a split into child nodes. If this number is not reached the node becomes a terminal node. Therefore, `min_n=tune()` indirectly determines the tree depth.  Default is $5$. Maybe you try 3, 5, 10, 30.

. . .

Below is a link for an R script that you can use to tune the *Random Forest* model.

::: footer
See: [Script to tune the Random Forest Model](https://econ.lange-analytics.com/RScripts/TreeBasedRandForTuningScript.R) 
:::



