---
title: "Linear Regression"
subtitle: "Univariate"
format: 
  revealjs:
    code-fold: true
    scrollable: true
---

```{r}
library(tidymodels)
library(rio)
library(kableExtra)
library(janitor)
DataMockup=import("https://ai.lange-analytics.com/data/DataStudyTimeMockup.rds")
```

## What Will You Learn {.scrollable .smaller}

-   Reviewing the basic idea behind linear regression

-   Learning how how to measure predictive quality with Mean Square Error ($MSE$).

-   Understanding the role of in a linear regression model (applies also to other models)

-   Calculating optimal regression parameters using OLS

-   Finding optimal regression parameters by trial and error

-   Finding optimal regression parameters with `lm` and `tidymodels`.

-   Distinguish between unfitted and fitted models

-   Using `tidymodels` to split observations randomly into a training and testing dataset.

-   Using `tidymodels` to analyze housing prices.


## Univariate Linear Regression - Data Table and Goal {.smaller}

::: columns

::: {.column width="60%"}
**The Data Table**

```{r }
kbl(DataMockup |> mutate(i=1:5) |> select(i,everything()), 
    caption="Mockup Training Dataset")|>
  add_header_above(c(" ", "y", "x"), escape=F) |> 
  kable_styling(bootstrap_options=c("striped","hover"), full_width = F, position="center")
```
:::

::: {.column width="40%"}
**The Regression:**

$$
\widehat{y}_{i} = \beta_{1}x_{i}+\beta_{2}
$$

$$
\widehat{Grade}_{i} = \beta_{1}StudyTime_{i}+\beta_{2}
$$

**The Goal**

Find values for $\beta_1$ and $\beta_2$ that minimize the prediction errors $$(\widehat{y}_{i}-y_i)^2$$
$$(\widehat{Grade}_{i}-Grade_i)^2$$
:::


:::

## Predict Grade Based on Study Time with Eyeballing {.smaller}


Copy the code from the code-section of the slide into an *R* script and try to adjust $\beta_1$ and $\beta_1$ to get the best prediction equation.

$$
\widehat{y}_{i} = \beta_{1}x_{i}+\beta_{2}
$$
$$
\widehat{Grade}_{i} = \beta_{1}StudyTime_{i}+\beta_{2}
$$

```{r}
#| eval: false
#| echo: true
Beta1=0
Beta2=50
source("https://ai.lange-analytics.com/source/RegrLinePlot4Slides.R")

```


## Univariate Linear Regression - Data Diagram and Goal {.smaller}

::: columns
::: {.column width="40%"}
**The Regression:**

$$
\widehat{y}_{i} = \beta_{1}x_{i}+\beta_{2}
$$
$$
\widehat{Grade}_{i} = \beta_{1}StudyTime_{i}+\beta_{2}
$$

**The Goal**

Find values for $\beta_0$ and $\beta_1$ that minimize the individual (squared) prediction errors: 
$$(\widehat{y}_{i}-y_i)^2$$
$$(\widehat{Grade}_{i}-Grade_i)^2$$
:::

::: {.column width="60%"}
**The Data Diagram**

```{r}
Model123=lm(Grade~StudyTime, data=DataMockup)
PredGrade=predict(Model123, DataMockup)
ggplot(DataMockup, aes(x=StudyTime,y=Grade)) +
  geom_line(aes(y=PredGrade), color="red", size=2.7)+
  geom_point(size=5, color="blue")+
  geom_point(aes(y=PredGrade), color="black", size=2.7)+
  geom_segment(aes(x = StudyTime, y = PredGrade,
                   xend = StudyTime, yend = Grade),size=1.2)+
  scale_x_continuous("Study Time", breaks=seq(1,8))+
  scale_y_continuous(limits=c(65,110), breaks=seq(60,100,5))
```

```{r}
#| eval: false
#| echo: true
Beta1=4
Beta2=64
```

:::
:::

# How to Measure Prediction Quality with the Mean Squared Error (MSE) {.smaller}

```{=tex}
\begin{eqnarray*}
MSE & = & \frac{1}{N} \sum_{i=1}^{N}(\widehat{y}_{i}-y_{i})^{2} \\
 & \Longleftrightarrow& \\
MSE & =  & \frac{1}{N} \sum_{i=1}^{N}(\underbrace{\overbrace{\beta_{1}x_{i}+\beta_2}^{\mbox{Prediction $i$}}-y_i}_{\mbox{Error $i$}})^2
\end{eqnarray*}
```
**Note, when the data are given (i.e.,** $x_i$ and $y_i$ are given), the $MSE$ depends only on the choice of $\beta_1$ and $\beta_2$ »

## How to Measure Prediction Quality with the MSE {.smaller .scrollable}

```{=tex}
\begin{eqnarray}
MSE & = & \frac{(\beta_1x_{1}+\beta_2-y_1)^2
       +(\beta_1x_{2}+\beta_2-y_2)^2 +  \cdots+
       (\beta_1x_{5}+\beta_2-y_{5})^2}{5} \nonumber \\
       & \Longleftrightarrow& \nonumber \\
MSE & = & \frac{1}{5}\left[ (\underbrace{\overbrace{\beta_1\cdot 2+\beta_2}^{\mbox{Prediction $1$}}-65}_{\mbox{Error $1$}})^2
       +(\underbrace{\overbrace{\beta_1\cdot 3+\beta_2}^{\mbox{Prediction $2$}}-82}_{\mbox{Error $2$}})^2\right.\nonumber \\
   &   &\nonumber \\
       && +
    (\underbrace{\overbrace{\beta_1\cdot 7+\beta_2}^{\mbox{Prediction $3$}}-93}_{\mbox{Error $3$}})^2 
+(\underbrace{\overbrace{\beta_1\cdot 8+\beta_2}^{\mbox{Prediction $4$}}-93}_{\mbox{Error $4$}})^2\nonumber \\
   &   &\nonumber \\
       && +\left. (\underbrace{\overbrace{\beta_1\cdot 4+\beta_2}^{\mbox{Prediction $5$}}-83}_{\mbox{Error $6$}})^2\right]
\end{eqnarray}
```


## Custom R Function to Calculate MSE {.scrollable}

```{r}
#| echo: false
FctMSE=function(VecBeta, data){
                        Beta1=VecBeta[1]
                        Beta2=VecBeta[2]
                        data=data |>
                        rename(Y=1, X=2) |> 
                        mutate(YPred=Beta1*X+Beta2) |>
                        mutate(Error=YPred-Y) |>
                        mutate(ErrorSqt=Error^2)
                        
                        MSE=mean(data$ErrorSqt)
                        
                        return(MSE)}
```

Function Call:

```{r}
#| echo: true
VecBeta=c(4,61)
ResultMSE=FctMSE(VecBeta, DataMockup)
print(ResultMSE)
```

Function Definition:»

```{r}
#| echo: true
FctMSE=function(VecBeta, data){
                        Beta1=VecBeta[1]
                        Beta2=VecBeta[2]
                        data=data |>
                        rename(Y=1, X=2) |> 
                        mutate(YPred=Beta1*X+Beta2) |>
                        mutate(Error=YPred-Y) |>
                        mutate(ErrorSqt=Error^2)
                        
                        MSE=mean(data$ErrorSqt)
                        
                        return(MSE)}
```

## How to Find Optimal Values for $\beta_1$ and $\beta_2$ {.smaller .scollable}

**Method 0: (not such a good idea)** 

:   We can use **unsystematic trial and error**.

    ```{r}
    #| echo: true
    library(rio)
    FctMSE=import("https://ai.lange-analytics.com/source/FctMSE.rds")
    VecBeta=c(7,60)
    ResultMSE=FctMSE(VecBeta, DataMockup)
    print(ResultMSE)
    ```


**Method 1:**

:   We can use a **systematic trial and error process**.

**Method 2:**

:   **Calculate optimal values** for the parameters (the $\beta s$) based on Ordinary Least Squares (OLS) using two formulas (**Note,** this method works only for linear regression)



## Method 1: Use a Systematic Trial and Error Process 🤓

-   **Grid Search (aka Brute Force):**

    1.  For a given range of $\beta_1$ and $\beta_2$ values, build a table with pairs of all combinations of these $\beta s$.
    2.  Then use our custom `FctMSE()` command to calculate a $MSE$ for each $\beta$ pair.
    3.  Find the $\beta$ pair with the lowest $MSE$

-   **Optimizer:** Use the R build-in optimizer. Push the start values for $\beta_1$ and $\beta_2$ together with the data to the optimizer as arguments. The rest is done by the optimizer.

-   See the R script in the footnote to see both algorithms in action.»

::: footer
See: [LinRegrScript.R script100](https://lange-analytics.com/AIBook/Scripts/LinRegrScript.R)
:::

## Method 2: Calculate Optimal Parameters (only for OLS!){.smaller}

```{=tex}
\begin{eqnarray*} 
\beta_{1,opt}&=& \frac
{N \sum_{i=1}^N y_i x_i- \sum_{i=1}^N y_i \sum_{i=1}^N x_i} 
{N \sum_{i=1}^N x_i^2 - \left (\sum_{i=1}^N x_i \right )^2}=3.96\\
&& \nonumber \\
\beta_{2,opt.}&=& \frac{ \sum_{i=1}^N y_i  - \beta_1 \sum_{i=1}^N x_i} {N} = 64.18
\end{eqnarray*}
```
```{r}
DataTable=DataMockup |>
  mutate(GradeXStudyTime=Grade*StudyTime) |>
  mutate(StudyTimeSquared=StudyTime^2) 


kbl(DataTable |> mutate(i=1:5) |> select(i,everything()), 
    caption="Mockup Training Dataset ")|>
  add_header_above(c(" ", "y", "x", "y x","x x"), escape=F) |> 
  kable_styling(bootstrap_options=c("striped","hover"), full_width = F, position="center")
```



# Using lm() Command from Base R to Run a Regression {.scrollable}

```{r}
#| echo: true
ModelStudyTime=lm(Grade~StudyTime, data=DataMockup)
summary(ModelStudyTime)
```

## Results from OLS Analysis and Sanity Check

**Unfitted Model**

$$Grade=\beta_1 StudyTime + \beta_2$$

**Fitted Model**

$$Grade=3.96 StudyTime + 64.18$$
Coefficient for StudyTime $\beta_1$ should be positive.

## Predicting with the Fitted Model - Generated Data {.smaller}

$$Grade=3.96 \cdot StudyTime + 64.18$$
```{r}
#| echo: true
ModelStudyTime=lm(Grade~StudyTime, data=DataMockup)

DataForPrediction=data.frame(StudyTime=c(1,2,3))

predict(ModelStudyTime, newdata=DataForPrediction)
```

Interpretation of $\beta_1=3.96$:

1. From the formula: If $StudyTime$ increases, Grade increases. -> makes sense!

2. From the formula: Quantitative impact: If $StudyTime$ increases by one unit, Grade increases by $3.96$ units.

3. From the prediction: Above we predicted $Grade$ for $Studyteimes$ of 1,2,3, which confirms point 1) and 2).

## Predicting with the Fitted Model - Training Data


$$Grade=3.96 StudyTime + 64.18$$
Generating the predictions:

```{r}
#| echo: true

ModelStudyTime=lm(Grade~StudyTime, data=DataMockup)

VecPred=predict(ModelStudyTime, newdata=DataMockup)
print(VecPred)
```

## Predicting with the Fitted Model - Training Data

$$Grade=3.96 StudyTime + 64.18$$
Augmenting the predictions to the data frame:

```{r}
#| echo: true
VecPred=predict(ModelStudyTime, newdata=DataMockup)
DataMockup=DataMockup |> 
           mutate(.pred=VecPred)
print(DataMockup)
```

## Using tidymodels for the Linear Regression

Creating the fitted (workflow) model:

```{r}
#| echo: true
library(tidymodels)
DataMockup=import("https://ai.lange-analytics.com/data/DataStudyTimeMockup.rds")
RecipeStudyTime=recipe(DataMockup, Grade~StudyTime)
ModelDesignStudyTime=linear_reg() |> 
                     set_engine("lm") |> 
                     set_mode("regression")
WFModelStudyTime= workflow() |> 
                  add_recipe(RecipeStudyTime) |> 
                  add_model(ModelDesignStudyTime) |> 
                  fit(DataMockup)
```

##
Getting the Results from a Workflow

```{r}
#| echo: true
ResultsStudyTime=tidy(WFModelStudyTime)
print(ResultsStudyTime)

```


## Adding Predictions to DataMockup

```{r}
#| echo: true
DataMockupWithPred=augment(WFModelStudyTime, DataMockup)
print(DataMockupWithPred)
```
## Evaluating Predictive Quality

```{r}
#| echo: true
metrics(DataMockupWithPred, truth=Grade, estimate=.pred)
```

# Real World Application

## Univariate OLS with a Real World Dataset

**Data Description:**

-   King County House Sale dataset (Kaggle 2015). House sales prices from May 2014 to May 2015 for King County in Washington State.
-   Several predictor variables. For now we use only $Sqft$
-   We will only use 500 randomly chosen observations from the total of 21,613 observations.

-We only use Sqft as predictor variable for now.


## Univariate OLS with a Real World Dataset

**Data:**

```{r}
#| echo: true
library(rio)
DataHousing =
  import("https://ai.lange-analytics.com/data/HousingDataSmall.csv")
head(DataHousing)
```

## Univariate OLS with a Real World Dataset


**Unfitted Model:** 

$$\widehat{Price}=\beta_1 Sqft + \beta_2$$
**Directed Acyclic Graphs (causal graphs):**

## Univariate OLS with a Real World Dataset {.scrollable}

**Splitting in Training and Testing Datasets**

```{r}
#| echo: true
library(tidymodels)
set.seed(777)
Split7030=initial_split(0.7,data=DataHousing, strata = Price, breaks = 5)
DataTrain=training(Split7030)
DataTest=testing(Split7030)
```

**DataTrain**

```{r}
print(DataTrain, max=6)
```

**DataTest**

```{r}
print(DataTest, max=6)
```

## Univariate OLS with a Real World Dataset

**Run the Analysis with lm()**

```{r}
#| echo: true
library(rio)
DataHousing =
  import("https://ai.lange-analytics.com/data/HousingDataSmall.csv")
library(tidymodels)
Split7030=initial_split(0.7,data=DataHousing, strata = Price, breaks = 5)
DataTrain=training(Split7030)
DataTest=testing(Split7030)

ModelHouses=lm(Price ~ Sqft, data=DataTrain)
summary(ModelHouses)
```

## The Same Analysis with tidymodels

```{r}
#| echo: true
library(tidymodels)

RecipeHouses=recipe(DataTrain, Price~Sqft)
ModelDesignHouses=linear_reg() |> 
                     set_engine("lm") |> 
                     set_mode("regression")

WFModelHouses= workflow() |> 
               add_recipe(RecipeHouses) |> 
               add_model(ModelDesignHouses) |> 
               fit(DataTrain)

Results=tidy(WFModelHouses)
print(Results)
```

## Predicting with tidymodels Training Data {.scrollable}

The augment() command first predicts and then augments the results to the related data frame

```{r}
#| echo: true
DataTrainWithPred=augment(WFModelHouses, new_data = DataTrain)
DataTestWithPred=augment(WFModelHouses, new_data = DataTrain)

head(DataTrainWithPred)
head(DataTestWithPred)

```

## Metrics for Predictive Quality Training Data

```{r}
#| echo: true
metrics(DataTrainWithPred, truth = Price, estimate = .pred)
```

## Metrics for Predictive Quality Testing Data

```{r}
#| echo: true
metrics(DataTrainWithPred, truth = Price, estimate = .pred)

```



