---
title: "GIS - Spatial Econometrics"
subtitle: "Joining two Tables"
format: 
  revealjs:
    echo: true
    code-fold: true
    incremental: false
    scrollable: true
---

```{r}
#| include: false
library(tidymodels)
library(rio)
library(janitor)
DataHouses= import("../../Data/DataHouseSampleCL.csv")
library(tidyverse)
NameAdresses=tibble(ID=c(1,22,3,4,5,6),
                    First=c("John","Joe","Jake",
                            "Hans","Anne", "Agatha"),                   
                  Last=c("Doe","Dull","Dune",
                          "Meier","Scholz","Smith"),
                  Street=c("649 C Street","648 A Street", "345 H Street",
                           "118 B Street","120 B Street","876 G Street"),
                  City=c("Pomona","Pomona","Pomona", 
                         "Pomona","Pomona","Pomona"))
Orders=tibble(ID=c(1,22,22,3,5,3,5,5,27,6),
              Item=c("Shoes","Shirt","Shoes","Shoe",
                     "Shirt","Shoes","Jeans","Jeans",
                    "Shoes","Shirt"),
              Amount=c(123,67,78,111,98,17,45,37,22,11)
  )                    


```

## What Will You Learn {.scrollable}

- Load a csv table to *My Content*
- Add a table from *My Content* to a *Web Map*

- What a relational database is
- Joining two tables into one table based on matching observations
    - Left Join
    - Inner Join
    - Outer or Full Join (rarely used)
- Join a *FeatureLayer*'s table with a *Table* in a *Web Ma*p

## ArcGis Online: 

1.  Load  Web map from `SchoolQualityYourInitial` from *My Content*

**Problem:** No school quality reported

## Adding Table with School Quality to ArcGIS Online

**We start with:** Showing School Quality in R: `Schools.R` (this step is not required)

**Next:** You add CSV file with school quality:<br> `DataEC4448/DataSchoolQuality.csv` to *My Content*:

Content -\> New Item -\> . . . (as feature table)


## ArcGis Online: Add table to Map

1.  Load  Web map from `SchoolQualityYourInitial` from *My Content*
2.  Add Table (table icon on the left)
3.   Choose: `DataSchoolQuality`

**Problem:** One layer has no school quality and the SchoolQuality table has no location information.

**Solution:**  Both have a school identifier (CDS code). So, we can join them.



## Joining two Tables {.smaller .scrollable}

Programs that allow joining two tables are called **relational databases**

```{r}
#| echo: false
Orders
```

```{r}
#| echo: false
NameAdresses
```

## Joining two Tables (Left Join) {.smaller}

::: columns
::: {.column width="40%"}
```{r}
#| echo: false
Orders
```
:::

::: {.column width="60%"}
```{r}
#| echo: false
NameAdresses
```
:::
:::

**Left Join**

. . .

```{r}
left_join(Orders,NameAdresses, by=join_by(ID==ID))
```

## Joining two Tables (Inner Join) {.smaller}

::: columns
::: {.column width="40%"}
```{r}
#| echo: false
Orders
```
:::

::: {.column width="60%"}
```{r}
#| echo: false
NameAdresses
```
:::
:::

**Inner Join**

. . .

```{r}
inner_join(Orders,NameAdresses, by=join_by(ID==ID))
```

## Joining two Tables (Outer Join)

::: columns
::: {.column width="40%"}
```{r}
#| echo: false
Orders
```
:::

::: {.column width="60%"}
```{r}
#| echo: false
NameAdresses
```
:::
:::

**Outer Join**

. . .

```{r}
full_join(Orders,NameAdresses, by=join_by(ID==ID))
```



## Joining two Tables in ArcGis Online

1.  Back to the web map `SchoolQualityYourInitial`

If you have not done it already: 

2.  Add Table (table icon on the left)
3.   Choose: `DataSchoolQuality`
4.  both have a school identifier (CDS code)

Joining based on attributes/features:

5.  **Joining:** Analysis -\> Summarize Data -\> Join Features
Result Layer: `SchoolQualityWithAdrYourInitial` Folder: `EC4448Schools` 

**Save Web map!!!**
