---
title: "Key Machine Learning Concepts"
subtitle: "Explained with Linear Regression"
format: 
  revealjs:
    code-fold: true
    incremental: false
    scrollable: true
    echo: true
---

## Loading Required Librarries

```{r}
#| code-fold: false
library(tidymodels)
library(rio)
library(kableExtra)
library(janitor)
DataMockup=import("https://ai.lange-analytics.com/data/DataStudyTimeMockup.rds")
```

## What Will You Learn {.scrollable .smaller}

-   Reviewing the basic idea behind linear regression

-   Learning how how to measure predictive quality with Mean Square Error ($MSE$).

-   Understanding the role of parameters in a machine learning model in general and in linear regression in particular

-   Calculating optimal regression parameters using OLS

-   Finding optimal regression parameters by trial and error

-   Distinguish between unfitted and fitted models

-   Using the `tidymodels` package to split observations from a dataset randomly into a training and testing dataset.

-   Understanding how categorical data such as the sex of a person (female/male) can be transformed into numerical dummy variable.

-   Being able to distinguish between dummy encoding and one-hot encoding

-   Using `tidymodels` including model design and data pre-processing (recipes) to analyze housing prices.

## Jumping Right Into It

### Univariate OLS with a Real World Dataset

**Data Description:**

::: incremental
-   King County House Sale dataset (Kaggle 2015). House sales prices from May 2014 to May 2015 for King County in Washington State.

-   Several predictor variables. For now we use only $Sqft$

-   We will only use 100 randomly chosen observations from the total of 21,613 observations.

-   We only use Sqft as predictor variable for now.
:::

## Loading the Data and Assigning Training and Testing Data (manually)



```{r}



DataHouses=
  import("https://ai.lange-analytics.com/data/HousingData.csv") |>
  clean_names("upper_camel") |>
  select(Price, Sqft=SqftLiving) 

# Manually generating DataTrain and DataTest
set.seed(7771)
DataTrain= sample_n(DataHouses, 100)
DataTest= sample_n(DataHouses, 50)
head(DataTrain)

```


## From Unfitted to Fitted Model

### How does the Unfitted Model Looks Like?

$$
\underbrace{\widehat{Price}}_\widehat{y}=\underbrace{\beta_1}_m \underbrace{Sqft}_x + \underbrace{\beta_1}_b
$$

### Fitting the Model with Tidymodels

```{r}
RecipeHouses= recipe(Price~Sqft, data=DataTrain)

ModelDesignOLS= linear_reg() %>% 
                set_engine("lm") %>% 
                set_mode("regression")

WFModelHouses = workflow() %>%  
  add_recipe(RecipeHouses) %>% 
  add_model(ModelDesignOLS) %>% 
  fit(DataTrain)

tidy(WFModelHouses)
```

## Unfitted Model vs Fitted Workflow Model {.smaller}

Unfitted Model: $$
\underbrace{\widehat{Price}}_\widehat{y}=\underbrace{\beta_1}_m \underbrace{Sqft}_x + \underbrace{\beta_1}_b
$$

```{r}
tidy(WFModelHouses)
```

Fitted Model: $$
\underbrace{\widehat{Price}}_\widehat{y}=\underbrace{240}_m \cdot\underbrace{Sqft}_x + \underbrace{52509}_b
$$

## Interpretation and Significance 

```{r}
#| echo: false
tidy(WFModelHouses)
```

$$
\begin{align}
\widehat{Price}&=240 \cdot Sqft +  52509\\
 (+240)&=240\cdot (+1) +  (+0)\\
 (+480)&=240\cdot (+2) +  (+0)\\
 (+720)&=240\cdot (+3) +  (+0)
\end{align}
$$ **For each extra** $Sqft$ the predicted price increases by \$240

**The variable** $Sqft$ is significant. I.e., the probability that the related coefficient $\beta_1$ equals zero is extremely small.


## Evaluating Predictive Quality with the Testing Dataset

```{r}
#| code-fold: false
DataTestWithPred=augment(WFModelHouses, new_data=DataTest)
metrics(DataTestWithPred, truth=Price, estimate=.pred)
```

## Univariate Linear Regression - Data Table and Goal {.smaller}

::: columns
::: {.column width="40%"}
**The Regression:**

$$
\widehat{Grade}_{i} = \beta_{1}StudyTime_{i}+\beta_{2}
$$

$$
\widehat{y}_{i} = \beta_{1}x_{i}+\beta_{2}
$$

**The Goal**

Find values for $\beta_1$ and $\beta_2$ that minimize the prediction errors $(\widehat{y}_{i}-y_i)^2$
:::

::: {.column width="60%"}
**The Data Table**

```{r }
#| echo: false
kbl(DataMockup |> mutate(i=1:5) |> select(i,everything()), 
    caption="Mockup Training Dataset")|>
  add_header_above(c(" ", "y", "x"), escape=F) |> 
  kable_styling(bootstrap_options=c("striped","hover"), full_width = F, position="center")
```
:::
:::

## Univariate Linear Regression - Data Diagram and Goal {.smaller}

::: columns
::: {.column width="40%"}
**The Regression:**

$$
\widehat{y}_{i} = \beta_{1}x_{i}+\beta_{2}
$$

**The Goal**

Find values for $\beta_1$ and $\beta_2$ that minimize the prediction errors $(\widehat{y}_{i}-y_i)^2$
:::

::: {.column width="60%"}
**The Data Diagram**

```{r}
Model123=lm(Grade~StudyTime, data=DataMockup)
PredGrade=predict(Model123, DataMockup)
ggplot(DataMockup, aes(x=StudyTime,y=Grade)) +
  geom_line(aes(y=PredGrade), color="red", size=2.7)+
  geom_point(size=5, color="blue")+
  geom_point(aes(y=PredGrade), color="black", size=2.7)+
  geom_segment(aes(x = StudyTime, y = PredGrade,
                   xend = StudyTime, yend = Grade),size=1.2)+
  scale_x_continuous("Study Time", breaks=seq(1,8))+
  scale_y_continuous(limits=c(65,110), breaks=seq(60,100,5))
```
:::
:::

# How to Measure Prediction Quality

```{=tex}
\begin{eqnarray*}
MSE & = & \frac{1}{N} \sum_{i=1}^{N}(\widehat{y}_{i}-y_{i})^{2} \\
 & \Longleftrightarrow& \\
MSE & =  & \frac{1}{N} \sum_{i=1}^{N}(\underbrace{\overbrace{\beta_{1}x_{i}+\beta_2}^{\mbox{Prediction $i$}}-y_i}_{\mbox{Error $i$}})^2
\end{eqnarray*}
```
**Note, when the data are given (i.e.,** $x_i$ and $y_i$ are given), the $MSE$ depends only on the choice of $\beta_1$ and $\beta_2$ »

## How to Measure Prediction Quality with the MSE {.smaller .scrollable}

```{=tex}
\begin{eqnarray}
MSE & = & \frac{(\beta_1x_{1}+\beta_2-y_1)^2
       +(\beta_1x_{2}+\beta_2-y_2)^2 +  \cdots+
       (\beta_1x_{5}+\beta_2-y_{5})^2}{5} \nonumber \\
       & \Longleftrightarrow& \nonumber \\
MSE & = & \frac{1}{5}\left[ (\underbrace{\overbrace{\beta_1\cdot 2+\beta_2}^{\mbox{Prediction $1$}}-65}_{\mbox{Error $1$}})^2
       +(\underbrace{\overbrace{\beta_1\cdot 3+\beta_2}^{\mbox{Prediction $2$}}-82}_{\mbox{Error $2$}})^2\right.\nonumber \\
   &   &\nonumber \\
       && +
    (\underbrace{\overbrace{\beta_1\cdot 7+\beta_2}^{\mbox{Prediction $3$}}-93}_{\mbox{Error $3$}})^2 
+(\underbrace{\overbrace{\beta_1\cdot 8+\beta_2}^{\mbox{Prediction $4$}}-93}_{\mbox{Error $4$}})^2\nonumber \\
   &   &\nonumber \\
       && +\left. (\underbrace{\overbrace{\beta_1\cdot 4+\beta_2}^{\mbox{Prediction $5$}}-83}_{\mbox{Error $6$}})^2\right]
\end{eqnarray}
```

## Custom R Function to Calculate MSE

```{r}
#| echo: false
FctMSE=function(VecBeta, data){
                        Beta1=VecBeta[1]
                        Beta2=VecBeta[2]
                        data=data |>
                        rename(Y=1, X=2) |> 
                        mutate(YPred=Beta1*X+Beta2) |>
                        mutate(Error=YPred-Y) |>
                        mutate(ErrorSqt=Error^2)
                        
                        MSE=mean(data$ErrorSqt)
                        
                        return(MSE)}
```

Function Call:

```{r}
#| echo: true
VecBetaTest=c(4,61)
ResultMSE=FctMSE(VecBetaTest, DataMockup)
print(ResultMSE)
```

Function Definition:»

```{r}
#| echo: true
FctMSE=function(VecBeta, data){
                        Beta1=VecBeta[1]
                        Beta2=VecBeta[2]
                        data=data |>
                        rename(Y=1, X=2) |> 
                        mutate(YPred=Beta1*X+Beta2) |>
                        mutate(Error=YPred-Y) |>
                        mutate(ErrorSqt=Error^2)
                        
                        MSE=mean(data$ErrorSqt)
                        
                        return(MSE)}
```

## How to Find Optimal Values for $\beta_1$ and $\beta_2$

**Method 1:**

:   **Calculate optimal values** for the parameters (the $\beta s$) based on Ordinary Least Squares (OLS) using two formulas (**Note,** this method works only for linear regression)

**Method 2:**

:   We can use a **systematic trial and error process**.

## Method 1: Calculate Optimal Parameters (only for OLS!)  {.smaller} 

```{=tex}
\begin{eqnarray*} 
\beta_{1,opt}&=& \frac
{N \sum_{i=1}^N y_i x_i- \sum_{i=1}^N y_i \sum_{i=1}^N x_i} 
{N \sum_{i=1}^N x_i^2 - \left (\sum_{i=1}^N x_i \right )^2}=3.96\\
&& \nonumber \\
\beta_{2,opt.}&=& \frac{ \sum_{i=1}^N y_i  - \beta_1 \sum_{i=1}^N x_i} {N} = 64.18
\end{eqnarray*}
```
```{r}
#| echo: false
DataTable=DataMockup |>
  mutate(GradeXStudyTime=Grade*StudyTime) |>
  mutate(StudyTimeSquared=StudyTime^2) 


kbl(DataTable |> mutate(i=1:5) |> select(i,everything()), 
    caption="Mockup Training Dataset ")|>
  add_header_above(c(" ", "y", "x", "y x","x x"), escape=F) |> 
  kable_styling(bootstrap_options=c("striped","hover"), full_width = F, position="center")
```

```{r}
#| echo: false
kbl(data.frame(as.list(colSums(DataTable))), 
    caption="Column Sums")|>
    kable_styling(bootstrap_options=c("striped","hover"), full_width = F, position="center")
```

## Method 2: Use a Systematic Trial and Error Process 🤓

-   **Grid Search (aka Brute Force):**

    1.  For a given range of $\beta_1$ and $\beta_2$ values, build a table with pairs of all combinations of these $\beta s$.
    2.  Then use our custom `FctMSE()` command to calculate a $MSE$ for each $\beta$ pair.
    3.  Find the $\beta$ pair with the lowest $MSE$

-   **Optimizer:** Use the R build-in optimizer. Push the start values for $\beta_1$ and $\beta_2$ together with the data to the optimizer as arguments. The rest is done by the optimizer.

-   See the R script in the footnote to see both algorithms in action.»

::: footer
See: [LinRegrScript.R script100](https://econ.lange-analytics.com/RScripts/LinRegrScript.R)
:::

# Multivariate OLS with a Real World Dataset

## Multivariate OLS with a Real World Dataset

**Data**

```{r}
#| echo: true
library(rio)
DataHousing =
  import("https://ai.lange-analytics.com/data/HousingDataSmall.csv")
```

-   King County House Sale dataset (Kaggle 2015). House sales prices from May 2014 to May 2015 for King County in Washington State.
-   Several predictor variables. 
-   We will use all 21,613 observations.



## Multivariate Analysis --- Three Predictor Variables {.scrollable}

`Sqft`: Living square footage of the house

`Grade` Indicates the condition of houses (1 (worst) to 13 (best))

`Waterfront`: Is house located at the waterfront (`yes` or `no`)

```{r}
#| echo: true
library(tidyverse);library(rio);library(janitor);library(tidymodels)
DataHousing =
  import("https://ai.lange-analytics.com/data/HousingData.csv")|>
  clean_names("upper_camel") |>
  select(Price, Sqft=SqftLiving, Grade, Waterfront)
```

**Unfitted Model:** » $$
Price=\beta_1 Sqft+\beta_2 Grade+\beta_3 Waterfront_{yes} +\beta_4
$$

## Multivariate Real World Dataset --- Splitting {.scrollable}

```{r}
#| echo: true
set.seed(777)
Split7030=initial_split(0.7,data=DataHousing, strata = Price, breaks = 5)
DataTrain=training(Split7030)
DataTest=testing(Split7030)
```



**DataTrain**

```{r}
#| echo: false
head(DataTrain)
```

**DataTest**

```{r}
#| echo: false
head(DataTest)
```

## Dummy and One-Hot Encoding {.scrollable}

**One-Hot Encoding**

```{r}
OneHotTable=tibble(Waterfront_yes=c(0,0,1,0),Waterfront_no=c(1,1,0,1))
print(OneHotTable)
```

*One-hot* encoding is easier to interpret but causes problems in OLS (dummy trap) because one variable is redundant. We can calculate one variable from the other (*perfect multicollinearity)*:

$$Waterfront_{yes}=1-Waterfront_{no}$$

## Dummy and One-Hot Encoding {.scrollable}

**Dummy Coding**

We use one variable less than we have categories. Waterfront has two categories. Therefore, we use one variable (e.g., `Waterfront_yes`):

**Dummy Encoding Example**

```{r}
DummyTable=tibble(Waterfront_yes=c(0,0,1,0))
print(DummyTable)
```

**Note,** dummy encoding can be done with `step_dummy()` in a *tidymodels recipe*.»

## Multivariate Analysis --- Building the Recipe

```{r}
#| echo: true
#| code-fold: false
RecipeHouses=recipe(Price ~ ., data=DataTrain) |> 
                    step_dummy(Waterfront)
```

Here is how the recipe later on (in the workflow) transforms the data:

```{r}
juice(RecipeHouses |> prep()) |> head()
```

## Multivariate Analysis --- Building the Model Design

**Unfitted Model:**

```{r}
#| echo: true
#| code-fold: false
ModelDesignHouses=linear_reg() |> 
  set_engine("lm") |> 
  set_mode("regression")
print(ModelDesignHouses)
```

»

## Multivariate Analysis --- Creating Workflow & Fitting to the Training Data

```{r}
#| echo: true
WFModelHouses = workflow() |>  
      add_recipe(RecipeHouses) |> 
      add_model(ModelDesignHouses) |> 
      fit(DataTrain)
tidy(WFModelHouses)
glance(WFModelHouses)
```

## Multivariate Analysis --- Predicting Testing Data and Metrics

```{r}
#| echo: true
DataTestWithPredictions = augment(WFModelHouses, new_data=DataTest)
metrics(DataTestWithPredictions, truth=Price, estimate=.pred)
```


## Exercise

**Run the Analysis**»

<https://ai.lange-analytics.com/exc/?file=05-LinRegrExerc100.Rmd>