---
title: "Digital Acyclical Graphs (DAGs)"
subtitle: "Graph Correlations Between Your Variables"

author:
  name: "Carsten Lange"
  email: "clange@cpp.edu"
  affiliation: "Cal Poly, Pomona"

execute:
  message: false
  warning: false

format: 
  revealjs:
    warning: false
    controls: true
    chalkboard:
      theme: whiteboard
      boardmarker-width: 2.7
    incremental: false
    scrollable: true
    footer: "<a href='https://dagitty.net'>https://dagitty.net</a> "


---





## What Will You Learn

-   Why is is important to graph the (assumed) correlation between all variables that will be part of the analysis
-   How to create a DAG

-   The very basics of backdoor paths



## How to Generate a Basic Graph

Assume: 

- `Stress` causes `Cancer`
- `Stress` causes `Smoking`
- `Smoking` causes `Cancer`

We want to analyze how `Smoking` influences `Cancer`, which makes `Cancer` our **Outcome**. The *predictor* `Smoking` becomes our **Exposure**:

## How to Generate a Basic Graph

```{r}
#| code-fold: true
library(tidyverse)
library(dagitty)
library(ggdag)
set.seed(123)
dagify(Cancer~Smoking+Stress,
       Smoking~Stress,
       outcome = "Cancer",
       exposure = "Smoking"
      ) |> 
ggdag_status(node_size = 22) + theme_void()
```
An online tool to create a DAG is available at: [https://dagitty.net/](https://dagitty.net/)


##  Use Your Domain Knowledge (beliefs) to Complete The Dag Below

*Directed Acylical Graphs (DAG's)* can help to structure our beliefs.

**Incomplete DAG**
```{r}
#| code-fold: true
library(tidyverse)
library(ggdag)
set.seed(123)
dagify(Wage~SexFem+Ten+Educ+Talent,
       outcome = "Wage",
       exposure = "SexFem"
      ) |> 
ggdag_status(node_size = 22) + theme_void()
```

Complete the DAG here: [https://dagitty.net/mfBystKCe](https://dagitty.net/mfBystKCe)