---
title: "Standard Deviation and Normal Distribution"
subtitle: "Review"
format: 
  revealjs:
    code-fold: true
    incremental: false
    scrollable: true
    echo: true
---

## What Will You Learn {.scrollable .smaller}

```{r}
#| echo: false
library(TeachHist)

```

-   You will review the concept of *Range*, *Absolute Deviation*, *Variance*, and *Standard Deviation* to measure the spread of a variable.

-   You will review histogram diagrams

-   You will review the concept of *Standard Deviation* to measure the spread of a variable in connection with the Normal Distribution.

-   You will use the *Normal Distribution* to predict individual variables.

-   You will use the *Normal Distribution* to predict properties of a sample such as the *Mean*.

-   You will distinguish between *Standard Deviation* and *Standard Error* and learn how to calculate the *Standard Error*

-   You will distinguish between *Normal Distribution* and *t-Distribution*

-   You will calculate a *Confidence Interval*

-   You be able to understand and perform a *Hypothesis Test*

## Some Questions to Introduce Center and Spread

::: incremental
-   Who runs faster, men or women?
    -   Is this true for all men and women?
-   Who is taller, men or women?
    -   Is this true for all men and women?
-   What is it the average temperature in Michigan/Coastal California?
    -   Why is the climate different in the two states?
-   Which of your products shows high or low volatility in sales?
    -   Why is that?
    -   Why is it important?
:::

## How to measure Center and Spread {.smaller}

### (1,000 synthetic observations each)

::: columns
::: {.column width="70%"}
```{r}
#| echo: false
library(rio)
library(tidyverse)
DataDerivative=import("DataDerivCryptoEuro.csv")
Plot1=ggplot(DataDerivative, aes(y=1,x=CryptoDeriv))+
       geom_jitter()+
       geom_vline(xintercept = mean(DataDerivative$CryptoDeriv), , color="red", linewidth=2)+
       labs(y="", x="Crypto Derivative in $")+
       scale_x_continuous(limits = c(10,40))+
       scale_y_continuous(labels = NULL)
Plot2=ggplot(DataDerivative, aes(y=1,x=EuroDeriv))+
       geom_jitter()+
       geom_vline(xintercept = mean(DataDerivative$EuroDeriv), , color="red", linewidth=2)+
       labs(y="", x="Euro Derivative in $")+
       scale_x_continuous(limits = c(10,40))+
       scale_y_continuous(labels = NULL) 
Plot1
```

```{r}
#| echo: false
Plot2
```
:::
:::

## How to measure Center and Spread {.smaller}

### (1,000 synthetic observations each)

::: columns
::: {.column width="60%"}
```{r}
#| echo: false
Plot1
```

```{r}
#| echo: false
Plot2
```
:::

::: {.column width="40%"}
$Range=x_{max}-x_{min}$

$AbsDev=\sum_{i=1}^{1000}|x_i-\overline {x}|$

$Var=\sum_{i=1}^{1000}(x_i-\overline {x})^2$

$Std=\sqrt{Var}$

```{r}
RangeCrypto=
  max(DataDerivative$CryptoDeriv)-min(DataDerivative$CryptoDeriv)
MADCrypto=mad(DataDerivative$CryptoDeriv, 
          center=mean(DataDerivative$CryptoDeriv), 
          constant=1)
VarCrypto=var(DataDerivative$CryptoDeriv)
StdCrypto=sd(DataDerivative$CryptoDeriv)

RangeEuro=
  max(DataDerivative$EuroDeriv)-min(DataDerivative$EuroDeriv)
MADEuro=mad(DataDerivative$EuroDeriv, 
          center=mean(DataDerivative$EuroDeriv), 
          constant=1)
VarEuro=var(DataDerivative$EuroDeriv)
StdEuro=sd(DataDerivative$EuroDeriv)
```

**Spread Crypto vs. Euro derivatives:**

Range: `r RangeCrypto` vs. `r RangeEuro`

MAD: `r MADCrypto` vs `r MADEuro`

Var: `r VarCrypto` vs `r VarEuro`

Std: `r StdCrypto` vs `r StdEuro`
:::
:::

## Measuring Female Height Spread for a Small Sample

::: columns
::: {.column width="60%"}
```{r echo=FALSE}
#| code-fold: true
DataSample=data.frame(Height=c(63, 66, 68, 70))
ggplot(DataSample, aes(x=Height, y=c(1:4)))+
  geom_vline(xintercept=mean(DataSample$Height), size=2)+
  geom_point(color="red", size=3)+
  ylab("")+
  scale_x_continuous(breaks=seq(63,70))+
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank() 
        )
```
:::

::: {.column width="40%"}
Sample Mean: $\bar x=???$

Range= ???

Absolute Deviation = ???

Sample Variance: $Var=\frac{\sum_{i=1}^N (x_i-\bar x)^2}{N-1}=???$

Sample Standard Deviation: ???
:::
:::

## How are Critical Chlosterol Values and other Blood Test Values Determined {.smaller}

**Assumption:** Cholesterol values are *normal distributed.*

**Normal Distribution:**

-   Most values are in the center.
-   Some are larger or smaller than the center.
-   Few are a lot larger or smaller than the center.
-   Very few are extremely large or smaller
-   The distribution is symmetric around the mean, leading to a bell curve.

For women 45 -- 54 years, the mean is 230.5 mg/100 ml cholesterol, and the standard deviation is 50.0 (source: <https://www.cdc.gov/nchs/data/series/sr_11/sr11_022.pdf>)

## Synthetic Data for 10,000 Women (45 - 54 years)

```{r}
library(TeachHist)
TeachHistRelFreq(Mean=230.5, Sd=50, SeedValue=123, NOfSimData=10000)
```

**Empirical Rule**, also known as the $68-95-99.7$ rule.

. . .

**0.95 Confidence Interval:** <br> $Mean \pm 1.96 \cdot StdDev$ <br> In this example $230.5 - 50 \cdot 1.96= 182.5$ and $230.5 + 50\cdot 1.96= 378.5$ <br> 95% of all 45 - 54 year women have a Cholesterol level higher than 182.5 and lower than 378.5.

## Synthetic Data for 10,000 Women (45 - 54 years)

### Histogram with Relative Frequencies as Column Area

```{r}
TeachHistDens(Mean=230.5, Sd=50, SeedValue=123, NOfSimData=10000, PlotNormCurv = FALSE)
```

Try column just right of mean: $0.0068\cdot 50=0.34$

Try changing the `BinWidth` for `TeachHistDensity()` and `TeachHistRelFreq()`

## Synthetic Data for 10,000 Women (45 - 54 years)

### Histogram with Relative Frequencies as Column Area and Normal Curve

```{r}
#| echo: false
TeachHistDens(Mean=230.5, Sd=50, SeedValue=123, NOfSimData=10000, PlotNormCurv = TRUE)
```

## Normal Curve {.smaller}

-   Most values are in the center.
-   Some are larger or smaller than the center.
-   Few are a lot larger or smaller than the center.
-   Very few are extremely large or small.
-   The distribution is symmetric around the mean, leading to a bell curve.

$$
Density = \frac{1}{\sqrt{2\pi\sigma^2}} e^{-\frac{(x - \mu)^2}{2\sigma^2}}
$$

where: $\mu$ is the mean and $\sigma$ is the standard deviation

```{r}
#| echo: false
ggplot(data.frame(x = c(30.5, 430.5)), aes(x = x)) +
stat_function(fun = dnorm, args = list(mean = 230.5, sd = 50)) +
labs(title = "Normal Distribution", x = "x", y = "Density")+
scale_x_continuous(breaks=seq(0,450,25))+
geom_vline(xintercept = 230.5, color="red")
```

## Synthetic Data for 10,000 Women (45 - 54 years)

### Histogram with Relative Frequencies as Column Area and Normal Curve {.smaller}

```{r}
#| echo: false
TeachHistDens(Mean=230.5, Sd=50, SeedValue=123, NOfSimData=10000, PlotNormCurv = TRUE)
```

`pnorm(x, mean=???, sd=???)` calculates the probability of getting a value smaller than x. **Try:**

-   `pnorm(480, mean=230.5, sd=50)`

-   `pnorm(130.5, mean=230.5, sd=50)`

-   `pnorm(330.5, mean=230.5, sd=50)-pnorm(130.5, mean=230.5, sd=50)`

## Normal Curve Women (45 - 54 years) with Mean: 230.5, SD=50

```{r}
#| echo: false
TeachHistDens(Mean=230.5, Sd=50, SeedValue=123, NOfSimData=10000, PlotNormCurv = TRUE)
```

`qnorm(x, mean=???, sd=???)` calculates the quantity for P percent of the values falling below that quantity.**Try:**

-   `qnorm(.5, mean=230.5, sd=50)`
-   `qnorm(.02, mean=230.5, sd=50)`
-   `qnorm(.05, mean=230.5, sd=50)` (meaning 5% are above this level)
-   `qnorm(.95, mean=230.5, sd=50)` (meaning 5% are above this level)
-   `qnorm(.977, mean=230.5, sd=50)- qnorm(.025, mean=230.5, sd=50)` (meaning 95% are in between this levels)

## Frequentists Statistics

*Frequentist Statistics* uses **Relative Frequency** to estimate **Probability**

## Applications/Exercise {.smaller}

Mean of male height in the U.S. is $70$ inches with a Standard Deviation of 3 inches ([click for source](https://tasks.illustrativemathematics.org/content-standards/tasks/1020#:~:text=The%20heights%20of%20adult%20men,standard%20deviation%20of%202.5%20inches):).

-   What is the probability to randomly meeting an U.S. male with a height smaller than 50 inches assuming height is normal distributed

-   What is the probability to randomly meeting an U.S. male with a height bigger than 80 inches assuming height is normal distributed

-   What is the probability to randomly meeting an U.S. male with a height bigger than 120 inches assuming height is normal distributed

-   What is the probability to randomly meeting an U.S. male with a height between 67 and 73 inches assuming height is normal distributed

-   What is the height range that 95% all men fall in

## Research about Groups (Samples) is Moe Important than Individuals {.smaller}

A group of 50 men are taking a class at Cal Poly. The average height in this class is 72 inches (remember the Mean of all men is $70$ inches and the Standard Deviation is $3$ inches.

. . .

**The mean of the sample is different than the mean of all men.** This should not come as surprise. If we had sampled another class, we would have measured a **different** mean.

. . .

**So, by how much would the *Sample Mean* spread, if we considered more courses?**
<br>
(Note, the spread of a sample mean is called *Standard Error*)

. . .

-   **Standard Error is less than the Standard Deviation of the population** (*mean of men height*). <br> 
**Reason:** In a group, tall and small men partially cancel out.

. . .



. . .

-   The *Standard Error* is smaller when the sample size ($N$) is bigger.






. . .

-   The *Standard Error* is bigger when the *Standard Deviation* of the population is bigger.



. . .

$$SE=\frac{StdDev}{\sqrt{N}}$$



## Back to the Male Students at Cal Poly

A group of 50 men are taking a class at Cal Poly. The average height in this class is 72 inches (remember the Mean of all men is $70$ inches and the Standard Deviation is $3$ inches).

Sample Mean: $\bar x =72$

Standard Error: $SE=\frac{Std}{\sqrt{N}}=\frac{3}{\sqrt{50}}=0.4242641$

```{r}
#| output-location: fragment
RelFreq=1-pnorm(72, mean = 70, sd=0.4242641)
cat("The relative frequency of a group of 50 men  (and thus the probibility)  having a mean of 72 inches or taller is", RelFreq)
```

## Another Application: Mean and Standard Deviation not Known {.smaller}

Our daily sales data for a new product are normally distributed. We do not know the *Mean* nor the *Standard Deviation*.

. . .

However, we have data from the last 36 days to estimate the *Mean* and *Standard Deviation* based on our sample data.

. . .

**Sample Size:** $N=36$

**Estimated Mean in (\$ 1,000):** $\bar x= 90$

**Estimated Standard Deviation:** $StdDev= 12$

**Estimated Standard Error:** $SE= \frac{12}{\sqrt{36}}=2$

## t-Distribution vs. Normal Distribution

**If the Standard Deviation is not known and is estimated from the sample:** <br> Normal Distribution changes to a t-Distribution

**However, Normal Distribution and t-Distribution are very similar with the t-Distribution having fatter tails.**



## t-Distribution vs. Normal Distribution ($N=10$)

```{r}
#| echo: false
DataPlot=data.frame(Sales=seq(90-5*2, 90+5*2, 0.1)) |> 
             mutate(SalesStd=(Sales-90)/2) |> 
             mutate(DensityNorm=dnorm(SalesStd)) |> 
             mutate(DensityT=dt(SalesStd, df=9))

ggplot(DataPlot, aes(x=Sales, y=DensityNorm))+
       geom_line(color="red", linewidth=2)+
       geom_line(aes(x=Sales, y=DensityT), color="blue", linewidth=1)
```

## t-Distribution vs. Normal Distribution ($N=36$)

```{r}
#| echo: false
DataPlot=data.frame(Sales=seq(90-5*2, 90+5*2, 0.1)) |> 
             mutate(SalesStd=(Sales-90)/2) |> 
             mutate(DensityNorm=dnorm(SalesStd)) |> 
             mutate(DensityT=dt(SalesStd, df=35))

ggplot(DataPlot, aes(x=Sales, y=DensityNorm))+
       geom_line(color="red", linewidth=2)+
       geom_line(aes(x=Sales, y=DensityT), color="blue", linewidth=1)
```

## Calculating the 95% Confidence Interval for Mean {.smaller}

```{r}
#| echo: false
set.seed(777)
DataSales=data.frame(Sales=rnorm(36, 90,12))
TeachHistDens(Mean = 90, Sd=2)
```

**Sample Size:** $N=36$
**Estimated Mean in (\$ 1,000):** $\bar x= 90$

**Estimated Standard Deviation:** $StdDev= 12$

**Estimated Standard Error:** $SE= \frac{12}{\sqrt{36}}=2$

Confidence Interval based on `qnorm()`:

```{r}
LowerBound=qnorm(0.025, 90, 2)
UpperBound=qnorm(0.975, 90, 2)
cat("95% Confidence Interval:", LowerBound, UpperBound)
```

. . .

Confidence Interval based on `t.test()`:

```{r}
t.test(DataSales$Sales)
```

## Hypothesis Test

**H0:** Before the sample of 36 days was taken, it was believed that the *Mean* for sales is 85?

**A (Research):** Research of the 36 days suggests that the *Mean* is different $(\bar x=90)$.

. . .

Is the mean from the sample just by accident different from 80 or is there more behind it? The latter would mean the H0 is likely wrong.

. . .

Let us work with 95% *Confidence*.

```{r}
TeachHistHypTest(NullHyp=85, StandardError=2, SampleMean=90, SeedValue=123)
```

## AB-Test {.smaller}

A very powerful and easy to use methodology to compare means of one or more *Treatment* groups to a *Control Group*. 

. . .

**Goal:** Determine if a treatment (e.g., number of conversions for a new  websites (*Treatment*)) is significant compared to the old website (*Control Group*).

. . .

**Methodology:** Two Sample t-Test

. . .

**Requirement:** Participants must be randomly assigned to the groups (no self-selection!).

. . .

**Problem:** While it is easy and technically straightforward to randomly assign website visitors to different webpages, it can be difficult or impossible in other cases. 

. . .

For example, it is not feasible to assign participants of a marketing event into a treatment group (drank a glass of champagne before the talk) and a control group (did not drink a glass of champagne before the talk)

## Example: Conversion Data (synthetic) for a New Website {.smaller}

New website is *Treatment* old website is *Control Group*

**Purchase value data from the last 100 visitors who purchased on each of the websites:**


```{r}
DataWebSites=import("DataWebSites.csv")[,2:3]
head(DataWebSites)
```

. . .

**Control Group:**
```{r}
cat("Mean of Control Group is:", mean(DataWebSites$PurchAmtControl))
cat("Standard Error of Control Group is:", sd(DataWebSites$PurchAmtControl)/sqrt(100))
```

. . .

**Treatment Group:**
```{r}
cat("Mean of Treatment Group is:", mean(DataWebSites$PurchAmtTreat))
cat("Standard Error of Treatment Group is:", sd(DataWebSites$PurchAmtTreat)/sqrt(100))
```

```{r}
library(rio)
DataWebSites=import("DataWebSites.csv")
```

## Two Sample t-Test

**Assumption:** Same Variance

**Distribution:** Normal

**Method:**

- Calculate difference between means

- Difference between means will also be normal distributed

- Standard error needs to be recalculated for difference and is approxiamtly double the one of the treatment/control group.

- **Hypothesis 0:** Difference is 0

- **Goal:** If Treatment website is better a difference of 0 should be very unlikely.

## Two Sample t-Test

```{r}
t.test(DataWebSites$PurchAmtControl,
       DataWebSites$PurchAmtTreat,
       var.equal=FALSE)
```


