---
title: "Standard Deviation and Normal Distribution"
subtitle: "Review"
format: 
  revealjs:
    code-fold: true
    incremental: false
    scrollable: true
    echo: true
---

## What Will You Learn {.scrollable .smaller}

```{r}
#| echo: false
library(kableExtra)

```



-   You will review the concept of Range, Absolute Deviation, Variance and Standard deviation to measure the spread of a variable.
-   You will review histogram diagrams
-   You will review the concept of Standard Deviation to measure the spread of a variable in more detail.
-   You will explain why Standard Deviation is so important for the concept of Normal Distribution.

## Some Questions to Introduce Center and Spread

::: {.incremental}
-   Who runs faster, men or women? 
    - Is this true for all men and women?

-   Who is taller, men or women? 
    - Is this true for all men and women?

-   What is it the average temperature in Michigan/Coastal California? 
    Why is the climate different in te two states?

-   Which of your products shows high or low volatility in sales? 
    - Why is that?
    - Why is it important?
:::



## How to measure Center and Spread {.smaller}

### (1,000 synthetic observations each)

::: columns
::: {.column width="70%"}
```{r}
#| echo: false
library(rio)
library(tidyverse)
DataDerivative=import("DataDerivCryptoEuro.csv")
Plot1=ggplot(DataDerivative, aes(y=1,x=CryptoDeriv))+
       geom_jitter()+
       geom_vline(xintercept = mean(DataDerivative$CryptoDeriv), , color="red", linewidth=2)+
       labs(y="", x="Crypto Derivative in $")+
       scale_x_continuous(limits = c(10,40))+
       scale_y_continuous(labels = NULL)
Plot2=ggplot(DataDerivative, aes(y=1,x=EuroDeriv))+
       geom_jitter()+
       geom_vline(xintercept = mean(DataDerivative$EuroDeriv), , color="red", linewidth=2)+
       labs(y="", x="Euro Derivative in $")+
       scale_x_continuous(limits = c(10,40))+
       scale_y_continuous(labels = NULL) 
Plot1
```

```{r}
#| echo: false
Plot2
```
:::
:::

## How to measure Center and Spread {.smaller}

### (1,000 synthetic observations each)

::: columns
::: {.column width="60%"}
```{r}
#| echo: false
Plot1
```

```{r}
#| echo: false
Plot2
```
:::

::: {.column width="40%"}
$Range=x_{max}-x_{min}$

$AbsDev=\sum_{i=1}^{1000}|x_i-\overline {x}|$

$Var=\sum_{i=1}^{1000}(x_i-\overline {x})^2$

$Std=\sqrt{Var}$

```{r}
RangeCrypto=
  max(DataDerivative$CryptoDeriv)-min(DataDerivative$CryptoDeriv)
MADCrypto=mad(DataDerivative$CryptoDeriv, 
          center=mean(DataDerivative$CryptoDeriv), 
          constant=1)
VarCrypto=var(DataDerivative$CryptoDeriv)
StdCrypto=sd(DataDerivative$CryptoDeriv)

RangeEuro=
  max(DataDerivative$EuroDeriv)-min(DataDerivative$EuroDeriv)
MADEuro=mad(DataDerivative$EuroDeriv, 
          center=mean(DataDerivative$EuroDeriv), 
          constant=1)
VarEuro=var(DataDerivative$EuroDeriv)
StdEuro=sd(DataDerivative$EuroDeriv)
```

**Spread Crypto vs. Euro derivatives:**

Range: `r RangeCrypto` vs. `r RangeEuro`

MAD: `r MADCrypto` vs `r MADEuro`

Var: `r VarCrypto` vs `r VarEuro`

Std: `r StdCrypto` vs `r StdEuro`
:::
:::

## Measuring Female Height Spread for a Small Sample

:::: columns
::: {.column width="60%"}

```{r echo=FALSE}
#| code-fold: true
DataSample=data.frame(Height=c(63, 66, 68, 70))
ggplot(DataSample, aes(x=Height, y=c(1:4)))+
  geom_vline(xintercept=mean(DataSample$Height), size=2)+
  geom_point(color="red", size=3)+
  ylab("")+
  scale_x_continuous(breaks=seq(63,70))+
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank() 
        )
```

:::

:::: columns
::: {.column width="40%"}

Sample Mean: $\bar x=???$

Range= ???

Absolute Deviation = ???

Sample Variance: $Var=\frac{\sum_{i=1}^N (x_i-\bar x)^2}{N-1}=???$

Sample Standard Deviation: ???

:::

::::


## How are Critical Diabetes/Chlosterol Values and other Blood Test Values Determined {.smaller}

Diabetics values are *normal distributed.*

**Normal Distribution:**

- Most values are in the center. 
- Some are larger or smaller than the center. 
- Few are a lot larger or smaller than the center. 
- Very few are extremely large or smaller
- The distribution is symmetric around the mean, leading to a bell curve.

For women 45 -- 54 years, the mean is 230.5 mg/100 ml cholesterol, and the standard deviation is 50.0 (source: <https://www.cdc.gov/nchs/data/series/sr_11/sr11_022.pdf>)

## Synthetic Data for 10,000 Women (45 - 54 years)

### Histogram with Relative Frequencies as Column Height

```{r}
library(TeachHist)
TeachHistRelFreq(Mean=230.5, Sd=50, SeedValue=123, NOfSimData=10000)
```

Empirical rule, also known as the 68-95-99.7 rule.

## Synthetic Data for 10,000 Women (45 - 54 years)

### Histogram with Relative Frequencies as Column Area

```{r}
TeachHistDens(Mean=230.5, Sd=50, SeedValue=123, NOfSimData=10000, PlotNormCurv = FALSE)
```

Try column just right of mean: 0.0068\*50=0.34

## Synthetic Data for 10,000 Women (45 - 54 years)

### Histogram with Relative Frequencies as Column Area and Normal Curve

```{r}
#| echo: false
TeachHistDens(Mean=230.5, Sd=50, SeedValue=123, NOfSimData=10000, PlotNormCurv = TRUE)
```

## Normal Curve {.smaller}

- Most values are in the center. 
- Some are larger or smaller than the center. 
- Few are a lot larger or smaller than the center. 
- Very few are extremely large or small.
- The distribution is symmetric around the mean, leading to a bell curve.

$$
Density = \frac{1}{\sqrt{2\pi\sigma^2}} e^{-\frac{(x - \mu)^2}{2\sigma^2}}
$$

where:
- $\mu$ is the mean
- $\sigma$ is the standard deviation

```{r}
#| echo: false
ggplot(data.frame(x = c(30.5, 430.5)), aes(x = x)) +
stat_function(fun = dnorm, args = list(mean = 230.5, sd = 50)) +
labs(title = "Normal Distribution", x = "x", y = "Density")+
scale_x_continuous(breaks=seq(0,450,25))+
geom_vline(xintercept = 230.5, color="red")
```


## Synthetic Data for 10,000 Women (45 - 54 years)

### Histogram with Relative Frequencies as Column Area and Normal Curve {.smaller}

```{r}
#| echo: false
TeachHistDens(Mean=230.5, Sd=50, SeedValue=123, NOfSimData=10000, PlotNormCurv = TRUE)
```

`pnorm(x, mean=???, sd=???)` calculates the probability of getting a value smaller than x. **Try:**

- `pnorm(480, mean=230.5, sd=50)`

-   `pnorm(130.5, mean=230.5, sd=50)`

-   `pnorm(330.5, mean=230.5, sd=50)-pnorm(130.5, mean=230.5, sd=50)`

## Normal Curve Women (45 - 54 years) wit Mean: 230.5, SD=50

```{r}
#| echo: false
TeachHistDens(Mean=230.5, Sd=50, SeedValue=123, NOfSimData=10000, PlotNormCurv = TRUE)
```

`qnorm(x, mean=???, sd=???)` calculates the quantity for P percent of the values falling below that quantity.**Try:**

-   `qnorm(.5, mean=230.5, sd=50)`
-   `qnorm(.02, mean=230.5, sd=50)`
-   `qnorm(.05, mean=230.5, sd=50)`
-   `qnorm(.95, mean=230.5, sd=50)` (meaning 5% are above this level)

## Distribution Other Than Normal Distribution

- Bernoulli 
- Uniform