---
title: "Standard Deviation and Normal Distribution"
subtitle: "Review"
format: 
  revealjs:
    code-fold: false
    incremental: false
    scrollable: true
    echo: true
---

```{r}
library(kableExtra)

```

## What Will You Learn {.scrollable .smaller}

-   You will review the concept of Variance to measure the spread of a variable.
-   You will review the concept of Standard Deviation to measure the spread of a variable.
- You will explain  why Standard Deviation is so important for the concept of Normal Distribution.

 

## The Data {.non-incremental}

**Source:**

- 2012 US Army Anthropometric Survey (ANSUR II) 
- DCPH (Defense Center for Public Health) - Aberdeen
- https://phc.amedd.army.mil/topics/workplacehealth/ergo/Pages/Anthropometric-Database.aspx
- 6068 observations (4082 male, 1986 female) with 108 measures (variables/columns)

**Download:**

```{r}
#| echo: true
library(rio)
library(janitor)
library(tidymodels)
DataSoldiersM=import("https://econ.lange-analytics.com/RData/Datasets/DataUSArmyBodyMeasures.xlsx", sheet="DataMale")|> 
              clean_names("upper_camel") |> 
              select(Height=Heightin, Weight=Weightlbs, Gender)
DataSoldiersF=import("https://econ.lange-analytics.com/RData/Datasets/DataUSArmyBodyMeasures.xlsx", sheet="DataFemale") |> 
              clean_names("upper_camel") |> 
              select(Height=Heightin, Weight=Weightlbs, Gender)

DataSoldiers=merge(DataSoldiersF, DataSoldiersM, all=TRUE)
head(DataSoldiersF)
head(DataSoldiersM)
head(DataSoldiers) #where are the males?
```

## Plot Height Male and Female Point Plot 

```{r}
ggplot(DataSoldiers, aes(x=Gender, y=Height, color=Gender))+
  geom_point()
  
```


## Plot Height Male and Female Jitter Plot

```{r}
ggplot(DataSoldiers, aes(x=Gender, y=Height, color=Gender))+
  geom_jitter()
  
```

## Plot Height Male and Female Jitter Plot alpha=0.1)

```{r}
ggplot(DataSoldiers, aes(x=Gender, y=Height, color=Gender))+
  geom_jitter(alpha=0.1)
```



# Height Regression

```{r}
RecipeHeightGender=recipe(DataSoldiers, Height~Gender) |> 
                   step_dummy(Gender)
ModelDesignLinear=linear_reg() |> 
                     set_engine("lm") |> 
                     set_mode("regression")
WFModelStudyTime= workflow() |> 
                  add_recipe(RecipeHeightGender) |> 
                  add_model(ModelDesignLinear) |> 
                  fit(DataSoldiers)
tidy(WFModelStudyTime)
```


