---
title: "Multi Variable Regression" 
author:
  name: "Carsten Lange"
  email: "clange@cpp.edu"
  affiliation: "Cal Poly, Pomona"

execute:
  message: false
  warning: false

format: 
  live-revealjs:
    theme: [moon, ../../CustomCL.scss]
    warning: false
    controls: true
    chalkboard:
      theme: whiteboard
      boardmarker-width: 2.7
    incremental: false
    scrollable: true
webr:
  packages:
    - dplyr
    - janitor
    - ggplot2
    - wooldridge
    - corrplot
  
engine: knitr    
---

{{< include ../../_extensions/r-wasm/live/_knitr.qmd >}}



## What Will You Learn {.scrollable .smaller}

```{webr}
#| echo: false
#| context: setup
DataHouseSqftCond=read.csv(
  "https://econ.lange-analytics.com/calcat/FrischWaughLovellRegr/Data/DataDiagram.csv") |>   rename(Cond=Grade)

ModelSqftCond97Cor=read.csv(
  "https://econ.lange-analytics.com/calcat/FrischWaughLovellRegr/Data/DataDiagram97Cor.csv")

DataWage=wage1 |> 
                 clean_names("upper_camel") |>  
                 select(Wage, SexFem=Female, Tenure, Exper, 
                        Educ, NonWhite=Nonwhite, MSA=Smsa, 
                        Married, NumDep=Numdep) |> 
                 mutate(SglPar=
                        ifelse((Married==0 & NumDep>0),1,0)) |> 
                 mutate(Married=NULL, NumDep=NULL)
```





- Using `lm`  to run a multivariate regression
- Interpret regression coefficients
- Interpret p-values
- Analyze pairwise correlation 







## Frisch-Waugh-Lovell Theorem {.smaller}

**Frisch-Waugh-Lovell (FWL) Theorem in a Nutshell:**

When the predictor variables in an *OLS* regression are **not correlated**,
we can run $j$ (e.g., two) Simple Regressions (regressions with only one
predictor variable) and use the resulting $\beta$-coefficients for a
joint multivariate regression.

**Example:** Predicting house $Price$ based on $Sft$ and $Cond$ (condition of a house with 0 (worst) -- 20 (best)):

$$\widehat{Price}_i= \beta_{Sqft} \cdot Sqft_i + \beta_{01}$$
$$\widehat{Price}_i= \beta_{Cond} \cdot Cond_i + \beta_{02}
$$
$$\widehat{Price}_i= \beta_{Sqft} \cdot Sqft_i+\beta_{Cond} \cdot Cond_i + \beta_0
$$

The resulting $\beta_0$ can be calculated based on the average of Price,
Square Feet, and Condition, <br> <br> $\overline{Price}$,
$\overline{Sqft}$ and $\overline{Cond}$, respectively.

$$\beta_0=\overline{Price}-\beta_{Sqft} \cdot \overline{Sqft} - \beta_{Cond} \overline{Cond}$$

## FWL Theorem with a Synthetic Dataset

We have a dataset for house prices ($Price_i$) together with predictor
variables for the houses' condition ($Cond_i$; 1 -- 20) and their square
footage ($Sqft_i$; 1000 -- 3100).

$Cond_i$ and $Sqft_i$ are not correlated:

```{webr}
DataHouseSqftCond=read.csv(
  "https://econ.lange-analytics.com/calcat/FrischWaughLovellRegr/Data/DataDiagram.csv") |>   
                  rename(Cond=Grade)

cat("Correlation between Sqft and Cond =",
    cor(DataHouseSqftCond$Sqft, DataHouseSqftCond$Cond))
ggplot(aes(x=Cond, y=Sqft), data=DataHouseSqftCond)+
geom_point()
```





## Running the Two Simple Regressions in R

$$\widehat{Price_i}=\beta_1 \widehat{Sqft}+\beta_0$$

```{webr PartRegressionUncorSqft}
ModelPriceSqft=lm(Price~Sqft,data=DataHouseSqftCond)
print(ModelPriceSqft)
```

$$\widehat{Price_i}=\beta_1 \widehat{Cond}+\beta_0$$

```{webr PartRegressionUncorCond}
ModelPriceCond=lm(Price~Cond,data=DataHouseSqftCond)
print(ModelPriceCond)
```

## Confirming the results with a Multivariate Regression {.smaller}

From the Simple Regressions we got:

$\beta_{Sqft}=213$ and $\beta_{Cond}=15,018$

From the multivariate regression we get:

```{webr ModelSqftCond}
ModelSqftCond=lm(Price~Sqft+Cond, data = DataHouseSqftCond)
summary(ModelSqftCond)
```

$$\widehat{Price}_i= 213 \cdot Sqft + 15,018 \cdot Cond_i + 298,100$$

# Correlated Predictor Variables <br>
(FWL Theorem does not work!)

## FWL Theorem with Another Synthetic Dataset {.smaller}

We have another dataset for house prices ($Price_i$) together with
predictor variables for the houses' condition ($Rooms_i$; 1 -- 8) and
square footage ($Sqft_i$; 1000 -- 3000).

**$Rooms_i$ and $Sqft_i$ are correlated. Each room has about 400 Sqft:**

$$Sqft_i\approx 400\cdot Rooms + 0$$

```{webr LoadingDataHouseSqftRooms97CorNotHidden}
DataHouseSqftRooms97Cor=read.csv(
  "https://econ.lange-analytics.com/calcat/FrischWaughLovellRegr/Data/DataDiagram97Cor.csv") 

cat("Correlation between Sqft and Cond =",
    cor(DataHouseSqftRooms97Cor$Sqft, DataHouseSqftRooms97Cor$Rooms))

ggplot(aes(x=Rooms, y=Sqft), data=DataHouseSqftRooms97Cor)+
geom_point()
```

**Note**, because of predictor correlation $\neq$ 0. *FWL* Theorem is
not applicable.


## Running the two simple regressions in R

### Simple Regression between `Sqft` and `Price`:

$\beta_{Sqft}=216$ and $\beta_{01}=291,181$

```{webr ModelPriceSqft97Cor}
ModelPriceSqft97Cor=lm(Price~Sqft, data=DataHouseSqftRooms97Cor)
summary(ModelPriceSqft97Cor)
```

### Simple Regression between `Rooms` and `Price`:

$\beta_{Rooms}=84,471$ and $\beta_{02}=297,937$

```{webr ModelPriceRooms97Cor}
ModelPriceRooms97Cor=lm(Price~Rooms, data=DataHouseSqftRooms97Cor)
summary(ModelPriceRooms97Cor)
```

## Comparing  the Results between two simple regressions and Multivariate Regression {.smaller}

### Multiple Regression between `Sqft`, `Rooms` and `Price`:

**From the Two Simple Regressions we got:**
$\beta_{Sqft}=216$ and  $\beta_{Rooms}=84,471$

**From the Multivariate Regression we get:**

```{webr ModelSqftCond97Cor}
ModelSqftCond97Cor=lm(Price~Sqft+Rooms, data = DataHouseSqftRooms97Cor)
summary(ModelSqftCond97Cor)
```

## Summary - Steps for a Linear Regression {.smaller}

1. Find variables that have a causal relationship with the outcome variable and are correlated with the outcome variable

2. Eliminate predictor variables that are correlated with each other

1. Write down the formula for the unfitted model. Such as:
$$\widehat{Price}=\beta_1 Sqft + \beta_2 Cond +\beta_0$$

3. Run the regression

4. Substitute the $\beta s$ in the formula with the values you got from the regression to get the fitted model.

5. Sanity test: Are the signs (positive/negative) what you expected. If not,  the related variable does not belong in your model

6. Interpret the $\beta (s)$. E.g., if predictor variable increases by one unit outcome increases by $\beta$ units. Note, the intercept coefficient cannot be interpreted in almost all cases!

7. Check if P value is low enough (e.g., smaller than 0.05=5%). This is the probability for the related coefficient to be 0 and thus irrelevant. You want a low probability for that event. If P value is too high the related variable does not belong in your model. Note, no need to interpret the P for the intercept.

## Real World Example: Sex Discrimination {.smaller}
### Data

Use `? wage1` to learn more about the data:

```{webr}
library(wooldridge)
? wage1
```

Note, in wage analysis the wage is often used in logarithmic form and tenure and experience are squared. For simplicity, we use the orginal values instead.

## Real World Example: Sex Discrimination {.smaller}
### Data
```{r}
#| echo: true
#| code-fold: true
library(janitor)
library(tidyverse)
library(wooldridge)
library(ggdag)
DataWage=wage1 |> 
         clean_names("upper_camel") |>  
         select(Wage, SexFem=Female, Tenure, Exper, 
                Educ, NonWhite=Nonwhite, MSA=Smsa, 
                Married, NumDep=Numdep)
head(DataWage)
```



## Example: Sex Discrimination (Impact of one Predictor) {.smaller}
### Univariate with SexFem only

**Can we trust a univariate regression with only $SexFem$ ???**

For example, education, tenure, experience etc. also influence $Wage$?

```{webr}
ModelSexOnly=lm(Wage~SexFem, DataWage)
summary(ModelSexOnly)
```

## Possible but Incomplete DAG
```{r}
#| code-fold: true
set.seed(123)
dagify(Wage~SexFem+Tenure+Educ+Exper+Talent+SglPar+NonWhite+MSA,
       SglPar~SexFem,
       Tenure~SexFem,
       Educ~SexFem,
       outcome = "Wage",
       exposure = "SexFem"
      ) |> 
ggdag_status(node_size = 22) + theme_void()
```

## Data-Engineering

We do not have the variables `Talent` and `SglPar`. 

`Talent` is no problem because it is not correlated with `SexFem`.<br> 
For `SglPar` we can use data-engineer:

```{r}
#| echo: true
DataWage=DataWage |> 
          mutate(SglPar=ifelse((Married==0 & NumDep>0),1,0)) |> 
          mutate(Married=NULL, NumDep=NULL)
head(DataWage)
```


## Which Independent Variables Should we Choose? {.smaller}

**Choosing as many as possible is a bad idea!!!**

## Which Independent Variables Should we Choose?


**If choosing all variables is a bad idea, we need a procedure to choose independent variables (predictors)!**

**Plan and think before you run a regression**



##  Exploratory Data Analysis with a Correlation Matrix


**Correlation Matrix**

```{webr}
library(corrplot)
corrplot(cor(DataWage))
```



## Multivariate Regression




**Experiment with the Wage Analysis:**

```{webr}
ModelSexDisc=lm(Wage~SexFem+Educ+Tenure+SglPar+MSA+Exper+NonWhite, data = DataWage)
summary(ModelSexDisc)
```


























