---
title: "Key Machine Learning Concepts"
subtitle: "Explained with Linear Regression"
format: 
  revealjs:
    code-fold: true
---

```{r}
library(tidymodels)
library(rio)
library(kableExtra)
library(janitor)
DataMockup=import("https://lange-analytics.com/AIBook/Data/DataStudyTimeMockup.rds")
```

## What Will You Learn {.scrollable}

-   Reviewing the basic idea behind linear regression

-   Learning how how to measure predictive quality with Mean Square Error ($MSE$).

-   Understanding the role of parameters in a machine learning model in general and in linear regression in particular

-   Calculating optimal regression parameters using OLS

-   Finding optimal regression parameters by trial and error

-   Distinguish between unfitted and fitted models

-   Using the `tidymodels` package to split observations from a dataset randomly into a training and testing dataset.

-   Understanding how categorical data such as the sex of a person (female/male) can be transformed into numerical dummy variable.

-   Being able to distinguish between dummy encoding and one-hot encoding

-   Using `tidymodels` including model design and data pre-processing (recipes) to analyze housing prices (see Section 6.4).

## Univariate Linear Regression - Data Table and Goal

::: columns
::: {.column width="40%"}
**The Regression:**

$$
\widehat{y}_{i} = \beta_{1}x_{i}+\beta_{2}
$$

**The Goal**

Find values for $\beta_1$ and $\beta_2$ that minimize the prediction errors $(\widehat{y}_{i}-y_i)^2$
:::

::: {.column width="60%"}
**The Data Table**

```{r }
kbl(DataMockup %>% mutate(i=1:5) %>% select(i,everything()), 
    caption="Mockup Training Dataset")%>%
  add_header_above(c(" ", "y", "x"), escape=F) %>% 
  kable_styling(bootstrap_options=c("striped","hover"), full_width = F, position="center")
```
:::
:::

## Univariate Linear Regression - Data Diagram and Goal

::: columns
::: {.column width="40%"}
**The Regression:**

$$
\widehat{y}_{i} = \beta_{1}x_{i}+\beta_{2}
$$

**The Goal**

Find values for $\beta_0$ and $\beta_1$ that minimize the prediction errors $(\widehat{y}_{i}-y_i)^2$
:::

::: {.column width="60%"}
**The Data Diagram**

```{r}
Model123=lm(Grade~StudyTime, data=DataMockup)
PredGrade=predict(Model123, DataMockup)
ggplot(DataMockup, aes(x=StudyTime,y=Grade)) +
  geom_line(aes(y=PredGrade), color="red", size=2.7)+
  geom_point(size=5, color="blue")+
  geom_point(aes(y=PredGrade), color="black", size=2.7)+
  geom_segment(aes(x = StudyTime, y = PredGrade,
                   xend = StudyTime, yend = Grade),size=1.2)+
  scale_x_continuous("Study Time", breaks=seq(1,8))+
  scale_y_continuous(limits=c(65,110), breaks=seq(60,100,5))
```
:::
:::

# How to Measure Prediction Quality

```{=tex}
\begin{eqnarray*}
MSE & = & \frac{1}{N} \sum_{i=1}^{N}(\widehat{y}_{i}-y_{i})^{2} \\
 & \Longleftrightarrow& \\
MSE & =  & \frac{1}{N} \sum_{i=1}^{N}(\underbrace{\overbrace{\beta_{1}x_{i}+\beta_2}^{\mbox{Prediction $i$}}-y_i}_{\mbox{Error $i$}})^2
\end{eqnarray*}
```
**Note, when the data are given (i.e.,** $x_i$ and $y_i$ are given), the $MSE$ depends only on the choice of $\beta_1$ and $\beta_2$ »

## Custom R Function to Calculate MSE

```{r}
#| echo: false
FctMSE=function(VecBeta, data){
                        Beta1=VecBeta[1]
                        Beta2=VecBeta[2]
                        data=data %>%
                        rename(Y=1, X=2) %>% 
                        mutate(YPred=Beta1*X+Beta2) %>%
                        mutate(Error=YPred-Y) %>%
                        mutate(ErrorSqt=Error^2)
                        
                        MSE=mean(data$ErrorSqt)
                        
                        return(MSE)}
```

Function Call:

```{r}
#| echo: true
VecBetaTest=c(4,61)
ResultMSE=FctMSE(VecBetaTest, DataMockup)
print(ResultMSE)
```



Function Definition:»

```{r}
#| echo: true
FctMSE=function(VecBeta, data){
                        Beta1=VecBeta[1]
                        Beta2=VecBeta[2]
                        data=data %>%
                        rename(Y=1, X=2) %>% 
                        mutate(YPred=Beta1*X+Beta2) %>%
                        mutate(Error=YPred-Y) %>%
                        mutate(ErrorSqt=Error^2)
                        
                        MSE=mean(data$ErrorSqt)
                        
                        return(MSE)}
```

## How to Find Optimal Values for $\beta_1$ and $\beta_2$

**Method 1:**

:   **Calculate optimal values** for the parameters (the $\beta s$) based on Ordinary Least Squares (OLS) using two formulas (**Note,** this method works only for linear regression)

**Method 2:**

:   We can use a **systematic trial and error process**.

## Method 1: Calculate Optimal Parameters (only for OLS!) {.smaller .scollable}

```{=tex}
\begin{eqnarray*} 
\beta_{1,opt}&=& \frac
{N \sum_{i=1}^N y_i x_i- \sum_{i=1}^N y_i \sum_{i=1}^N x_i} 
{N \sum_{i=1}^N x_i^2 - \left (\sum_{i=1}^N x_i \right )^2}=3.96\\
&& \nonumber \\
\beta_{2,opt.}&=& \frac{ \sum_{i=1}^N y_i  - \beta_1 \sum_{i=1}^N x_i} {N} = 64.18
\end{eqnarray*}
```
```{r}
DataTable=DataMockup %>%
  mutate(GradeXStudyTime=Grade*StudyTime) %>%
  mutate(StudyTimeSquared=StudyTime^2) 


kbl(DataTable %>% mutate(i=1:5) %>% select(i,everything()), 
    caption="Mockup Training Dataset ")%>%
  add_header_above(c(" ", "y", "x", "y x","$x x"), escape=F) %>% 
  kable_styling(bootstrap_options=c("striped","hover"), full_width = F, position="center")
```

## Method 2: Use a Systematic Trial and Error Process 🤓

-   **Grid Search (aka Brute Force):**

    1.  For a given range of $\beta_1$ and $\beta_2$ values, build a table with pairs of all combinations of these $\beta s$.
    2.  Then use our custom `FctMSE()` command to calculate a $MSE$ for each $\beta$ pair.
    3.  Find the $\beta$ pair with the lowest $MSE$

-   **Optimizer:** Use the R build-in optimizer. Push the start values for $\beta_1$ and $\beta_2$ together with the data to the optimizer as arguments. The rest is done by the optimizer.

-   See the R script in the footnote to see both algorithms in action.»

::: footer
See: [LinRegrScript.R script100](https://lange-analytics.com/AIBook/Scripts/LinRegrScript.R)
:::

## Univariate OLS with a Real World Dataset

**Data**

```{r}
#| echo: true
library(rio)
DataHousing =
  import("https://lange-analytics.com/AIBook/Data/HousingDataSmall.csv")
```

- King County House Sale dataset (Kaggle 2015). House sales prices from May 2014 to May 2015 for King County in Washington State.
- Several predictor variables. For now we use only $Sqft$
- We will only use 500 randomly chosen observations from the total of 21,613 observations.

. . .

**Unfitted Model:**
$$\widehat{Price}=\beta_1 Sqft + \beta_2$$

## Univariate OLS with a Real World Dataset {.scrollable}

**Splitting in Training and Testing Datasets**

```{r}
#| echo: true
library(tidymodels)
Split7030=initial_split(0.7,data=DataHousing, strata = Price, breaks = 5)
DataTrain=training(Split7030)
DataTest=testing(Split7030)
```

**DataTrain**
```{r}
print(as_tibble(DataTrain))
```

**DataTest**
```{r}
print(as_tibble(DataTest))
```

## Univariate OLS with a Real World Dataset

**Run the Analysis**

<https://lange-analytics.com/AIBook/Exercises/handler.html?file=05-LinRegrExerc100.Rmd>


## Multivariate Real World Dataset --- Splitting {.scrollable}

```{r}
#| echo: true
library(tidyverse);library(rio);library(janitor);library(tidymodels)
DataHousing =
  import("https://lange-analytics.com/AIBook/Data/HousingData.csv")%>%
  clean_names("upper_camel") %>%
  select(Price, Sqft=SqftLiving, Grade, Waterfront)

set.seed(777)
Split7030=initial_split(0.7,data=DataHousing, strata = Price, breaks = 5)
DataTrain=training(Split7030)
DataTest=testing(Split7030)
```

```{r}
DataTrain=as_tibble(DataTrain)
DataTest=as_tibble(DataTest)
```



**DataTrain**
```{r}
print(DataTrain)
```

**DataTest**
```{r}
print(DataTest)
```

## Dummy and One-Hot Encoding {.scrollable}


**One-Hot Encoding**
```{r}
OneHotTable=tibble(Waterfront_yes=c(0,0,1,0),Waterfront_no=c(1,1,0,1))
print(OneHotTable)
```
 
*One-hot* encoding is easier to interpret but causes problems in OLS (dummy trap) because one variable is redundant. We can calculate one variable from the other (*perfect multicollinearity)*:

$$Waterfront_{yes}=1-Waterfront_{no}$$

## Dummy and One-Hot Encoding {.scrollable}

**Dummy Coding**

We use one variable less than we have categories. Waterfront has two categories. Therefore, we use one variable (e.g., `Waterfront_yes`):

**One-Hot Encoding**
```{r}
DummyTable=tibble(Waterfront_yes=c(0,0,1,0))
print(DummyTable)
```