---
title: "Regularization"
subtitle: "Ridge and Lasso"
format: 
  revealjs:
    code-fold: true
    scrollable: true
    echo: true
    incremental: false
---


## What Will You Learn/Review {.scrollable .smaller}

-   Reviewing the basic idea behind linear regression

-   Learning how  to measure predictive quality with Mean Square Error ($MSE$).

-    Calculating optimal OLS regression parameters using `tidymodels`

-   Distinguish between unfitted and fitted models

- How to interpret the OLS regression parameters and their significance

- Using metrics to evaluate prediction quality on the testing 

## Loading the Libraries and the Data

```{r}
library(tidymodels)
library(rio)
library(janitor)
set.seed(123)
DataHouses=import("https://ai.lange-analytics.com/data/HousingData.csv") |> 
           clean_names("upper_camel") |> 
           select(Price,Sqft=SqftLiving,Bedrooms,Condition) |> 
           sample_n(144) |> 
           mutate(Price=228*Sqft+32543+rnorm(144,0,50000))
head(DataHouses)

```
## Splitting in Training and Testing Data:

```{r}
#| code-fold: false
set.seed(Seed)
Split7030=initial_split(DataHouses,prop=0.7, strata = Price)
DataTrain=training(Split7030)
DataTest=testing(Split7030)
```
## How much is a House Worth in King County?

. . .

**A house with average properties should be predicted with an average price!**

```{r}
MeanSqft=mean(DataHouses$Sqft)
cat("The mean square footage of a house in King county is:", MeanSqft)

MeanPrice=mean(DataHouses$Price)
cat("The mean price of a house in King county is:", MeanPrice)
```

## Predicting the Price of an Average Sized House as the Average of all House Prices

```{r}
#| echo: false
ggplot(aes(x=Sqft, y=Price), data=DataHouses)+
  geom_point(color="red")+
  geom_hline(yintercept=mean(DataHouses$Price))+
  geom_vline(xintercept=mean(DataHouses$Sqft))+
  scale_y_continuous(limits=c(0,900000),  breaks = seq(0,900000,100000), labels = scales::comma)+
  scale_x_continuous(limits=c(0,5000),  breaks = seq(0,5000,500), labels = scales::comma)
```
# How to Measure Prediction Quality with the Mean Squared Error (MSE) {.smaller}

```{=tex}
\begin{eqnarray*}
MSE & = & \frac{1}{N} \sum_{i=1}^{N}(\widehat{y}_{i}-y_{i})^{2} \\
 & \Longleftrightarrow& \\
MSE & =  & \frac{1}{N} \sum_{i=1}^{N}(\underbrace{\overbrace{\beta_{1}x_{i}+\beta_0}^{\mbox{Prediction $i$}}-y_i}_{\mbox{Error $i$}})^2
\end{eqnarray*}
```
**Note, when the data are given (i.e., $x_i$ and $y_i$ are given), the $MSE$ depends only on the choice of $\beta_1$ and $\beta_0$** »


## Including Sqft as Determinat of Price
### Preparing the Data

Blueprint for the data:
```{r}
#| code-fold: false
RecipeHouses=recipe(Price~Sqft, data=DataTrain)
```

## Choosing the Model Blueprint

Blueprint for the model:
```{r}
#| code-fold: false
ModelDesignOLS=linear_reg() |> 
            set_engine("lm") |> 
            set_mode("regression")
```

## How does the Unfitted Model Looks Like?

$$
\underbrace{\widehat{Price}}_\widehat{y}=\underbrace{\beta_1}_m \underbrace{Sqft}_x + \underbrace{\beta_0}_b
$$

## Using a Workflow to Fit the Model to the Data (finding the optimal $\beta_1$ and $\beta_0$ values

$$
\underbrace{\widehat{Price}}_\widehat{y}=\underbrace{\beta_1}_m \underbrace{Sqft}_x + \underbrace{\beta_0}_b
$$
```{r}
#| code-fold: false
WFModelHouses=workflow() |> 
              add_recipe(RecipeHouses) |> 
              add_model(ModelDesignOLS) |> 
              fit(DataTrain)
```

## Unfitted Model vs Fitted Workflow Model {.smaller}



Unfitted Model:
$$
\underbrace{\widehat{Price}}_\widehat{y}=\underbrace{\beta_1}_m \underbrace{Sqft}_x + \underbrace{\beta_0}_b
$$

```{r}
tidy(WFModelHouses)
```

Fitted Model:
$$
\underbrace{\widehat{Price}}_\widehat{y}=\underbrace{238}_m \cdot\underbrace{Sqft}_x + \underbrace{6584}_b
$$

Predict the price for a house with 1,000 sqft and send it to me in a private chat!

## Interpretation and Significance {.smaller}

```{r}
#| echo: false
tidy(WFModelHouses)
```

$$
\begin{align}
\widehat{Price}&=238 \cdot Sqft +  6584\\
 (+238)&=238\cdot (+1) +  (+0)\\
 (+476)&=238\cdot (+2) +  (+0)\\
 (+714)&=238\cdot (+3) +  (+0)
\end{align}
$$
**For each extra $Sqft$ the predicted price increases by \$238**

**The variable $Sqft$ is significant. I.e., the probability that the related coefficient $\beta_1$ equals zero is extremely small.**

## How Does the fitted model that considers SQFT improves the prediction compared to a simple average

```{r}
#| echo: false
ggplot(aes(x=Sqft, y=Price), data=DataTrain)+
  geom_point(color="red")+
  geom_vline(xintercept=mean(DataHouses$Sqft))+
  geom_hline(yintercept=mean(DataHouses$Price))+
  geom_smooth(method="lm", se=FALSE, color="blue")+
  scale_y_continuous(limits=c(0,900000),  breaks = seq(0,900000,100000), labels = scales::comma)+
  scale_x_continuous(limits=c(0,5000),  breaks = seq(0,5000,500), labels = scales::comma)
```
## Evaluating Predictive Quality with the Testing Dataset

```{r}
#| code-fold: false
DataTestWithPred=augment(WFModelHouses, new_data=DataTest)
metrics(DataTestWithPred, truth=Price, estimate=.pred)
```

## Project: Analysis with all Variables

```{r}
#| code-fold: false
library(rio)
library(janitor)
library(tidyverse)
DataHouses=import("https://ai.lange-analytics.com/data/HousingData.csv") |> 
           clean_names("upper_camel") |> 
           select(Price,Sqft=SqftLiving,Bedrooms,Condition)
```

