---
title: "Quarto Technical Writing and Data Template"
subtitle: "A Guide to References, Math, and Data Manipulation"
author: "Carsten Lange (vibe-coded with Gemini)"
date: last-modified
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    embed-resources: true
---


## Introduction {#sec-intro}

This document serves as a template for your assignments. It demonstrates how to handle technical elements like cross-referencing, mathematical notation, and basic data manipulation using `tidyverse`. 

You can refer back to the introduction using a label such as `{#sec--intro}` directly behind the section title and the reference that label in your text with  `@sec-intro`. Like here: @sec-intro

## Tables and Figures

### Figures
To make a figure referenceable and print a caption, you must provide a label starting with `fig-` at the beginning of the code chunk options. E.g.:

` ```{r}`<br>
`#| label: fig-plot`<br>
`#| fig-cap: "A scatterplot of displacement vs. highway mileage using the mpg dataset."`

```{r}
#| label: fig-plot
#| fig-cap: "A scatterplot of displacement vs. highway mileage using the mpg dataset."
#| message: false
#| code-fold: true

library(tidyverse)
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(aes(color = class)) +
  theme_minimal()
```

As shown in @fig-plot, there is a clear relationship between engine size and fuel efficiency.

### Tables

Similarly, tables require a label starting with tbl-. We use `knitr::kable()` to format the raw data frames into clean tables for the report.

**Code Snippet:**



```{r}
#| label: tbl-data
#| tbl-cap: "The first six rows of the mpg dataset."
set.seed(123)
knitr::kable(slice_sample(mpg, n=6))
```

The raw data values are detailed in @tbl-data.

## Mathematical Notation

### Inline Math

To include math within a sentence, wrap the LaTeX code in single dollar signs $. For example, the sample mean is defined as $\bar{x} = \frac{\sum x_i}{n}$.

Note: Do not leave spaces between the $ and your math code (e.g., use $x$, not $ x $).

### Display Math

For equations on their own line, use double dollar signs \$\$. 

$$f(x) = \frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{1}{2}\left(\frac{x-\mu}{\sigma}\right)^2}$$


To number and reference an equation, add a label starting with `#eq-` in curly braces immediately after the closing \$\$. Start a new line for each \$\$. E.g.:

`$$`<br>
`E = mc^2 `<br>
`$$ {#eq-einstein}`

$$
E = mc^2 
$$ {#eq-einstein}

We can see the **relationship** between *energy* and *mass* in @eq-einstein.

## Data Manipulation with tidyverse

### Grouping and Summarizing

Use `group_by()` to categorize data and `summarize()` to calculate aggregate statistics (like mean, sum, or count) for those categories.

```{r}
#| label: tbl-summary
#| tbl-cap: "Average Highway Mileage and Count by Vehicle Class."
#| message: false

library(tidyverse)

AggregatedMPG=mpg |> 
  group_by(class) |>
  summarize(miles_per_gallon_hwy = mean(hwy),
    observations = n()
  ) 
  knitr::kable(AggregatedMPG)
```

The summary statistics in @tbl-summary show how mileage varies by vehicle type.

### Left Join
A `left_join()` merges two tables. It keeps every row from the "left" (first) table and adds columns from the "right" (second) table where the keys match.

```{r}
#| label: tbl-lefttable
#| tbl-cap: "Student Data (left table)"

# Create example data frames
students <- data.frame(
  student_id = c(101, 102, 103),
  name = c("Alice", "Bob", "Charlie")
)
knitr::kable(students)
```


```{r}
#| label: tbl-righttable
#| tbl-cap: "Grades Data (right table)"

grades <- data.frame(
  student_id = c(101, 102),
  grade = c("A", "B")
)
knitr::kable(grades)
```

```{r}
#| label: tbl-joinedtable
#| tbl-cap: "Students with their grades (joined table)"

# Perform the join
students_with_grades =  left_join(students, grades, by = c("student_id" = "student_id")) 
knitr::kable(students_with_grades)
```

As seen in @tbl-join, "Charlie" remains in the dataset, but his grade is `NA` because there was no matching ID in the grades table.


### Important Reminders for Students:


* **Labels:** Labels for references (e.g., `#fig-`, `#tbl-`) **must** be inside the code chunk options (the lines starting with `#|`).
* **Indentation:** In the YAML header at the top, indentation is strictly enforced. Ensure there are two spaces for each level.