---
title: "Flight Data Analysis: NYC 2013"
author: "Your Name"
date: last-modified
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
---

# Introduction {#sec-intro}

In this lab, we explore the `nycflights13` dataset to understand departure patterns and delays. We will practice data cleaning, joining tables, and visualizing results.

# Methodology {#sec-method}

To analyze the delays, we will calculate the arithmetic mean ($\bar{x}$) of departure delays. The formula for the mean is defined in @eq-mean:

$$
\bar{x} = \frac{1}{n} \sum_{i=1}^{n} x_i
$$ {#eq-mean}

We will also use a mutating join to combine flight data with airport metadata.

# Data Import {#sec-data-prep}

We use the `rio` package to import the CSV files and `tidyverse` for data manipulation.

```{r}
#| label: setup
#| message: false
#| warning: false

library(rio)
library(tidyverse)

# Import local data files
flights  <- import("data/Flights.csv")
airports <- import("data/Airports.csv")
```

# Data Merging and Cleaning {#sec-join}

To see the full names of the destination airports, we perform an `inner_join()`. We match the `dest` column in the flights data to the `faa` code in the airports data.

```{r}
#| label: tbl-joined
#| tbl-cap: "First 6 rows of the joined flight and airport data."

flights_joined <- flights %>%
  inner_join(airports, by = c("dest" = "faa"))

# Display a preview
head(flights_joined) %>% 
  select(year, month, day, carrier, dest, name)
```

As shown in @tbl-joined, we now have the full name of each airport associated with every flight record.

# Analysis and Visualization {#sec-viz}

## Descriptive Statistics

We calculate the average departure delay for the three NYC origin airports: JFK, LGA, and EWR.

```{r}
#| label: tbl-stats
#| tbl-cap: "Mean and Median Departure Delays by Origin Airport"

flights_joined %>%
  group_by(origin) %>%
  summarize(
    mean_delay = mean(dep_delay, na.rm = TRUE),
    median_delay = median(dep_delay, na.rm = TRUE)
  )
```

## Visualizing Delays by Destination

The following figure illustrates the average delay for the destinations with the most traffic.

```{r}
#| label: fig-delays
#| fig-cap: "Average Departure Delay for Top 10 Destinations"

flights_joined %>%
  group_by(name) %>%
  summarize(
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    n = n()
  ) %>%
  filter(n > 1000) %>%
  slice_max(avg_dep_delay, n = 10) %>%
  ggplot(aes(x = reorder(name, avg_dep_delay), y = avg_dep_delay)) +
  geom_col(fill = "midnightblue") +
  coord_flip() +
  labs(
    x = "Airport Name",
    y = "Average Delay (Minutes)"
  ) +
  theme_minimal()
```

# Conclusion

Based on the analysis in @sec-viz, specifically the data visualized in @fig-delays, we can identify which destination airports experience the highest average departure delays from NYC.
```

---

### Key Features Included:
* **YAML Header:** Configured for a numbered TOC.
* **Math:** Includes both inline ($\bar{x}$) and a labeled display equation (@eq-mean).
* **Cross-References:** Uses `@sec-`, `@tbl-`, and `@fig-` syntax throughout the narrative.
* **Tidyverse/Rio:** Uses `import()` and `inner_join()` as requested.