---
title: "Lab: NYC Flight Analysis with R & Quarto (Worksheet)"
author: "all names of your group"
date: last-modified
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    embed-resources: true
---

# Overview {#sec-overview}

In this lab, you will explore the `nycflights13` dataset using the **Positron IDE**. Your goal is to move beyond simple coding and produce a **reproducible research report**. You will practice data loading with `rio`, relational joins with the `tidyverse`, and professional formula typesetting with LaTeX.

# Technical Setup {#sec-setup}

Before you begin, ensure you have the following installed in your environment:

* **Required Packages:** `install.packages(c("rio", "tidyverse", "janitor"))`
* **Working Directory:** Create a folder in *Positron* 
* **Data Files:** Download and place `Flights.csv` and `Airports.csv` in the same folder as this `.qmd` file (in your working directory).

# Quarto Document Requirements {#sec-reqs}

Your submission must utilize Quarto's advanced formatting features:
1.  **YAML Header:** Configured for a Table of Contents and numbered sections.
2.  **Cross-References:** You must reference at least one Section, one Table, one Equation and one Figure using the `@` syntax.
3.  **Code Chunks:** Use `#| label:` and `#| tbl-cap:` or `#| fig-cap:` for all outputs.

# Methodology & Math 

In your analysis, you must define your variables. Use **inline math** for variables like the sample mean $\bar{x}$ or the standard deviation $\sigma$. 

We will calculate the average delay using the arithmetic mean formula:

here goes the formula

# Task 1: Data Import and Data Engineering {#sec-import}
## Task 1a: Data Import

We use the `rio` package to load our data. This package simplifies the process by automatically handling the CSV format. The `knitr::kable()` command from the `knitr` package formats a data frame as a table.

```{r}
#| message: false
#| warning: false


library(rio)
library(tidyverse)
library(janitor)

# INSTRUCTIONS: Load your CSV files here
Flights  = import("Flights.csv") |> 
           clean_names("upper_camel")
Airports = import("Airports.csv") |> 
           clean_names("upper_camel")

knitr::kable(head(Flights))
```

```{r}
#| label: tbl-OrgAirports
#| tbl-cap: "Original Airports Data Frame"
knitr::kable(head(Airports))
```

## Task 1b:  Data Engineering

For both data frames reduce the variables only to the one needed with `select()`! 

For `Flights` you need four variables including `Origin`, `SchedDepTime`, `DepTime`, and `DepDelay`. For the flights data frame you need the airport code `Faa` and the `Name` of the airport.

```{r}
Flights=
knitr::kable(head(Flights))
```

```{r}
Airports=Airports
```

**Explain here** how the variable `DepDelay` was calculated and comment on the variable types.


# Task 2: Relational Joins 

The `flights` dataset contains airport codes (e.g., "EWR"). To display full names (e.g., "Newark Liberty Intl"), you must perform an `inner_join()` with the `airports` table. Then rename the variable `Name` to `OriginName`

**Note:** You must match the `origin` column from flights to the `faa` column in airports.

```{r}

FlightsJoined=

knitr::kable(head(FlightsJoined))
```

# Task 3: Descriptive Statistics

Using the joined data, calculate the *mean* and *median* `dep_delay` for each origin airport (EWR, JFK, LGA). Refer to @eq-mean for the calculation logic.

```{r}

# INSTRUCTIONS: Use group_by() and summarize()

```

If you get `NA` values and/or negative values for `MedianDelay` check your data!

# Task 4: Visualization

Create a bar chart showing the average departure delay for the top three New York airports. 

```{r}

```

# Troubleshooting & FAQ {#sec-faq}

* **Reference shows as ??**: Check that your labels start with the correct prefix (e.g., `fig-` for figures, `tbl-` for tables).
* **Mean is NA**: Data often contains missing values. Ensure you use `na.rm = TRUE` inside your `mean()` or `median()` functions.
* **Join results in 0 rows**: Ensure you mapped the keys correctly: `by = c("dest" = "faa")`.
* **YAML Error**: Ensure there are no tabs in the YAML header; use spaces only.

# Pre-Submission Checklist {#sec-checklist}

* [ ] Does the document render to HTML without errors?
* [ ] Is the Table of Contents visible and correctly numbered?
* [ ] Does the text include at least one reference to @eq-mean?
* [ ] Are all figures and tables properly captioned and referenced (e.g., "As seen in @fig-delays...")?
* [ ] Have you removed any "empty" code chunks or experimental code?

# Conclusion {#sec-conclusion}

*Write a brief summary of your findings here, referencing the table in @sec-stats and the figure in @sec-viz.*
```
