---
title: "Lab: NYC Flight Analysis with R & Quarto"
author: "Carsten Lange (based on a Gemini draft)"
date: last-modified
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    embed-resources: true
---

# Overview {#sec-overview}

In this lab, you will explore the `nycflights13` dataset using the **Positron IDE**. Your goal is to move beyond simple coding and produce a **reproducible research report**. You will practice data loading with `rio`, relational joins with the `tidyverse`, and professional typesetting with LaTeX.

# Technical Setup {#sec-setup}

Before you begin, ensure you have the following installed in your environment:

* **Required Packages:** `install.packages(c("rio", "tidyverse"))`
* **Data Files:** Place `flights.csv` and `airports.csv` in the same folder as this `.qmd` file.

# Quarto Document Requirements {#sec-reqs}

Your submission must utilize Quarto's advanced formatting features:
1.  **YAML Header:** Configured for a Table of Contents and numbered sections.
2.  **Cross-References:** You must reference at least one Section, Table, and Figure using the `@` syntax.
3.  **Code Chunks:** Use `#| label:` and `#| tbl-cap:` or `#| fig-cap:` for all outputs.

# Methodology & Math {#sec-math}

In your analysis, you must define your variables. Use **inline math** for variables like the sample mean $\bar{x}$ or the standard deviation $\sigma$. 

We will calculate the average delay using the arithmetic mean formula defined in @eq-mean:

$$
\bar{x} = \frac{1}{n} \sum_{i=1}^{n} x_i
$$ {#eq-mean}

# Task 1: Data Import {#sec-import}

We use the `rio` package to load our data. This package simplifies the process by automatically handling the CSV format.

```{r}
#| label: setup
#| message: false
#| warning: false

library(rio)
library(tidyverse)

# INSTRUCTIONS: Load your CSV files here
# flights  <- import("flights.csv")
# airports <- import("airports.csv")
```

# Task 2: Relational Joins {#sec-join}

The `flights` dataset contains airport codes (e.g., "EWR"). To display full names (e.g., "Newark Liberty Intl"), you must perform an `inner_join()` with the `airports` table. 

**Note:** You must match the `dest` column from flights to the `faa` column in airports.

```{r}
#| label: tbl-joined
#| tbl-cap: "Sample of Flights Joined with Airport Metadata"

# INSTRUCTIONS: Perform the inner_join here
# flights_joined <- flights %>% 
#   inner_join(airports, by = c("dest" = "faa"))

# head(flights_joined)
```

# Task 3: Descriptive Statistics {#sec-stats}

Using the joined data, calculate the mean and median `dep_delay` for each origin airport (EWR, JFK, LGA). Refer to @eq-mean for the calculation logic.

```{r}
#| label: tbl-delays
#| tbl-cap: "Summary Statistics for Departure Delays"

# INSTRUCTIONS: Use group_by() and summarize()
# Remember to use na.rm = TRUE
```

# Task 4: Visualization {#sec-viz}

Create a bar chart showing the average departure delay for the top 10 most frequent destinations. 

```{r}
#| label: fig-delays
#| fig-cap: "Average Departure Delay for Top 10 Destinations"

# INSTRUCTIONS: Create a ggplot showing average delay by airport name
```

# Troubleshooting & FAQ {#sec-faq}

* **Reference shows as ??**: Check that your labels start with the correct prefix (e.g., `fig-` for figures, `tbl-` for tables).
* **Mean is NA**: Data often contains missing values. Ensure you use `na.rm = TRUE` inside your `mean()` or `median()` functions.
* **Join results in 0 rows**: Ensure you mapped the keys correctly: `by = c("dest" = "faa")`.
* **YAML Error**: Ensure there are no tabs in the YAML header; use spaces only.

# Pre-Submission Checklist {#sec-checklist}

* [ ] Does the document render to HTML without errors?
* [ ] Is the Table of Contents visible and correctly numbered?
* [ ] Does the text include at least one reference to @eq-mean?
* [ ] Are all figures and tables properly captioned and referenced (e.g., "As seen in @fig-delays...")?
* [ ] Have you removed any "empty" code chunks or experimental code?

# Conclusion {#sec-conclusion}

*Write a brief summary of your findings here, referencing the table in @sec-stats and the figure in @sec-viz.*
```
