---
title: "Interactive Section"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: Exercises

---

<style>
  div.yellowbox {
    background-color: #f1dfa6;
    border: 1px solid black;
    padding: 21px;
  }

  strong {
    color: #d10056
  }

  .center {
    text-align: center;
  }
</style>


```{r setup, include=FALSE}
library(learnr)
library(rio)
library(tidyverse)
library(tidymodels)
library(janitor)
library(kableExtra)

knitr::opts_chunk$set(exercise=TRUE, exercise.eval=FALSE, exercise.lines=6, attr.source = ".numberLines")
```

:::: {.yellowbox data-latex=""}

::: {.center data-latex=""}
**How to Work with this Interactive Section**
:::

Here you will find the complete interactive section from the book, including book content, and *R* code blocks. Read the content in a browser as you would read it in the textbook. 

Within the *R* code blocks, you can work in various ways:

* Use the `Run Code` button to execute the *R* code and see the result.

* Change the *R* code, click the `Run Code` button, and then see the new results.

* To get the original *R* code back, click the `Start Over` button at the left-top corner of each *R* code block. 

* To reset the *R* code for all code blocks, click the `Start Over` in the left menu.

**Make sure you installed all needed packages!** See Section "R Packages Required"  in this Chapter for a list of packages that need to be installed. Section 3.2 explain how to install a packege in *RStudio*.



**For Troubleshooting:** [Click here](https://blog.lange-analytics.com/2024/01/interactsessions.html)
::::


 
 
 <clex id="Exc100">

```{r TrainTest-Exerc100-LoadKWineData, eval=TRUE, message=FALSE, warning=FALSE, exercise=FALSE, include=FALSE}
library(tidymodels); library(rio); library(janitor)

DataWine=readRDS(gzcon(url("https://ai.lange-analytics.com/data/WineData.rds"))) |> 
         clean_names("upper_camel") |> 
         rename(Sulfur=TotalSulfurDioxide) |> 
         mutate(WineColor=as.factor(WineColor))

set.seed(876)
Split7030=initial_split(DataWine,prop=0.7, strata=WineColor)
DataTrain=training(Split7030)
DataTest=testing(Split7030)

ModelDesign=nearest_neighbor(neighbors=tune()) |>
               set_engine("kknn") |> 
               set_mode("classification")

Recipe=recipe(WineColor~., data=DataTrain) |>
       step_rm(Quality) |>           
       step_normalize(all_predictors())

ParGrid=tibble(neighbors=c(1:15))


set.seed(123)
FoldsForTuning=vfold_cv(DataTrain, v=5, repeats=1, strata=WineColor)

TuneWFModel=workflow() |> 
            add_recipe(Recipe) |>
            add_model(ModelDesign) 

if(exists("RerunTuningForUniVarkNeighWine")){
  if (RerunTuningForUniVarkNeighWine){
    TuneResults=TuneWFModel |> tune_grid(FoldsForTuning, 
          grid=ParGrid,
          metrics=metric_set(accuracy, specificity, sensitivity))
    saveRDS(TuneResults,"data/TuneResultsWine.rds")
  }
}
TuneResults=readRDS(gzcon(url("https://ai.lange-analytics.com/data/TuneResultsWine.rds")))

BestHyperPar=select_best(TuneResults, metric="accuracy")

WFModelBest=TuneWFModel |> 
                            finalize_workflow(BestHyperPar) |> 
                            fit(DataTrain)


DataTestWithPredBestModel=augment(WFModelBest,DataTest)
```

```{r SetupForCodeChunkOutsideExerc, include=FALSE,  exercise=FALSE}
library(tidymodels); library(rio); library(janitor)

DataWine=readRDS(gzcon(url("https://ai.lange-analytics.com/data/WineData.rds"))) |> 
         clean_names("upper_camel") |> 
         rename(Sulfur=TotalSulfurDioxide) |> 
         mutate(WineColor=as.factor(WineColor))

set.seed(876)
Split7030=initial_split(DataWine,prop=0.7,strata=WineColor)
DataTrain=training(Split7030)
DataTest=testing(Split7030)
Recipe=recipe(WineColor~., data=DataTrain) |>
       step_rm(Quality) |>           
       step_normalize(all_predictors())
ModelDesign=nearest_neighbor(neighbors=tune()) |>
            set_engine("kknn") |> 
            set_mode("classification")
TuneResults=readRDS(gzcon(url("https://ai.lange-analytics.com/data/TuneResultsWine.rds")))
BestHyperPar=select_best(TuneResults, metric="accuracy")
```

**Step 1 - Generating Training and Testing Data:**

:   As before, you use the wine dataset and split the data into training
    (`DataTrain`) and testing data (`DataTest`). Since you use the same
    value in the `set.seed()` command, the (random) split will be
    identical to the one we used before with the $k=4$ *Nearest
    Neighbor* model. The code below has been executed already.

    ```{r TrainTest-Exerc100-CreateTrainTestData, exercise=FALSE}
    library(tidymodels); library(rio); library(janitor)
    DataWine=import("https://ai.lange-analytics.com/data/WineData.rds") |> 
             clean_names("upper_camel") |> 
             rename(Sulfur=TotalSulfurDioxide) |> 
             mutate(WineColor=as.factor(WineColor)) 

    set.seed(876)
    Split7030=initial_split(DataWine, prop=0.7, strata=WineColor)
    DataTrain=training(Split7030)
    DataTest=testing(Split7030)

    head(DataTrain)
    ```

**Step 2 - Create a Recipe:**

:   Here, you will create a *recipe* and store it in the *R* object
    `Recipe`. Use `step_rm()` to remove the predictor variable $Quality$
    because it is not related to $WineColor$, and the command
    `step_normalize()` to normalize all remaining predictors. Please,
    substitute `<THESE>` placeholders accordingly and execute the code.
    Note that the data frame `DataTrain` has already been loaded in the
    background:

    ```{r TrainTest-Exerc100-Recipe, exercise.setup="TrainTest-Exerc100-LoadKWineData", exercise=TRUE, eval=FALSE, exercise.lines=5}
    Recipe=recipe(WineColor~., data=DataTrain) |>
           step_<NAME OF STEP>(<ARGUMENT OF STEP>) |>           
           step_<NAME OF STEP>(all_predictors())
    print(Recipe)
    ```

    ```{r TrainTest-Exerc100-Recipe-hint, exercise=FALSE,  eval=FALSE, include=FALSE, exercise.lines=17}
    'Substitute <NAME OF STEP> with rm and normalize. 
    Also, substitute <ARGUMENT OF STEP> with `Quality`.

    The next HINT provides the solution. So, keep trying before you peek.'

    ```

    ```{r TrainTest-Exerc100-Recipe-solution, exercise=FALSE,  eval=FALSE, include=FALSE, exercise.lines=17}
    Recipe=recipe(WineColor~., data=DataTrain) |>
           step_rm(Quality) |>           
           step_normalize(all_predictors())
    print(Recipe)
    ```

**Step 3 - Create a Model Design:**

:   Next, you will create the *model design* and store it into the *R*
    object `ModelDesign`. Since you plan to tune the argument
    `neighbors=` (stands for the $k$ in *Nearest Neighbors*), you have
    to add it as an argument into the `nearest_neighbor()` command by
    substituting `<ARGUMENT(S) OF COMMAND>` with the argument and its
    value. Remember that you cannot set `neighbors` to a specific
    numerical value because you want to tune the *hyper-parameter*
    `neighbors` later on. Therefore you have to assign the placeholder
    `tune()` to the argument `neighbors`.

    ```{r TrainTest-Exerc100-ModelDesign, exercise=TRUE, eval=FALSE, exercise.lines=5}
    ModelDesign=nearest_neighbor(<ARGUMENT(S) OF COMMAND>) |>
                set_engine("kknn") |>
                set_mode("classification")
    print(ModelDesign)
    ```

    ```{r TrainTest-Exerc100-ModelDesign-hint, exercise=FALSE,  eval=FALSE, include=FALSE, exercise.lines=7}
    'Instead of assigning a number to the hyper-parameter neighbors 
    use the command tune() as a place holder.

    The next HINT provides the solution. So, keep trying before you peek.'
    ```

    ```{r TrainTest-Exerc100-ModelDesign-solution, exercise=FALSE,    eval=FALSE, include=FALSE}
    ModelDesign=nearest_neighbor(neighbors=tune()) |>
               set_engine("kknn") |> 
               set_mode("classification")

    print(ModelDesign)
    ```

**Step 4 - Add the Recipe and the Model Design to a Workflow:**

:   The code block below adds the *R* object `Recipe` and the model
    design object `ModelDesign` to a *workflow* model named
    `TuneWFModel`. The *R* code has been executed already.

    ```{r TrainTest-Exerc100-CreateWorkflow, exercise=FALSE}
    TuneWFModel=workflow() |> 
                add_recipe(Recipe) |>
                add_model(ModelDesign) 
    print(TuneWFModel)
    ```

    You can see in the printout above that the *workflow* is not
    *finalized* because the number of `neighbors` has not been set in
    the *model design*. The *hyper-parameter* `neighbors` is set to
    `tune()` instead and will later, in Step 7, be replaced in a trial
    and error process with different values for `neighbors`.

**Step 5 - Create a Hyper-Parameter Grid:**

:   Later, when tuning is executed in Step 7, values reaching from 1 --
    15 for the *hyper-parameter* `neighbors` shall be tried out.

    You need to provide these values in a data frame column that is
    named the same as the *hyper-parameter*. Below, replace the
    `<LIST OF VALUES>` to define a column `neighbors` in the data frame
    `ParGrid`. The column should contain values from 1 -- 15:

    ```{r TrainTest-Exerc100-Grid, eval=FALSE, exercise=TRUE, exercise.lines=3}
    ParGrid=data.frame(neighbors=c(<LIST OF VALUES>))
    print(ParGrid)
    ```

    ```{r TrainTest-Exerc100-Grid-hint, eval=FALSE, exercise=FALSE,  eval=FALSE, include=FALSE, exercise.lines=7}
    'Use the c() command to create a vector containing the numbers 1-15
    either separated by a comma, or you can use a colon (:) to indicate from/to.

    The next HINT provides the solution. So, keep trying before you peek.'
    ```

    ```{r TrainTest-Exerc100-Grid-solution, eval=FALSE, exercise=FALSE, include=FALSE, exercise.lines=5}
    ParGrid=data.frame(neighbors=c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15))
    print(ParGrid)
    ```

**Step 6 - Creating Resamples for Cross-Validation:**

:   The values you have created above for $k$ (*hyper-parameter*
    `neighbors`) will be evaluated later using five folds (*resamples*).
    Each fold contains the complete training data, but different
    sections are used for training and assessment in each fold.

    Please create five folds below by substituting `<NUMBER OF FOLDS>`
    accordingly.

    The folds will be saved in the *R* object `FoldsForTuning`:

    ```{r TrainTest-Exerc100-Folds, exercise.setup="TrainTest-Exerc100-LoadKWineData", exercise=TRUE, exercise.lines=5, eval=FALSE}
    set.seed(123)
    FoldsForTuning=vfold_cv(DataTrain, v=<NUMBER OF FOLDS>, 
                            strata=WineColor)
    print(FoldsForTuning)
    ```

    ```{r TrainTest-Exerc100-Folds-hint, exercise=FALSE,  eval=FALSE, include=FALSE, exercise.lines=17}
    'The argument v= determines the number of folds to be used for tuning.

    The next HINT provides the solution. So, keep trying before you peek.'
    ```

    ```{r TrainTest-Exerc100-Folds-solution, exercise=FALSE,  eval=FALSE, include=FALSE, exercise.lines=17}
    set.seed(123)
    FoldsForTuning=vfold_cv(DataTrain, v=5, strata=WineColor)

    print(FoldsForTuning)
    ```

**Step 7 - Tune the Workflow and Train All Models:**

:   Now it is time to run the tuning procedure using the `tune_grid()`
    command. Be patient because it will take some time to fully execute.
    Since we have to try out 15 parameters and use five folds for each
    model, the `tune_grid()` command has to fit 75 models
    $(15\cdot 5=75)$.

    Please substitute `<LIST OF METRICS>` with a list of metrics to be
    calculated. Use the metrics *accuracy*, *specificity*, and
    *sensitivity*.

    After the tuning is finished, all results are stored in the *R*
    object `TuneResults`, and they can be evaluated by different
    *metrics* commnands.

    For example the command `autoplot()` provides a diagrammatic
    overview of the results for all three *metrics*.

    ```{r TrainTest-Exerc100-TuneWorkkFlow, exercise.setup="TrainTest-Exerc100-LoadKWineData", exercise=TRUE, exercise.lines=4, eval=FALSE}
    TuneResults=tune_grid(TuneWFModel, resamples=FoldsForTuning,
    grid=ParGrid, metrics=metric_set(<LIST OF METRICS>))
    autoplot(TuneResults)
    ```

    ```{r TrainTest-Exerc100-TuneWorkkFlow-hint, exercise=FALSE,  eval=FALSE, include=FALSE}
    'The command metric_set takes a comma-separated list 
    of metrics to be evaluated.

    Consider that you want to evaluate the metrics accuracy, specificity, 
    sensitivity.

    The next HINT provides the solution. So, keep trying before you peek.'
    ```

    ```{r TrainTest-Exerc100-TuneWorkkFlow-solution, exercise=FALSE, include=FALSE, eval=FALSE}
    TuneResults=TuneWFModel |> tune_grid(FoldsForTuning, grid=ParGrid ,  
                       metrics=metric_set(accuracy, specificity, sensitivity))
    autoplot(TuneResults)
    ```

    The three graphs that you will create in the exercise above are also
    displayed in Figure
    \@ref(fig:TrainTest-Exerc100-ThreeMetricGraphsBook). They show for
    each *metric* the related *accuracy*, *specificity*, and
    *sensitivity* for all tried out *hyper-parameters* (`neighbors`).
    The results for the five folds are averaged.

    ```{r TrainTest-Exerc100-ThreeMetricGraphsBook, exercise=FALSE, fig.cap="Tuning Results for Various k Values", fig.alt="This figure shows three graphs representing accuracy, sensitivity, and specificity related to the tuning parameter k.", echo=FALSE, message=FALSE, warning=FALSE,}
    library(rio); library(tidymodels)
    TuneResults=readRDS(gzcon(url("https://ai.lange-analytics.com/data/TuneResultsWine.rds")))
    autoplot(TuneResults)+ scale_x_continuous(breaks=seq(1,15,1))
    ```

    You can see that `neighbors` values between 1 and 4 produce the best
    results for *accuracy* (predicting red and white wines (positive and
    negative class) correctly) and *sensitivity* (predicting red wines
    (positive class) correctly). If you look at *specificity*
    (predicting white wines (negative class) correctly), you can see
    that $k=5$ produces the best result. However, $k=5$ also leads to a
    sharp decrease in *sensitivity* and a decrease in *accuracy* as
    well.

    It seems reasonable to use the best result for *accuracy*, which
    means choosing a $k$ between 1 and 4.

**Step 8 - Extract the Best Hyper-Parameter(s):**

:   All assessment results for the specified metrics are stored in the
    *tuning* object `TuneResults`. Choose the metric *accuracy* by
    substituting `<METRIC>` accordingly. Afterward, when you execute the
    code the `select_best()` command extracts the best *hyper-parameter*
    (value for `neighbors`) for the metric you specified:

    ```{r TrainTest-Exerc100-BestHypPar, exercise.setup="TrainTest-Exerc100-LoadKWineData", exercise=TRUE, exercise.lines=3, eval=FALSE}
    BestHyperPar=select_best(TuneResults, metric="<METRIC>") 
    print(BestHyperPar)
    ```

    ```{r TrainTest-Exerc100-BestHypPar-hint, exercise=FALSE,  eval=FALSE, include=FALSE, exercise.lines=17}
    'Substitute <METRIC> with accuracy and make sure to not delete the quotes.

    The next HINT provides the solution. So, keep trying before you peek.'
    ```

    ```{r TrainTest-Exerc100-BestHypPar-solution, exercise=FALSE,  eval=FALSE, include=FALSE, exercise.lines=3}
    BestHyperPar=select_best(TuneResults, metric="accuracy") 
    print(BestHyperPar)
    ```

    As you will see in the printout after executing the *R* code, the
    best value for the *hyper-parameter* `neighbors` based on *accuracy*
    is `r select_best(TuneResults, metric="accuracy")[[1,1]]`. It is
    saved in the data frame `BestHyperPar`.

    You can try other metrics in the code block above and see if the
    result changes. Why would the metric `rmse` cause an error?

**Step 9 - Finalize and Train the Best Workflow Model:**

:   The code block below is executed already. The `finalize_workflow()`
    command used the value from `BestHyperPar` to substitute the
    `tune()` placeholder in the *R* object `TuneWFModel` (created in
    Steps 2 -- 4). The *hyper-parameter* is now set to
    ``` neighbors=``r select_best(TuneResults, metric="accuracy")[[1,1]] ```
    completing the *workflow*. The command `fit(DataTrain)` calibrates
    the *workflow* to the training data and the result is saved into
    `WFModelBest`.

    ```{r TrainTest-Exerc100-FinalizeSilent, exercise=FALSE, exercise.lines=6, eval=FALSE}
    WFModelBest=TuneWFModel |> 
                finalize_workflow(BestHyperPar) |>  
                fit(DataTrain)

    print(WFModelBest)
    ```

    ```{r TrainTest-Exerc100-FinalizeOut, echo=FALSE, exercise=FALSE, exercise.lines=6}
        WFModelBest=TuneWFModel  |> 
                    finalize_workflow(BestHyperPar)  |>  
                    fit(DataTrain)

        PrintOut=capture.output(WFModelBest)
        PrintOut=strtrim(PrintOut,70)
        cat(PrintOut,sep = "\n")
    ```

    At the end of the printout above, you can see that $(k)$ was set to
    `r select_best(TuneResults, metric="accuracy")[[1,1]]` rather than
    `tune()`. You can also see that `WFModelBest` is fitted to the
    training data because the "Minimum missclassification" is reported
    (based on the training data).

**Step 10: Assess Prediction Quality Based on the Testing Data:**

:   Since `WFModelBest` is a fitted model, you can use it for
    predictions. In this last step, you use the `augment()` command to
    predict `WineColor`. The `augment()` command will then add the
    prediction results as column `.pred` to the testing data.

    The `conf_mat()` command compares the predictions in column `.pred`
    to the true values to create a *confusion matrix*. But before, you
    have to substitute `<OUTCOME VARIABLE>` with the variable name
    (column name) for the outcome variable.

    ```{r TrainTest-Exerc100-FinalConfMat, exercise.setup="TrainTest-Exerc100-LoadKWineData", exercise=TRUE, exercise.lines=3, eval=FALSE}
    DataTestWithPredBestModel=augment(WFModelBest, DataTest)
    conf_mat(DataTestWithPredBestModel, truth=<OUTCOME VARIABLE>, 
             estimate=.pred_class)
    ```

    ```{r TrainTest-Exerc100-FinalConfMat-hint, exercise=FALSE,  eval=FALSE, include=FALSE, exercise.lines=17}
    'The (column) name for the outcome variable is WineColor.

    The next HINT provides the solution. So, keep trying before you peek.'
    ```

    ```{r TrainTest-Exerc100-FinalConfMat-solution, exercise=FALSE,  eval=FALSE, include=FALSE, exercise.lines=17}
    DataTestWithPredBestModel=augment(WFModelBest,DataTest)
    conf_mat(DataTestWithPredBestModel, truth=WineColor, estimate=.pred_class)
    ```

    You will see in the *confusion matrix* that based on the testing
    data from 480 red wines, only six were misclassified. Likewise, from
    the 480 white wines, only seven were misclassified. Given that the
    classification was only based on the chemical properties of the
    wines, the results are excellent.

    Most likely, a true wine expert might have reached a similar
    impressive result, but it would have taken them a long time to
    classify 960 wines.

</clex>