---
title: "Interactive Project: Tune k-Nearest Neighbors to Detect Wine Color"
subtitle: Companion Exercises for [Practical Machine Learning with R](https://ai.lange-analytics.com)"
author:
  name: "Carsten Lange"
  email: "clange@cpp.edu"
  affiliation: "Cal Poly, Pomona"
format: live-html
engine: knitr
webr:
  repos:
    - 'https://tidymodels.r-universe.dev'
    - 'https://repo.r-wasm.org'
  packages:
    - rsample
    - tune
    - dplyr
    - janitor
    - recipes
    - yardstick
    - parsnip
    - workflows
    - kknn
---
{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ../_extensions/r-wasm/live/_gradethis.qmd >}}


::: {.content-visible when-runtime="loading"}
::: {.callout-note}

## Please be Patient: Loading R Environment
The interactive R environment and the required packages must be loaded in your browser. This typically takes 1–2 minutes. You can monitor the progress in the bottom-right corner. 

**After packages are loaded, it still takes some time** because the website must run a pre-tuning procedure. Please be patient!!!

When everything is loaded, you get a **"Ready to roll"** message below.
Thank you for your patience!
:::
:::



```{webr}
#| auto-run: true
#| echo: false
#| label: exc_all
DataWine=readRDS(url("https://ai.lange-analytics.com/data/WineData.rds")) |> 
         clean_names("upper_camel") |> 
         rename(Sulfur=TotalSulfurDioxide) |> 
         mutate(WineColor=as.factor(WineColor))
set.seed(876)
Split7030=initial_split(DataWine,prop=0.7, strata=WineColor)
DataTrain=training(Split7030)
DataTest=testing(Split7030)

ModelDesignKNN=nearest_neighbor(neighbors=tune()) |>
               set_engine("kknn") |> 
               set_mode("classification")

RecipeWine=recipe(WineColor~., data=DataTrain) |>
       step_rm(Quality) |>           
       step_normalize(all_predictors())

ParGrid=tibble(neighbors=c(1:15))

set.seed(123)

FoldsForTuning=vfold_cv(DataTrain, v=5, repeats=1, strata=WineColor)

TuneWFModelKNN=workflow() |> 
            add_recipe(RecipeWine) |>
            add_model(ModelDesignKNN) 

TuneResults=TuneWFModelKNN |> tune_grid(FoldsForTuning, 
          grid=ParGrid,
          metrics=metric_set(accuracy, specificity, sensitivity))


BestHyperPar=select_best(TuneResults, metric="accuracy")

WFModelBest=TuneWFModelKNN |> 
finalize_workflow(BestHyperPar) |> 
fit(DataTrain)

DataTestWithPredBestModel=augment(WFModelBest,DataTest)
cat("\033[1;32mReady to roll\033[0m\n")
```

```{r}
#| echo: false
#| warning: false
#| message: false
# needed for code almost at end
library(tidymodels); library(janitor)
DataWine=readRDS(url("https://ai.lange-analytics.com/data/WineData.rds")) |> 
         clean_names("upper_camel") |> 
         rename(Sulfur=TotalSulfurDioxide) |> 
         mutate(WineColor=as.factor(WineColor))
set.seed(876)
Split7030=initial_split(DataWine,prop=0.7, strata=WineColor)
DataTrain=training(Split7030)
DataTest=testing(Split7030)

ModelDesignKNN=nearest_neighbor(neighbors=tune()) |>
               set_engine("kknn") |> 
               set_mode("classification")

RecipeWine=recipe(WineColor~., data=DataTrain) |>
       step_rm(Quality) |>           
       step_normalize(all_predictors())

ParGrid=tibble(neighbors=c(1:15))

set.seed(123)

FoldsForTuning=vfold_cv(DataTrain, v=5, repeats=1, strata=WineColor)

TuneWFModelKNN=workflow() |> 
            add_recipe(RecipeWine) |>
            add_model(ModelDesignKNN) 

TuneResults=TuneWFModelKNN |> tune_grid(FoldsForTuning, 
          grid=ParGrid,
          metrics=metric_set(accuracy, specificity, sensitivity))


BestHyperPar=select_best(TuneResults, metric="accuracy")

WFModelBest=TuneWFModelKNN |> 
finalize_workflow(BestHyperPar) |> 
fit(DataTrain)

DataTestWithPredBestModel=augment(WFModelBest,DataTest)
ConFusMat=conf_mat(DataTestWithPredBestModel, truth = WineColor, estimate=.pred_class)

```

**Libraries** already loaded in the background:

- *tidyverse*
- *tidymodels*
- *kknn*



**Step 1 - Generating Training and Testing Data:**

As before, you use the wine dataset and split the data into training
(`DataTrain`) and testing data (`DataTest`). Since you use the same
value in the `set.seed()` command, the (random) split will be
identical to the one we used before with the $k=4$ *Nearest
Neighbor* model. The code below has been executed already.

```{r}
#| echo: true
#| message: false
#| warning: false
library(tidymodels); library(rio); library(janitor); library(kableExtra)
DataWine=import("https://ai.lange-analytics.com/data/WineData.rds") |> 
         clean_names("upper_camel") |> 
         rename(Sulfur=TotalSulfurDioxide) |> 
         mutate(WineColor=as.factor(WineColor)) 

set.seed(876)
Split7030=initial_split(DataWine, prop=0.7, strata=WineColor)
DataTrain=training(Split7030)
DataTest=testing(Split7030)

kable(head(DataWine))
```

**Step 2 - Create a Recipe:**

Here, you will create a *recipe* and store it in the *R* object
`RecipeWine`. Use `step_rm()` to remove the predictor variable $Quality$
because it is not related to $WineColor$, and the command
`step_normalize()` to normalize all remaining predictors. Please,
substitute `<THESE>` placeholders accordingly and execute the code.

Note that the data frame `DataTrain` has already been loaded in the
background:

```{webr}
#| exercise: exc_step2
#| exercise.setup: exc_all
RecipeWine=recipe(WineColor~., data=DataTrain) |>
           ______(______) |>           
           step_normalize(all_predictors())
print(RecipeWine)
```

::: {.hint exercise="exc_step2"}
**Here is a hint:**

The step to remove a variable such as `Quality` is called `step_rm()`.
:::

::: {.solution exercise="exc_step2"}
**Here is the Solution:**


```{webr}
#| exercise: exc_step2
#| solution: true
RecipeWine=recipe(WineColor~., data=DataTrain) |>
           step_rm(Quality) |>           
           step_normalize(all_predictors())
print(RecipeWine)
```
:::

```{webr}
#| exercise: exc_step2
#| check: true
gradethis::grade_this_code()
```

**Step 3 - Create a Model Design:**

Next, you will create the *model design* and store it into the *R*
object `ModelDesignKNN`. Since you plan to tune the argument
`neighbors=` (stands for the $k$ in *Nearest Neighbors*), you have
to add it as an argument into the `nearest_neighbor()` command by
substituting `______` with the argument and its
value. Remember that you cannot set `neighbors` to a specific
numerical value because you want to *tune* the *hyper-parameter*
`neighbors` later on. Therefore you have to assign the placeholder
`tune()` to the argument `neighbors`.

```{webr}
#| exercise: exc_step3
#| exercise.setup: exc_all
ModelDesignKNN=nearest_neighbor(neighbors=______) |>
                set_engine("kknn") |>
                set_mode("classification")
print(ModelDesignKNN)
```

::: {.hint exercise="exc_step3"}
**Here is a hint:**

Instead of assigning a number to the hyper-parameter neighbors 
use the command `tune()` as a place holder.
:::

::: {.solution exercise="exc_step3"}
**Here is the Solution:**

```{webr}
#| exercise: exc_step3
#| solution: true
ModelDesignKNN=nearest_neighbor(neighbors=tune()) |>
               set_engine("kknn") |> 
               set_mode("classification")
print(ModelDesignKNN)
```
::: 

```{webr}
#| exercise: exc_step3
#| check: true
gradethis::grade_this_code()
```

**Step 4 - Add the RecipeWine and the Model Design to a Workflow:**

The code block below adds the *R* object `Recipe` and the model
design object `ModelDesignKNN` to a *workflow* model named
`TuneWFModelKNN`. The *R* code has been executed already.

```{webr}
#| autorun: true
#| exercise.setup: exc_all
TuneWFModelKNN=workflow() |> 
                add_recipe(RecipeWine) |>
                add_model(ModelDesignKNN) 
print(TuneWFModelKNN)
```

You can see in the printout above that the *workflow* is not *finalized* because the number of `neighbors` has not been set in    the *model design*. The *hyper-parameter* `neighbors` is set to
`tune()` instead. Later, in Step 7, `tune()` will be replaced in a trial
and error process with different values for `neighbors`.

**Step 5 - Create a Hyper-Parameter Grid:**

When tuning is executed in Step 7, values reaching from 1 --15 for the *hyper-parameter* `neighbors` shall be tried out. You need to provide these values in a data frame column that is named the same as the *hyper-parameter*. Below, replace the `______` to define a column `neighbors` in the data frame`ParGrid`. The column should contain values from 1 -- 15:

```{webr}
#| exercise: exc_step5
#| exercise.setup: exc_all
ParGrid=data.frame(neighbors=c(______))
print(ParGrid)
```

::: {.hint exercise="exc_step5"}
**Here is a hint:**

Use the `c()` command to create a vector containing the numbers $1 - 15$
either separated by a comma, or you can use a colon (*:*) to indicate *from/to*.
:::

::: {.solution exercise="exc_step5"}
**Here is the Solution:**

```{webr}
#| exercise: exc_step5
#| solution: true
ParGrid=data.frame(neighbors=c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15))
ParGrid
```

:::

```{webr}
#| exercise: exc_step5
#| check: true
grade_this({
  # .result refers to the output of the student's code
  if (sum(.result)==sum(1:15)) {
    pass()
  } else {
    fail()
  }
})
```

**Step 6 - Creating Resamples for Cross-Validation:**

The values you have created above for $k$ (*hyper-parameter*
`neighbors`) will be evaluated later using five folds (*resamples*).

Each fold contains the complete training data, but different 
sections are used for training and assessment in each fold.
Please create five folds below by substituting `______`
accordingly.

The folds will be saved in the *R* object `FoldsForTuning`:

```{webr}
#| exercise: exc_step6
#| exercise.setup: exc_all
set.seed(123)
FoldsForTuning=vfold_cv(DataTrain, v=______, strata=WineColor)
print(FoldsForTuning)
```

::: {.hint exercise="exc_step6"}
**Here is a hint:**

The argument `v=` determines the number of folds to be used for tuning.
:::

::: {.solution exercise="exc_step6"}
**Here is the Solution:**

```{webr}
#| exercise: exc_step6
#| solution: true
set.seed(123)
FoldsForTuning=vfold_cv(DataTrain, v=5, strata=WineColor)
print(FoldsForTuning)
```
:::

```{webr}
#| exercise: exc_step6
#| check: true
gradethis::grade_this_code()
```

**Step 7 - Tune the Workflow and Train All Models:**

Now it is time to run the tuning procedure using the `tune_grid()`
command. Be patient because it will take some time to fully execute.
Since we have to try out 15 parameters and use five folds for each
model, the `tune_grid()` command has to fit 75 models
$(15\cdot 5=75)$.

Please substitute `______` with a list of metrics to be
calculated. Use the metrics *accuracy*, *specificity*, and
*sensitivity*.

After the tuning is finished, all results are stored in the *R*
object `TuneResults`, and they can be evaluated by different
*metrics* commands.

For example the command `autoplot()` provides a diagrammatic
overview of the results for all three *metrics*.

```{webr}
#| exercise: exc_step7
#| exercise.setup: exc_all
TuneResults=tune_grid(TuneWFModelKNN, resamples=FoldsForTuning,
                      grid=ParGrid, metrics=metric_set(______))
autoplot(TuneResults)
```

::: {.hint exercise="exc_step7"}
**Here is a hint:**

The command metric_set takes a comma-separated list 
of metrics to be evaluated.

Consider that you want to evaluate the metrics `accuracy`, `specificity`, 
`sensitivity`.
:::

::: {.solution exercise="exc_step7"}
**Here is the Solution:**

```{webr}
#| exercise: exc_step7
#| solution: true
TuneResults=TuneWFModelKNN |> tune_grid(FoldsForTuning, grid=ParGrid ,  
                       metrics=metric_set(accuracy, specificity, sensitivity))
autoplot(TuneResults)
```
:::

```{webr}
#| exercise: exc_step7
#| check: true
gradethis::grade_this_code()
```

The three graphs that you created in the exercise above show for
each *metric* the related *accuracy*, *specificity*, and
*sensitivity* for all tried out *hyper-parameters* (`neighbors`).
The results for the five folds are averaged.

You can see that `neighbors` values between 1 and 4 produce the best
results for *accuracy* (predicting red and white wines (positive and
negative class) correctly) and *sensitivity* (predicting red wines
(positive class) correctly). If you look at *specificity*
(predicting white wines (negative class) correctly), you can see
that $k=5$ produces the best result. However, $k=5$ also leads to a
sharp decrease in *sensitivity* and a decrease in *accuracy* as
well.

It seems reasonable to use the best result for *accuracy*, which
means choosing a $k$ between 1 and 4.

**Step 8 - Extract the Best Hyper-Parameter(s):**

All assessment results for the specified metrics are stored in the
*tuning* object `TuneResults`. Choose the metric *accuracy* by
substituting `______` accordingly. Afterward, when you execute the
code the `select_best()` command extracts the best *hyper-parameter*
(value for `neighbors`) for the metric you specified:

```{webr}
#| exercise: exc_step8
#| exercise.setup: exc_all
BestHyperPar=select_best(TuneResults, metric="______") 
print(BestHyperPar)
```

::: {.hint exercise="exc_step8"}
**Here is a hint:**

Substitute `______` with `accuracy` and make sure to not delete the quotes.
:::

::: {.solution exercise="exc_step8"}
**Here is the Solution:**

```{webr}
#| exercise: exc_step8
#| solution: true
BestHyperPar=select_best(TuneResults, metric="accuracy") 
print(BestHyperPar)
```
:::

```{webr}
#| exercise: exc_step8
#| check: true
gradethis::grade_this_code()
```
As you will see in the printout after executing the *R* code, the
best value for the *hyper-parameter* `neighbors` based on *accuracy*
is $neighbors=$ `r  BestHyperPar[[1,1]]`. It is
saved in the data frame `BestHyperPar`.

You can try other metrics in the code block above and see if the
result changes. 

**Step 9 - Finalize and Train the Best Workflow Model:**

The code block below is executed already. The `finalize_workflow()`
command used the value from `BestHyperPar` to substitute the
`tune()` placeholder in the *R* object `TuneWFModelKNN` (created in
Steps 2 -- 4). The *hyper-parameter* is now set to $neighbors=$ `r  BestHyperPar[[1,1]]`.

    
```{webr}
#| autorun: true
#| exercise.setup: exc_all
WFModelBest=TuneWFModelKNN |> 
            finalize_workflow(BestHyperPar) |>  
            fit(DataTrain)

print(WFModelBest)
```

At the end of the printout above, you can see that $(k)$ was set to
`neighbors=1` rather than
`tune()`. You can also see that `WFModelBest` is fitted to the
training data because the "Minimum missclassification" is reported
(based on the training data).

**Step 10: Assess Prediction Quality Based on the Testing Data:**

Since `WFModelBest` is a fitted model, you can use it for
predictions. In this last step, you use the `augment()` command to
predict `WineColor`. The `augment()` command will then add the
prediction results as column `.pred_class` to the testing data.

The `conf_mat()` command compares the predictions in column `.pred_class`
to the true values to create a *confusion matrix*. But before, you
have to substitute `______` with the variable name
(column name) for the outcome variable.

```{webr}
#| exercise: exc_step10
#| exercise.setup: exc_all
DataTestWithPredBestModel=augment(WFModelBest, DataTest)    
conf_mat(DataTestWithPredBestModel, truth=______, 
             estimate=.pred_class)
```

::: {.hint exercise="exc_step10"}
**Here is a hint:**

The (column) name for the outcome variable is `WineColor` and the predictions are stored in the column `.pred_class`.
:::

::: {.solution exercise="exc_step10"}
**Here is the Solution:**

```{webr}
#| exercise: exc_step10
#| solution: true
DataTestWithPredBestModel=augment(WFModelBest,DataTest)
conf_mat(DataTestWithPredBestModel, truth=WineColor, estimate=.pred_class)
```
:::

You can see in the *confusion matrix* that based on the testing
data from the 480 red wines only `r ConFusMat$table["white", "red"]` were misclassified as being *white*. Likewise, from
the 480 white wines, only `r ConFusMat$table["red", "white"]` were misclassified. Given that the
classification was only based on the chemical properties of the
wines, the results are excellent.

Most likely, a true wine expert might have reached a similar
impressive result, but it would have taken them a long time to
classify 960 wines.
