---
title: "Interactive Section"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: Exercises

---

<style>
  div.yellowbox {
    background-color: #f1dfa6;
    border: 1px solid black;
    padding: 21px;
  }

  strong {
    color: #d10056
  }

  .center {
    text-align: center;
  }
</style>


```{r setup, include=FALSE}
library(learnr)
library(rio)
library(tidyverse)
library(tidymodels)
library(janitor)
library(kableExtra)

knitr::opts_chunk$set(exercise=TRUE, exercise.eval=FALSE, exercise.lines=6, attr.source = ".numberLines")
```

:::: {.yellowbox data-latex=""}

::: {.center data-latex=""}
**How to Work with this Interactive Section**
:::

Here you will find the complete interactive section from the book, including book content, and *R* code blocks. Read the content in a browser as you would read it in the textbook. 

Within the *R* code blocks, you can work in various ways:

* Use the `Run Code` button to execute the *R* code and see the result.

* Change the *R* code, click the `Run Code` button, and then see the new results.

* To get the original *R* code back, click the `Start Over` button at the left-top corner of each *R* code block. 

* To reset the *R* code for all code blocks, click the `Start Over` in the left menu.

**Make sure you installed all needed packages!** See Section "R Packages Required"  in this Chapter for a list of packages that need to be installed. Section 3.2 explain how to install a packege in *RStudio*.



**For Troubleshooting:** [Click here](https://blog.lange-analytics.com/2024/01/interactsessions.html)
::::


 
 
 <clex id="Exc100">

In the previous section, we worked with a univariate *Polynomial Model* to demonstrate *regularization* and to evaluate *overfitting* scenarios. In the real world, data scientists usually use multivariate models to include more than one variable in their analysis. Therefore, this interactive exercise will teach you how to use a multivariate approach.

```{r RidgeLasso-Exerc100-LoadHouseData, message=FALSE, warning=FALSE, exercise=FALSE, include=FALSE}
library(tidymodels); library(rio); library(janitor); library(glmnet)

set.seed(777)
DataHousing=import("https://ai.lange-analytics.com/data/HousingData.csv") |>
  clean_names("upper_camel") |> 
  select(Price,Sqft=SqftLiving, Bedr=Bedrooms, Year=YrBuilt) |> 
  filter(Price<=1300000, Sqft<4500) 

set.seed(777)
Split005=DataHousing |> 
  initial_split(prop=0.005, strata = Price, breaks = 5)
DataTrain=training(Split005)
DataTest=testing(Split005)

RecipeHousesMultivar=recipe(Price~., data=DataTrain) |> 
  step_mutate(Sqft2=Sqft^2, Bedr2=Bedr^2, Year2=Year^2,
              SqftxBedr=Sqft*Bedr, SqftxYear=Sqft*Year, 
              BedrxYear=Bedr*Year, Sqft2xBedr=Sqft2*Bedr,
              SqftxBedr2=Sqft*Bedr2, Sqft2xBedr2=Sqft2*Bedr2,
              Sqft2xYear=Sqft2*Year, SqftxYear2=Sqft*Year2,
              Sqft2xYear2=Sqft2*Year2, Bedr2xYear=Bedr2*Year,
              BedrxYear2=Bedr*Year2,Bedr2xYear2=Bedr2*Year2)|> 
  step_normalize(all_predictors())

ModelDesignBenchmark=linear_reg() |> 
                  set_engine("lm") |> 
                  set_mode("regression")

WFModelBenchmark=workflow() |> 
  add_model(ModelDesignBenchmark) |> 
  add_recipe(RecipeHousesMultivar) |> 
  fit(DataTrain)

ModelDesignElastNetTune=linear_reg(penalty=tune(), mixture=tune()) |>
                        set_engine("glmnet") |> 
                        set_mode("regression")

WFModelElastNetTune=workflow() |>
add_model(ModelDesignElastNetTune) |>
add_recipe(RecipeHousesMultivar)

set.seed(777)
FoldsForTuning=vfold_cv(DataTrain, v=3, strata=Price, repeats=10)

```

### The Data

You \index{Housing Dataset} \index{Datasets!Housing} will again predict house prices using the **King County House Sale** dataset (@Kaggle2015). In the code block below, the dataset is loaded. Instead of selecting only one variable to predict the price of a house $(Price_i)$, we use the `select()` command to choose the number of bedrooms $(Bedr_i)$ and the year a house was built $(Year)$, together with the house's square footage $(Sqft)$. We filter out houses with a price of more than \$1,300,000 and square footage 4,500 sqft or more, which are outliers:

```{r RidgeLasso-Exerc100-Data, eval=TRUE, message=FALSE, warning=FALSE, exercise=FALSE}
library(tidymodels); library(rio); library(janitor); library(glmnet)

DataHousing=
  import("https://ai.lange-analytics.com/data/HousingData.csv")|>
  clean_names("upper_camel") |> 
  select(Price, Sqft=SqftLiving, Bedr=Bedrooms, Year=YrBuilt) |> 
  filter(Price<=1300000, Sqft<4500) 

set.seed(777)
Split005=initial_split(DataHousing, prop=0.005, strata=Price, breaks=5)
DataTrain=training(Split005)
DataTest=testing(Split005)
```

Note when creating the training dataset, we use the argument `prop=0.005`, leading to a training dataset with about 100 observations. This size is small, but in general realistic for a training dataset. However, given a total of `r format(nrow(DataHousing), big.mark=",")` available observations, we could have used the more common `prop=0.7`. In that case, we would end up with `r format(round(nrow(DataHousing)*0.7), big.mark=",")` observations for the training dataset. With such a large training dataset, the unregularized model we use later in this section would likely not lead to *overfitting*, and thus *regularization* would no longer be needed --- which would defeat the purpose of this interactive exercise.

### Unregularized Benchmark Model {#RidgeLasso-Exerc100-Unregularaized}

As a benchmark, we begin with an unregularized model. In this case, a multivariate *Degree-2 Polynomial Model*. It includes the variables $Sqft$, $Bedr$, and $Year$. They are used in their original form, squared, and as various *interaction terms*:

```{=tex}
\begin{eqnarray}
Price&=&
\beta_1 Sqft+\beta_2 Bedr+\beta_3 Year\nonumber \\
&+&\beta_4 Sqft^2+\beta_5 Bedr^2+\beta_6 Year^2\nonumber \\
&+&\beta_7 Sqft Bedr+\beta_8 Sqft Year+\beta_9 Bedr Year\nonumber \\
&+&\beta_{10} Sqft^2 Bedr+\beta_{11} Sqft Bedr^2+\beta_{12} Sqft^2 Bedr^2\nonumber \\
&+&\beta_{13} Sqft^2 Year+\beta_{14} Sqft Year^2+\beta_{15} Sqft^2 Year^2\nonumber \\
&+&\beta_{16} Bedr^2 Year+\beta_{17} Bedr Year^2+\beta_{18} Bedr^2 Year^2
(\#eq:RidgeLasso-MultiVarPredict)
\end{eqnarray}
```
As before, we use a *recipe* to generate and normalize the additional predictors:

```{r RidgeLasso-Exerc100-Recipe, eval=TRUE, message=FALSE, warning=FALSE, exercise=FALSE}
RecipeHousesMultivar=recipe(Price~., data=DataTrain) |> 
                     step_mutate(Sqft2=Sqft^2,Bedr2=Bedr^2,Year2=Year^2,
                       SqftxBedr=Sqft*Bedr,SqftxYear=Sqft*Year, 
                       BedrxYear=Bedr*Year,Sqft2xBedr=Sqft2*Bedr, 
                       SqftxBedr2=Sqft*Bedr2,Sqft2xBedr2=Sqft2*Bedr2,
                       Sqft2xYear=Sqft2*Year,SqftxYear2=Sqft*Year2,
                       Sqft2xYear2=Sqft2*Year2,Bedr2xYear=Bedr2*Year,
                       BedrxYear2=Bedr*Year2,Bedr2xYear2=Bedr2*Year2) |> 
                     step_normalize(all_predictors())
```

Afterward, we create the *model design* (`ModelDesignBenchmark`), the *workflow model* (`WFModelBenchmark`), and we calibrate `WFModelBenchmark` to the training data:

```{r RidgeLasso-Exerc100-WorkflowBenchmark, eval=TRUE, message=FALSE, warning=FALSE, exercise=FALSE}
ModelDesignBenchmark=linear_reg() |> 
                     set_engine("lm") |> 
                     set_mode("regression")

WFModelBenchmark=workflow() |> 
                 add_model(ModelDesignBenchmark) |> 
                 add_recipe(RecipeHousesMultivar) |> 
                 fit(DataTrain)
```

When you execute the `tidy()` command in the code block below, you will see that all predictor variable parameters ($\beta s$), except the one for the intercept, are in the multi-million range, and some even in the multi-billion range (see column `estimate`).

```{r RidgeLasso-Exerc100-TidyCoeffBenchmark, exercise.setup="RidgeLasso-Exerc100-LoadHouseData", exercise=TRUE, eval=FALSE, exercise.lines=1}
tidy(WFModelBenchmark)
```

When you execute the code block below, the model's training and testing data predictions are compared to the true values ($Price_i$), and the related metrics are calculated:

```{r RidgeLasso-Exerc100-TrainPerfBenchmark, exercise.setup="RidgeLasso-Exerc100-LoadHouseData", exercise=TRUE, eval=FALSE, exercise.lines=8}
DataTrainWithPredBenchmark=augment(WFModelBenchmark,DataTrain)
print("Metrics for Training Data:")
metrics(DataTrainWithPredBenchmark, truth=Price, estimate=.pred)

DataTestWithPredBenchmark=augment(WFModelBenchmark,DataTest)
print("Metrics for Testing Data:")
metrics(DataTestWithPredBenchmark, truth=Price, estimate=.pred)
```

```{r RidgeLasso-Exerc100-DataForPargraphBelow, eval=TRUE, message=FALSE, warning=FALSE, exercise=FALSE, include=FALSE}
DataTrainWithPredBenchmark=augment(WFModelBenchmark,DataTrain)
FirstNumber=metrics(DataTrainWithPredBenchmark, truth=Price, estimate=.pred)[[3,3]]
DataTestWithPredBenchmark=augment(WFModelBenchmark,DataTest)
SecondNumber=metrics(DataTestWithPredBenchmark, truth = Price, estimate = .pred)[[3,3]]
```

After executing the code above, you will see that the `mae` for the training data is about \$`r format(round(FirstNumber,-2), big.mark = ",")` and the `mae` for the testing data is about \$`r format(round(SecondNumber,-2), big.mark = ",")`. Such a low training `mae`, when compared to the testing `mae`, strongly indicates *overfitting*.

In the following section, you will use an *Elastic-Net* model to improve the predictive quality.

### Regularized Elastic-Net Polynomial Model {#RidgeLasso-ElastNet}

When you substitute the $\dots$ after `mixture=` and `penalty=` in the code further below with values of your choice and execute, the model's $\beta$ values will be calibrated and printed. Also, the metrics for the testing data are calculated and printed.

Recall, if `mixture=0`, you will run a *Ridge* model; if `mixture=1`, you will run a *Lasso* model. For anything in between, you will run a *mixed* model.

Finding an appropriate value for the argument `penalty=` is more tricky. For the *Ridge* *penalty*, values between $15{,}000$ and $154{,}000{,}000$ will work. For the *Lasso* *penalty*, values between $15$ and $154{,}000$ will work.[^07-ridgelasso-5]

[^07-ridgelasso-5]: These recommendations are based on the 100 $\lambda$ values that `glmnet` analyzes by default, depending on the model and the training data. Values outside these ranges might produce surprising results.

Try a few settings and see how the results are changing. For example, use `mixture=1` (*Lasso*) and `penalty=100000`. Afterward, increase and decrease `penalty=` and leave `mixture=1` constant. How are the results changing? Then try `mixture=0` and leave `penalty=100000` to see the different results for a *Ridge* model. You can also set `mixture=0.5` or `mixture=0.7` to try out different *Elastic-Net* models. Playing with different *hyper-parameter* values and observing the resulting $\beta s$ and *metrics* will help you to better understand the $Ridge$, $Lasso$, and *Elastic-Net* approaches.

```{r RidgeLasso-Exerc100-TrialRegularization, exercise.setup="RidgeLasso-Exerc100-LoadHouseData", eval=FALSE, message=FALSE, warning=FALSE, exercise=TRUE, exercise.lines=15}
ModelDesignRegularized=linear_reg(penalty=... , mixture=...) |> 
                       set_engine("glmnet") |> 
                       set_mode("regression")

WFModelRegularized=workflow() |> 
                   add_model(ModelDesignRegularized) |> 
                   add_recipe(RecipeHousesMultivar) |> 
                   fit(DataTrain)

print("Beta Values (see Column Estimate:)")
tidy(WFModelRegularized)

DataTestWithPredRgularized=augment(WFModelRegularized, DataTest)
print("Metrics for Testing Data:")
metrics(DataTestWithPredRgularized, truth=Price, estimate=.pred)
```

**Warning:** In the code block above, you changed *hyper-parameters* and evaluated the effects based on the testing data to better understand *regularization*. **Never use the testing data to find the best hyper-parameters.** Otherwise, you end up with a model that is highly specialized to the testing data, but when used on new data in the production phase, it would likely perform very poorly.

Instead, use validation data derived from the training dataset to evaluate the performance of *hyper-parameters*. This is the underlying principle of *tuning*, and in the next section, you will tune the *workflow model* to improve performance.

### Tuning the Elastic-Net Polynomial Model {#RidgeLasso-TuneElasticNet}

In this section, you will use the *10-Step Template* to tune the *hyper-parameters* `penalty` and `mixture`.

**Step 1 - Generate Training and Testing Data:**

:   The training and testing data have already been generated (see Section \@ref(RidgeLasso-ElastNet) above).

**Step 2 - Create a Recipe:**

:   The *recipe* to create the predictors for the multivariate *Degree-2 Polynomial* model (`RecipeHousesMultivar)` was created in connection with the unregularized model and can be reused here. It establishes the predictors with various powers and the interaction terms.

**Step 3 - Create a Model Design:**

:   The *model design* from the previous section needs slight modifications. We have to add the `tune()` placeholder for the *hyper-parameters*:

    ```{r RidgeLasso-Exerc100-TuneModelDesign, exercise=FALSE, eval=FALSE}
    ModelDesignElastNetTune =linear_reg(penalty=tune(), mixture=tune()) |> 
                             set_engine("glmnet") |> 
                             set_mode("regression")
    ```

**Step 4 - Add the Recipe and the Model Design to a Workflow:**

:   As before, we add the *recipe* and the (modified) *model design* to the *workflow*:

    ```{r RidgeLasso-Exerc100-TuneWorkflow, exercise=FALSE, eval=FALSE}
    TuneWFModel=workflow() |>
                add_recipe(RecipeHousesMultivar) |>     
                add_model(ModelDesignElastNetTune)
    ```

**Step 5 - Create a Hyper-Parameter Grid:**

:   Previously, when only one *hyper-parameter* was tuned, you added the values to be tried out in a data frame column named with the same name as the *hyper-parameter*. When you have more than one *hyper-parameter*, it gets a little more complicated because you want to try out different combinations of *hyper-parameters*.

    If you decide to try out all combinations of the values from two or more *hyper-parameters*, you can use the `crossing()` command. The arguments of the `crossing()` command are the *hyper-parameters* you would like to try out. The assigned values to these arguments are the values you want to try out for each of the *hyper-parameters*. They are assigned to each argument as *R* `vector` objects. The `crossing()` command then generates all possible combinations for the values of the *hyper-parameters*.

    For example, in this project you will later try 66 *penalty* values from 15,000 to 80,000 with an increment of 1,000 and three *mixture* values (0, 0.5, and 1). You can use the crossing() command to generate all possible combinations:

    ```{r RidgeLasso-Exerc100-TuneHypParGrid, exercise=FALSE, eval=TRUE}
    ParGridHouses=crossing(mixture=c(0, 0.5, 1),
                           penalty=seq(15000, 80000, 1000))
    print(ParGridHouses)
    ```

    The *R* `vector` object assigned to `mixture` is created with the `c()` command by listing the related values. The *R* `vector` object assigned to `penalty` is created with the `seq()` command. The first value in `seq()` determines where the sequence starts ($15{,}000$) and the second where it ends ($80{,}000$). The third argument determines the increment (the step size; $1{,}000$).

    You can see in the printout above that `crossing()` generates all possible combinations of the *penalty* values and the *mixture* values. Each of these 198 combinations will be tried out when you tune the *workflow* later on.

    The `crossing()` command is very helpful when you combine multiple *hyper-parameters* with several values each. Imagine you want to create a grid with all combinations of five *hyper-parameters* with ten values each, `crossing()` would create $100{,}000$ parameter combinations ($10\cdot 10\cdot 10\cdot 10\cdot 10=10^5$) for you. Be advised, tuning $100{,}000$ *hyper-parameter* combinations would use a lot of computing time.

    In general, computing time increases exponentially with the number of *hyper-parameters*, while computing time increases proportionally when the values for each *hyper-parameter* increase. These facts must be considered when deciding how many *hyper-parameters* you want to tune and how many values you like to consider for each *hyper-parameter*.

**Step 6: - Create Resamples for Cross-Validation:**

:   To create *resamples* for *Cross-Validation*, we use the `vfold()` command. However, although usually recommended, creating ten *Cross-Validation* folds creates a problem. Given that we have only `r nrow(DataTrain)` observations, the validation set of each fold would consist only of about ten observations. Therefore, we will create sets of three folds (`v=3`), which leaves about $30$ observation for each validation dataset. To offset the small number of folds, we repeat the process of creating folds ten times (see the argument `repeats=10`) creating a total of $30$ folds:

    ```{r RidgeLasso-Exerc100-TuneCreateFolds, exercise=FALSE, eval=FALSE}
    FoldsForTuning=vfold_cv(DataTrain, v=3, strata=Price, repeats=10)
    ```

**Step 7 -- Step 10:**

Now, you are ready to finalize the tuning process in the interactive project. The training and testing data (from Step 1), the *workflow* including the *recipe* and the *model design* (from Steps 2 -- 4), as well as the *resamples* for validation (from Step 6) have already been loaded in the background.

In the code block below, you only have to define the data frame with the *parameter grid* as explained in Step 5 to try out various combinations for the *hyper-parameters* `penalty=` and `mixture=`.

For the first time you run the code, it is recommended to use the values as determined in Step 5 (you can later change them). Note, executing the code can take a while because `penalty=seq(15000,80000,1000)` will create 66 *penalty* values, and `mixture=seq(0,1,0.5)` will create 3 *mixture* values. The `crossing()` command will then create 198 combinations ($66\cdot3$) which will be all tried out and evaluated in the tuning process in Step 7.

When *R* outputs the results, look carefully at the $\beta$ parameters, the plot generated by `autoplot()` after Step 7, and the metrics for the testing data. Then change the `c()` and `seq()` commands that generate the *hyper-parameter* values for `penalty=` and `mixture=`. Afterward, execute the code again and see how the results change.

Please recall, that for this setup *Ridge* *penalty* values between $15{,}000$ and $154{,}000{,}000$ will work and *Lasso* *penalty* values between $15$ and $154{,}000$ will work.[^07-ridgelasso-6]

[^07-ridgelasso-6]: These recommendations are based on the 100 $\lambda$ values that `glmnet` tries by default, depending on the model and the training data. Values outside these ranges might produce surprising results.

```{r RidgeLasso-Exerc100-TuneElstNetExerc, exercise.setup="RidgeLasso-Exerc100-LoadHouseData", exercise=TRUE, eval=FALSE, exercise.lines=37}
# Step 5 - Create a Hyper-Parameter Grid: 
#
# ParGridHouses contains the hyper-parameter combinations to be 
# tried out.
# Change the values for the seq() commands to try out different values.
ParGridHouses=crossing(penalty=seq(15000, 80000, 1000), 
                       mixture=seq(0, 1, 0.5))
print("Hyper-Parameter Combinations to be Tried Out:")
ParGridHouses

# Step 7: Tune the Workflow and Train All Models
set.seed(777)
TuneResultsElastNet=tune_grid(WFModelElastNetTune,  
                              resamples=FoldsForTuning, 
                              grid=ParGridHouses, 
                              metrics = metric_set(rmse))

# The diagram shows how the  *root mean squared error* (`rmse`) 
# changes with increasing regularization:
autoplot(TuneResultsElastNet)

# Step 8 - Extract the Best Hyper-Parameter(s):
BestHyperPara=select_best(TuneResultsElastNet, metric="rmse")
print("Best values for the Hyper-Parameters")
BestHyperPara

# Step 9: Finalize and Train the Best Workflow 
BestWFModel=finalize_workflow(WFModelElastNetTune, BestHyperPara) |> 
            fit(DataTrain)

print("Optimal Beta Parameters (see estimate column)")
tidy(BestWFModel)

# Step 10: Assess Prediction Quality Based on the Testing Data:
DataTestBestWFModelWithPred=augment(BestWFModel, DataTest)
print("Metrics for the Testing Data")
metrics(DataTestBestWFModelWithPred, truth = Price, estimate = .pred)
```

```{r RidgeLasso-Exerc100-TuneElstNetExerc-hint, exercise=FALSE,  eval=FALSE, include=FALSE, exercise.lines=17}
For the first trial, leave ParGridHouses in Step 5 as it is and 
just execute the code.

It takes a while. So be patient and have a coffee.

Next, try different *hyper-parameters* by changing the seq() commands 
in crossing(). What are the consequences in terms of beta parameters and
performance?

The next HINT provides a tuning that works faster. 

In the "solution", mixture is changed to:
mixture=seq(0,1,1)  
using only values 0 (Ridge) and 1 (Lasso) but no mix.

In the solution penalty is changed to:
penalty=seq(15000,80000,2000) covering the same range but with 
a higher increment (2000) and thus fewer values.
```

```{r RidgeLasso-Exerc100-TuneElstNetExerc-solution, exercise=FALSE,  eval=FALSE, include=FALSE, exercise.lines=34}
# Step 5 - Create a Hyper-Parameter Grid: 
#
# ParGridHouses contains the *hyper-parameter* combinations to be tried out.
# Change the values for the seq() commands to try out different values.
ParGridHouses=crossing(penalty=seq(15000,80000,2000), mixture=c(0,1))
print("Hyper-Parameter Combinations to be Tried Out:")
ParGridHouses

# Step 7: Tune the Workflow and Train All Models
set.seed(777)
TuneResultsElastNet=tune_grid(WFModelElastNetTune,  
                              resamples=FoldsForTuning, 
                              grid=ParGridHouses, 
                              metrics = metric_set(rmse))

# The diagram shows how the rmse changes with increasing regularization:
autoplot(TuneResultsElastNet)

# Step 8 - Extract the Best Hyper-Parameter(s):
BestHyperPara=select_best(TuneResultsElastNet, metric="rmse")
print("Best values for the Hyper-Parameters")
BestHyperPara

# Step 9: Finalize and Train the Best Workflow 
BestWFModel=finalize_workflow(WFModelElastNetTune, BestHyperPara) |> 
            fit(DataTrain)

print("Optimal Beta Parameters (see estimate column)")
tidy(BestWFModel)

# Step 10: Assess Prediction Quality Based on the Testing Data:
DataTestBestWFModelWithPred=augment(BestWFModel, DataTest)
print("Metrics for the Testing Data")
metrics(DataTestBestWFModelWithPred, truth = Price, estimate = .pred)
```

</clex>