---
title: "Interactive Section"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: Exercises

---

<style>
  div.yellowbox {
    background-color: #f1dfa6;
    border: 1px solid black;
    padding: 21px;
  }

  strong {
    color: #d10056
  }

  .center {
    text-align: center;
  }
</style>


```{r setup, include=FALSE}
library(learnr)
library(rio)
library(tidyverse)
library(tidymodels)
library(janitor)
library(kableExtra)

knitr::opts_chunk$set(exercise=TRUE, exercise.eval=FALSE, exercise.lines=6, attr.source = ".numberLines")
```

:::: {.yellowbox data-latex=""}

::: {.center data-latex=""}
**How to Work with this Interactive Section**
:::

Here you find the complete interactive section from the book, including book content, and R-code-blocks. Read the content in a browser as you would read it in the textbook. 

With the R-code blocks, you can work in various ways:

* Use the `Run Code` button to execute the R-code and see the result.

* Change the R-code, click the `Run Code` button, and then see the new results.

* To get the original R-code back, click the `Start Over` button at the left-top corner of each R-code block. 

* To reset the R-code for all code blocks, click the `Start Over` in the left menu.

**Make sure you installed all needed packages!** See Section "R Packages Required"  in this Chapter for a list of packages that need to be installed. Section 3.2 explain how to install a packege in *RStudio*.



**For Troubleshooting:** [Click here](https://blog.lange-analytics.com/2024/01/interactsessions.html)
::::


 
 
 <clex id="Exc100">

```{r Interpret-Exerc100-LoadDataEtc, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, exercise=FALSE, include=FALSE}
# Make sure to set DraftMode=FALSE at the beginning of the rmd file eval=!DraftMode

library(rio); library(janitor); library(tidymodels)
library(DALEX); library(DALEXtra)
DataVaxOrg=import("https://ai.lange-analytics.com/data/DataVax.rds") 

DataCountyInfo=DataVaxOrg |> 
               mutate(CountyID=row_number()) |> 
               select(CountyID, Fips, County, State)

DataVax=DataVaxOrg |>   
        select(PercVacFull,PercRep,
               PercAsian,PercBlack,PercHisp,
               PercYoung25,PercOld65, 
               PercFoodSt,Population) |> 
               mutate(Population=frequency_weights(Population))
set.seed(2021)
Split85=DataVax |> initial_split(prop=0.85,
                                 strata=PercVacFull,
                                 breaks=3)

DataTrain=training(Split85)
DataTest=testing(Split85)

set.seed(777)
RecipeVax=recipe(PercVacFull~., data=DataTrain)
  
ModelDesign=rand_forest() |> 
    set_engine("ranger") |> 
    set_mode("regression")
  
WfModelVax=workflow() |>
    add_model(ModelDesign) |>
    add_recipe(RecipeVax) |> 
    fit(DataTrain)

DataTrainPredOnly=DataTrain|> 
          select(-PercVacFull,-Population)
ExplainerRandForest=explain_tidymodels(WfModelVax, 
                    data=DataTrainPredOnly,y=DataTrain$PercVacFull, verbose = FALSE) 
```

To generate a *SHAP* diagram for a continental U.S. county of your choice, you have to proceed in two steps:

-   In *Step 1* you enter a search string for the county of your choice, and you will get a *CountyID* as a result. This *CountyID* is the row in the data frame `DataVax` representing the observation of your county, and you need to enter it in *Step 2*.

-   In *Step 2* you enter the *CountyID* in the second code chunk and execute the code. *R* will run in your browser and generate the *SHAP* diagram for you. This can take a while. So, be a little patient.

**Step 1:**

The code chunk below helps you to find the $CountyID$ for the county you are interested in. You need the $CountyID$ for the next step below to create the *SHAP* diagram.

To find the $CountyID$, substitute the $/dots$ in the code block below with the name or part of the name of the county you are interested in (do not include the word "county"!):

```{r Interpret-Exerc100-FindCounty, eval=FALSE, exercise=TRUE, exercise.lines=5, exercise.setup="Interpret-Exerc100-LoadDataEtc", exercise.timelimit=1800}
CountySearchStr="..."

######## Do not change code below this line ###################
VecMatchingCountyNames=grep(CountySearchStr, DataCountyInfo$County, 
                       ignore.case=TRUE)
VecMatchingFIPS=grep(CountySearchStr, as.character(DataCountyInfo$Fips), 
                       ignore.case=TRUE)
VecMatching=sort(c(VecMatchingCountyNames, VecMatchingFIPS))
unique(DataCountyInfo[VecMatching,])
DataMatchingCounties=DataCountyInfo[VecMatching,]
print(DataMatchingCounties)
```

```{r Interpret-Exerc100-FindCounty-hint, exercise=FALSE,  eval=FALSE, include=FALSE, exercise.lines=17}
'To find the CountyID for Los Angeles County, you could use:'

'CountySearchStr="Los Angeles" or CountySearchStr="Angeles"

'Do not include the word "County" in your search string!'

'The next HINT provides the solution for Los Angeles.'
```

```{r Interpret-Exerc100-FindCounty-solution, exercise=FALSE,  eval=FALSE, include=FALSE, exercise.lines=17}
CountySearchStr="Los Angeles"

######## Do not change code below this line ###################
VecMatchingCountyNames=grep(CountySearchStr, DataCountyInfo$County, 
                       ignore.case=TRUE)
VecMatchingFIPS=grep(CountySearchStr, as.character(DataCountyInfo$Fips), 
                ignore.case=TRUE)
VecMatching=sort(c(VecMatchingCountyNames, VecMatchingFIPS))
unique(DataCountyInfo[VecMatching,])
DataMatchingCounties=DataCountyInfo[VecMatching,]
print(DataMatchingCounties)
```

**Step 2:**

Here you can create the *SHAP* diagram and the related table with some extra info. All you have to do is substitute the $\dots$ in the code block below with the *CountyID* that you got in *Step 1* (the number only; no quotes).

Executing the code takes a while (depending on your computer between 1 -- 8 minutes). You can change the `B=` argument (the number of scenarios to be calculated) to a lower value to speed things up, but the trade-off is a lower precision.

After you get the results, take a moment to interpret the *Shap* values and decide which *SHAP* values should not be interpreted (hint: take the value of the *SHAP value* and the box plot into account.)

```{r Interpret-Exerc100-GenerateShap, eval=FALSE, exercise=TRUE, exercise.lines=10, exercise.setup="Interpret-Exerc100-LoadDataEtc", exercise.timelimit=1800}
CountyID= ...

set.seed(2021)

SHAPObject=predict_parts(
           explainer=ExplainerRandForest, 
           new_observation=DataVax[CountyID,], 
           type="shap", B=25)

### Do not change code below this line ##########

CountyPred=round(attr(SHAPObject, "prediction"),3)
VacRateCoalNone=round(attr(SHAPObject, "intercept"),3)    
TitleShap=paste(substr(DataCountyInfo[[CountyID,3]],1,15), 
                "(PredVac=", CountyPred,
                "Coalition: 'none'=", VacRateCoalNone,")")
    
SHAPObject$label=TitleShap
plot(SHAPObject)

ShapTable=SHAPObject |> 
          as_tibble() |> 
          filter(B==0) |> 
          select(Variable=variable_name, SHAP=contribution,
                 County=variable_value) |> 
          mutate(Nation=c(0.063,0.136,0.115,0.191,0.173, 
                          0.478,0.310)) |> 
          mutate(County= round(as.numeric(County),3))|> 
          mutate(SHAP= round(SHAP,3))

kable(ShapTable, caption = TitleShap) |> 
kable_styling(bootstrap_options=c("striped","hover"), 
              position="center", full_width = T)
```

```{r Interpret-Exerc100-GenerateShap-hint, exercise=FALSE,  eval=FALSE, include=FALSE, exercise.lines=5}
'To generate the SHAP Diagram  and table for Los Angeles, use:'

'CountyID=133'


'The next HINT provides the solution for Los Angeles.'
```

```{r Interpret-Exerc100-GenerateShap-solution, exercise=FALSE,  eval=FALSE, include=FALSE, exercise.lines=10}

CountyID= 133

set.seed(2021)

SHAPObject=predict_parts(
           explainer=ExplainerRandForest, 
           new_observation=DataVax[CountyID,], 
           type="shap", B=25)

### Do not change code below this line ##########

CountyPred=round(attr(SHAPObject, "prediction"),3)
VacRateCoalNone=round(attr(SHAPObject, "intercept"),3)    
TitleShap=paste(substr(DataCountyInfo[[CountyID,3]],1,15), 
                "(PredVac=", CountyPred, 
                "Coalition: 'none'=", VacRateCoalNone,")")
    
SHAPObject$label=TitleShap
plot(SHAPObject)

ShapTable=SHAPObject |> 
          as_tibble() |> 
          filter(B==0) |> 
          select(Variable=variable_name, SHAP=contribution,
                 County=variable_value) |> 
          mutate(Nation=c(0.063,0.136,0.115,0.191,0.173, 
                          0.478,0.310)) |> 
          mutate(County= round(as.numeric(County),3))|> 
          mutate(SHAP= round(SHAP,3))

kable(ShapTable, caption = TitleShap) |> 
kable_styling(bootstrap_options=c("striped","hover"), 
              position="center", full_width = T)
```

</clex>