---
title: "Hypothesis Tests"
subtitle: "Inference, Sigificance, Coefficient of Determination ($R^2$)"
format: 
  revealjs
multiplex: true
code-fold: false
scrollable: true
echo: true
incremental: false
smaller: false
---



## What Will You Learn {.scrollable .smaller}

```{r}
#| echo: false
library(tidyverse)
library(rio)
library(kableExtra)
library(janitor)
```


- What are **Statistics** (e.g., coefficients of a regresion)?
- Why can **Statistics change** (sampling distribution)?
- How to measure the spread of the change of statistics with the **Standard Error**
- **Standard Deviation vs Standard Error**.
- *Significance (*P-value*) of a $\beta$ coefficient.
-  Coefficient of Determination ($R^2$)

## What are **Statistics**?

**Statistics  are numerical values calculated from a sample.**

Examples include: mean, standard deviation, or the coefficients from a regression.

## The Statistic $\beta_1$ for a Sample of N=100 for a Univariate Regression.

```{=tex}
\begin{equation} 
\beta_{1,opt}= \frac
{{100} \sum_{i=1}^{100} y_i x_i- \sum_{i=1}^{100} y_i \sum_{i=1}^{100} x_i} 
{{100} \sum_{i=1}^{100} x_i^2 - \left (\sum_{i=1}^{100} x_i \right )^2}
\end{equation}
```

$\beta_1$ is clearly a **Statistic**!

## Example 1: {.smaller}
### Statistic $\beta_1$ for a Sample of N=100 for a Univariate Regression set.seed(123) 

```{r}
#| code-fold: true
library(rio)
library(janitor)
library(tidyverse)
DataSoldiersM=import("https://econ.lange-analytics.com/RData/Datasets/DataUSArmyBodyMeasures.xlsx", sheet="DataMale")|> 
              clean_names("upper_camel") |> 
              select(Height=Heightin, Weight=Weightlbs, Gender)
set.seed(123)

DataTrain=sample_n(DataSoldiersM,100)

Results=lm(Weight~Height, data=DataTrain) |> 
        summary()
print(Results)
```

$$Weight=\beta_1 Height +\beta_2$$

$$\beta_1=`r Results$coefficients[2,1]`$$
$$std.error=`r Results$coefficients[2,2]`$$ 

$SE_{\beta_1}$ measures the spread of the coefficient $\beta_1$.

How can $\beta_1$ have spread, as we have only one sample???

## Example 2: {.smaller}
### Statistic $\beta_1$ for a Sample of N=100 for a Univariate Regression set.seed(321) 

```{r}
#| code-fold: true
library(rio)
library(janitor)
library(tidyverse)
DataSoldiersM=import("https://econ.lange-analytics.com/RData/Datasets/DataUSArmyBodyMeasures.xlsx", sheet="DataMale")|> 
              clean_names("upper_camel") |> 
              select(Height=Heightin, Weight=Weightlbs, Gender)
set.seed(321)

DataTrain=sample_n(DataSoldiersM,100)

Results=lm(Weight~Height, data=DataTrain) |> 
        summary()
print(Results)
```

$$Weight=\beta_1 Height +\beta_2$$

$$\beta_1=`r Results$coefficients[2,1]`$$
$$std.error=`r Results$coefficients[2,2]`$$ 

$SE_{\beta_1}$ measures the spread of $\beta_1$. 




## How Can R Estimate the SE for $\beta_1$ Based on only One Sample?{.smaller}

The estimated *Standard Error* for $\beta_1$ is related to the *MSE* of the regression and the *Variance* of the *predictor(independend) variable* $(X)$. The formula for a uni-variate regression is:

$$
SE_{\beta_1}=\sqrt{\frac{MSE}{VAR_X}\frac{1}{N}}\cdot \underbrace{\sqrt{\frac{N-2}{N-1}}}_{\approx 1}
$$

- with: $MSE=\sum_{i=1}^N (\widehat Y_i -Y_i)^2$

- with: $VAR_X= \sum_{i=1}^N (X-\bar X)^2$ 

- with (in our case): $Y=Weight$

- with (in our case): $X=Height$

- with (in our case): $N=100$

The greater $N$ (the sample size)  the smaller is $SE$ (try N=400 instead og N=100) in the simulation and the $SE$ will be cut in half.

## Using the Formulas from the previous slide {.smaller}

```{r}
print(Results)
```

Note, R uses the $t-distribution$ here, instead of the standard normal distribution, because the standard error was not known but estimated.  For $N>100$ the two distributions are almost exactly the same.

**t-distribution has slightly fatter tails than the normal distribution to account for the estimation of the standard error**

## Finding the Standard Error with Repeated Sampling
### In reality this is not possible because we have only on sample!


Goto [https://econ.lange-analytics.com/ibm6510-40/SamplingDistributionLM.R](https://econ.lange-analytics.com/ibm6510-40/SamplingDistributionLM.R)

**Assumption:**
$\beta_1$ is normal distributed with repeated sampling.

**Central Limit Theorem:**

If a sufficiently large samples from a population is taken, the sample's statistics will be normally distributed, even if the population is not normally distributed.


In theory *sufficiently large* means infinity. However, in most cases samples with 30 or more observations are *sufficiently large*.







## Coefficient of Determination {.smaller}
### Left: SST (sum of squares, total) --------- Right: SSR (sum of squares, residuals)

![Source: [https://en.wikipedia.org/wiki/Coefficient_of_determination](https://en.wikipedia.org/wiki/Coefficient_of_determination)](CoefficientOfDetermination.png)

## Coefficient of Determination - Definition
[https://econ.lange-analytics.com/calcat/linregrmeans/](https://econ.lange-analytics.com/calcat/linregrmeans/)

$$
R^2=1 - \frac{SSR}{SST}=\frac{SST-SSR}{SST}=\frac{
\overbrace{\frac{SST-SSR}{N-1}}^{\mbox {Expl. Variance}}
}
{
\underbrace{\frac{SST}{N-1}}_{\mbox{Variance}}
}
$$

. . .

$$R^2=
\frac{\mbox {Expl. Variance}}{\mbox {Variance}}
$$


**$R^2$ is the proportion of explained variance of the depended variable. $R^2$ takes values from 0 (worst) to 1 (best)**
