---
title: "Logistic Regression"
subtitle: "Working with Imbalanced Datasets"
author:
  name: "Carsten Lange"
  email: "clange@cpp.edu"
  affiliation: "Cal Poly, Pomona"

format: 
  live-revealjs:
    theme: [moon, ../CustomCL.scss]
    smaller: true
    controls: true
    chalkboard:
      theme: whiteboard
      boardmarker-width: 2.7
    incremental: false
    scrollable: true
    footer: "<a href='https://ai.lange-analytics.com/htmlbook/TreeBased.html#TreeBased-RandForest' target='_blank'>Textbook </a> "
   
webr:
  packages: ['dplyr', 'rsample', 'janitor',  'recipes', 'broom', 'yardstick', 'parsnip', 'workflows',  'themis'] 
  

engine: knitr
---

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}

## What Will You Learn {.smaller}

```{webr setup}
#| context: setup
#| echo: false
DataChurn=read.csv("https://ai.lange-analytics.com/data/TelcoData.csv") |>
clean_names("upper_camel") |> 
          select(Churn,Gender,SeniorYes=SeniorCitizen,
                 Tenure, MonthlyCharges) |>
          mutate(ChurnYes=ifelse(Churn=="Yes",1,0), Churn=NULL) |>
          mutate(MaleYes=ifelse(Gender=="Male",1,0), Gender=NULL) |>
          mutate(ChurnYes=as.factor(ChurnYes)) |> 
          na.omit(TotalCharges)
set.seed(123)
Split7030=initial_split(DataChurn, prop=0.7,strata=ChurnYes)
DataTrain=training(Split7030)
DataTest=testing(Split7030)

RecipeChurn=recipe(ChurnYes~MaleYes+SeniorYes+MonthlyCharges+Tenure, data=DataTrain)

ModelDesignLogistic=logistic_reg() |>
                    set_engine("glm") |> 
                    set_mode("classification")

WFModelChurn=workflow() |>
             add_recipe(RecipeChurn) |>
             add_model(ModelDesignLogistic) |>
             fit(DataTrain)

DataTestWithPred=augment(WFModelChurn, new_data = DataTest)
```

-   What distinguishes data with a binary dependent variable (dummy variables) from data with a continuous dependent variable.

-   Why a linear regression (Linear Probability Model; LPM) is not suitable for classification (estimating dummy variables)

-   Why a Logistic Regression function is well suited for classification (estimating dummy variables)

-   How to interpret the parameters of a Logistic Regression

-   How to deal with unbalanced datasets
    - downsampling
    - upsampling
    - SMOTE
    

## A Mock Up Example {.smaller}

```{r rSetUpYachts}
#| echo: false
library(rio)
library(janitor)
library(tidymodels)
library(kableExtra)
DataPlot=tibble(Name=c("Jack","Sarah","Carl","Eric","Zoe",
                          "James","Enrico","Erica","Stephanie","Susan"),
                   Income=c(45,50,55,60,67,250,280,320,370,500),
                   Yacht=c(1,0,0,0,0,1,1,1,1,1))
kbl(DataPlot)
```

##  {.smaller background-iframe="loglinappdata.html" background-interactive="true"}

::: columns
::: {.column width="80%"}
:::

::: {.column width="20%"}
-   Click on legend to show/hide graphs.

-   The data are already shown.

-   Activate the OLS regression line to see how data are approximated.
:::
:::

##  {.smaller background-iframe="logisticfuncapp.html" background-interactive="true"}

::: columns
::: {.column width="75%"}
:::

::: {.column width="25%"}
$$P=\frac{1}{1+e^{-(\beta_1* Inc + \beta_2)}}$$

-   Adjust beta1 and beta2 to see the effect on the Logistic Function
:::
:::

##  {.smaller background-iframe="loglinapp0.html" background-interactive="true"}

::: columns
::: {.column width="70%"}
:::

::: {.column width="30%"}
$$P=\frac{1}{1+e^{-(\beta_1* Inc + \beta_2)}}$$

-   Click on legend to show/hide graphs.

-   50%-Level and Decision Boundary graphs are currently hidden

-   Adjust beta1 and beta2 to improve MSE of Logistic Regression
:::
:::

##  {.smaller background-iframe="loglinappopt.html" background-interactive="true"}

::: columns
::: {.column width="70%"}
:::

::: {.column width="30%"}
$$P=\frac{1}{1+e^{-(\beta_1* Inc + \beta_2)}}$$

-   Click on legend to show/hide graphs.

-   50%-Level and Decision Boundary graphs are currently hidden

-   Use 50%-Level and Decision Boundary to see predicted yacht ownership fot data
:::
:::

## Churn at the Telco Company

**Research Question:**

From a marketing standpoint, should we pay more attention to women vs. men, seniors vs. non seniors, customers with high monthly charges or high tenure to more effectively prevent churn?

<!-- ## Churn at the Telco Company --- Digital Acylical Graph -->


## Churn at the Telco Company

```{r rSetupChurn}
#| code-line-numbers: "1-4|5|6"
#| echo: true
library(rio)
library(janitor)
library(tidymodels)
library(SmartEDA)
DataChurnOrg=import("https://ai.lange-analytics.com/data/TelcoData.csv") |>
             clean_names("upper_camel") 
```

```{r}
#| echo: true
# ExpReport(DataChurnOrg, op_file="EDAChurnData.html")
```

[EDAChurnData.html](EDAChurnData.html){target="_blank"}

## Data Engineering

```{r}
#| code-line-numbers: "1|2-3|4|5|6"
#| echo: true
DataChurn=DataChurnOrg |>
          select(Churn,Gender,SeniorYes=SeniorCitizen,
                 Tenure,MonthlyCharges, TotalCharges) |>
          mutate(ChurnYes=ifelse(Churn=="Yes",1,0), Churn=NULL) |>
          mutate(MaleYes=ifelse(Gender=="Male",1,0), Gender=NULL) |>
          na.omit(TotalCharges)
```

. . .

```{r}
#| echo: true
#| code-line-numbers: "1|2"
library(corrplot)
corrplot(cor(DataChurn))
```


## Splitting into Training and Testing Data

```{r SplitForR}
DataChurn=DataChurn |> 
          mutate(ChurnYes=as.factor(ChurnYes)) 
set.seed(123)
Split7030=initial_split(DataChurn, prop=0.7,strata=ChurnYes)
DataTrain=training(Split7030)
DataTest=testing(Split7030)
```

```{webr SplitForWebR}
#| exercise: exc1
set.seed(789)
Split7030=initial_split(DataChurn, prop=0.7,strata=ChurnYes)
DataTrain=training(______)
DataTest=testing(______)

print(Split7030)
head(DataTrain)
```

::: {.solution exercise="exc1"}
::: { .callout-tip collapse="false"}

## Fully worked solution:

```r
set.seed(789)
Split7030=initial_split(DataChurn, prop=0.7,strata=ChurnYes)
DataTrain=training(Split7030)
DataTest=testing(Split7030)

print(Split7030)
head(DataTrain)
```
:::
:::

## Analysis --- Creating Fitted Workflow

```{webr RecipeAndModelOrg}
#| exercise: exc2
RecipeChurn=______(______~MaleYes+SeniorYes+MonthlyCharges+Tenure, data=______)

ModelDesignLogistic=logistic_reg() |>
                    set_engine("glm") |> 
                    set_mode("______")

WFModelChurn=workflow() |>
             add_recipe(______) |>
             add_model(______) |>
             fit(______)

#Printing Parameters
tidy(WFModelChurn)
```

::: {.solution exercise="exc2"}
::: { .callout-tip collapse="false"}

## Fully worked solution:

```r
RecipeChurn=recipe(ChurnYes~MaleYes+SeniorYes+MonthlyCharges+Tenure, data=DataTrain)

ModelDesignLogistic=logistic_reg() |>
                    set_engine("glm") |> 
                    set_mode("classification")

WFModelChurn=workflow() |>
             add_recipe(RecipeChurn) |>
             add_model(ModelDesignLogistic) |>
             fit(DataTrain)

#Printing Parameters
tidy(WFModelChurn)
```
:::
:::

## Preparing Result Analysis


```{webr}
#| exercise: exc3
DataTestWithPred=______(______, new_data = DataTest)
head(DataTestWithPred)
```

::: {.solution exercise="exc3"}
::: { .callout-tip collapse="false"}

## Fully worked solution:

```r
DataTestWithPred=augment(WFModelChurn , new_data = DataTest)
head(DataTestWithPred)
```
:::
:::

## Analysis --- Summary Result

**Confusion Matrix:**

```{webr ConfusionMatrixOrg}
#| exercise: exc4
conf_mat(DataTestWithPred, truth=______, estimate= .______)
```

::: {.solution exercise="exc4"}
::: { .callout-tip collapse="false"}

## Fully worked solution:

```r
conf_mat(DataTestWithPred, truth=ChurnYes, estimate= .pred_class)
```
:::
:::

. . .

**Selected Metrics:**

```{webr}
#| exercise: exc5
FctMetricsCL=metric_set(accuracy, sensitivity, specificity)
FctMetricsCL(______, truth=______, estimate=______)
```

::: {.solution exercise="exc5"}
::: { .callout-tip collapse="false"}

## Fully worked solution:

```r
FctMetricsCL=metric_set(accuracy, sensitivity, specificity)
FctMetricsCL(DataTestWithPred, truth=ChurnYes, estimate= .pred_class)
```
:::
:::

## **Warning: Be careful with the Accuracy Rate** {.smaller}

#### The Story of Dr. Nebulous's Gamblers System

Dr. Nebulous offers a **97% Machine Learning Gambling Prediction**. Here is how it works: Gamblers can buy a prediction for a fee of \$5. Dr. Nebulous will then run his famous machine learning model and send a closed envelope with the prediction. The gambler is supposed to open the envelope in the casino, right before placing a bet of \$100 on a number in roulette. The envelope contains a message that states either "You will win" or "You will lose", which allows the gambler to act accordingly by either bet or not bet.

Dr. Nebulous claims that a "clinical trial" of 1000 volunteers, who opened the envelope after they had bet on a number in roulette, shows an accuracy of 97.3%.

**How could Dr. Nebulous have such a precise model?**

## **Warning: Be careful with the Accuracy Rate** {.smaller}

#### The Story of Dr. Nebulous's Gamblers System

The trick is Dr. Nebulous's machine learning model uses the *naive prognosis*: It always predicts "You will lose".

Here is the confusion matrix from the 1,000 volunteers trial:

```{r echo=FALSE}
# needed for Dr. Nebulous

# Define your expected classes
lvls <- c("Class1", "Class2")

# Create a zero-filled table with those levels
empty_table <- table(
  Prediction = factor(levels = lvls),
  Truth = factor(levels = lvls)
)

# Convert to a tidymodels conf_mat object
ConMatNebulous <- conf_mat(empty_table)

rownames(ConMatNebulous$table)=c("Win","Lose")
colnames(ConMatNebulous$table)=c("Win","Lose")
ConMatNebulous$table[1]=0
ConMatNebulous$table[2]=27
ConMatNebulous$table[3]=0
ConMatNebulous$table[4]=997
print(ConMatNebulous)
```

Roulette has 37 numbers to bet on. Chance to win is: $\frac{1}{37}=0.027$.

Out of the 1000 volunteers, 27 are expected to win, and 973 are expected to lose.

$$Accuracy=\frac{0+973}{1000}=0.973$$

## **Warning: Be careful with the Accuracy Rate** {.smaller}

#### The Story of Dr. Nebulous's Gamblers System

```{r echo=FALSE}
rownames(ConMatNebulous$table)=c("Win","Lose")
colnames(ConMatNebulous$table)=c("Win","Lose")
ConMatNebulous$table[1]=0
ConMatNebulous$table[2]=27
ConMatNebulous$table[3]=0
ConMatNebulous$table[4]=997
print(ConMatNebulous)
```

However, when we look at the correct positive and the correct negative rate separately, we see that Dr. Nebulous' accuracy rate (although correct) makes little sense.

-   The correct negative rate (**specificity**) is 100%

-   The correct positive rate (**sensitivity**) is zero (out of the 27 winners, all were falsely predicted as "You will lose").



**This example shows: When interpreting the confusion matrix, you must look at accuracy, sensitivity, and specificity simultaneously**

## Fixing the Problem of Unbalanced Data {.smaller}
### The Problem

```{r}
count(DataTrain, ChurnYes)
```

::: {.callout-note}
## Imbalanced Response Variable
We had **3614 customers not-churning** and only **1308 customers churning**
:::

::: incremental
1.  `step_downsample()`: Delete observations from the majority class. \
    With the argument `underatio=` you can stop *downsampling* when the majority class is reduced to a defined multiple of the minority class. E.g., `step_downsample(under_ratio=1.2)` means that *downsampling* stops when the majority class is 1.2 times the minority class (default is `under_ratio=1`)

2.  `step_upsample()`: Copy/Paste observations from the minority class. \
    With the argument `over_ratio=` you can stop *upsampling* when the minority class grows to a defined fraction of the majority class (default is `over_ratio=1`). E.g., `step_upsample(over_ratio=0.9)` *upsampling* stops when the minority class reaches 0.9 (90%) of the majority class:

3.  `step_smote()`: Create records similar to existing minority class observations. \
    *SMOTE* uses *k-Nearest-Neighbor* to find the most similar observation to a randomly selected observation and th,e generates a mix out of these two observations until the chosen `over_ratio=` ratio is reached (default is `over_ratio=1`)
:::

## Fixing the Problem of Unbalanced Data {.smaller}
### Up-, Downsampling and SMOTE in `tidymodels` Workflow

To implement `step_upsample()`, `step_downsample()`, or  `step_smote()` only the `recipe()` needs to be changed and updated in the workflow.

::: {.callout-note}
## `library(themis)`

**Note**, `step_upsample()`, `step_downsample()`, and `step_smote()` are not included in `tidymodels`. They are part of the `themis` package, which needs to be installed ones and loaded (`library(themis)`). 
:::

## Using Downsampling to Balance the Training Data

`library(themis)` is needed for `step_downsample()` and has been loaded already in the background.

```{webr}
RecipeChurnDown=recipe(ChurnYes~MaleYes+SeniorYes+MonthlyCharges+Tenure, data=DataTrain) |>
            step_downsample(ChurnYes, under_ratio=1.2)

# Showing count of ChurnYes
count(RecipeChurnDown |>prep() |>juice(),ChurnYes)  
```

. . .

```{webr}
WFModelChurnDown=workflow() |>
             add_recipe(RecipeChurnDown) |>
             add_model(ModelDesignLogistic) |>
             fit(DataTrain)

tidy(WFModelChurnDown)
```

## Analyzing the Results for Downsampling

```{webr}
DataTestWithPredDown=augment(WFModelChurnDown, new_data = DataTest)

conf_mat(DataTestWithPredDown, truth=ChurnYes, estimate= .pred_class)
```

. . .

```{webr}
FctMetricsCL=metric_set(accuracy, sensitivity, specificity)
FctMetricsCL(DataTestWithPredDown, truth=ChurnYes, estimate= .pred_class)
```

## Using SMOTE to Balance the Training Data

`library(themis)` is needed for `step_smote()` and has been loaded already in the background.

```{webr}
RecipeChurnSmote=recipe(ChurnYes~MaleYes+SeniorYes+MonthlyCharges+Tenure, 
                        data=DataTrain) |>
            step_smote(ChurnYes, over_ratio=0.8)

# Showing count of ChurnYes
count(RecipeChurnSmote|>prep()|>juice(),ChurnYes)
```

. . .

```{webr}
WFModelChurnSmote=workflow() |>
             add_recipe(RecipeChurnSmote) |>
             add_model(ModelDesignLogistic) |>
             fit(DataTrain)
tidy(WFModelChurnSmote)
```

## Analyzing the Results for SMOTE

```{webr}
DataTestWithPredSmote=augment(WFModelChurnSmote, new_data = DataTest)

conf_mat(DataTestWithPredSmote, truth=ChurnYes, estimate= .pred_class)
```

. . .

```{webr}
FctMetricsCL=metric_set(accuracy, sensitivity, specificity)
FctMetricsCL(DataTestWithPredSmote, truth=ChurnYes, estimate= .pred_class)
```