---
title: "Logistic Regression"
subtitle: "A Powerful Tool for Classification"
format: 
  revealjs:
    code-fold: true
---


## Basics of Logistic Regression

Classification Algorithm:

- **One outcome variable** (analysed in what follows)
- Multiple outcome variables
  - ordered logistic regression (not covered here) 
  - unordered logistic regression (not covered here) »

## A Mock-Up Example to Introduce the Idea {.smaller}

```{r}
#| echo: true
library(tidymodels); library(kableExtra); library(janitor); library(rio)
DataYachts=import("https://lange-analytics.com/AIBook/Data/DataYachts.csv")%>% 
  mutate(YachtNum=Yacht, Yacht=as.factor(Yacht)) 
kbl(DataYachts %>% select(-YachtNum), caption="Income and Yacht Ownership") %>% 
  kable_styling(bootstrap_options=c("striped","hover"), position="center", full_width = F)
```

## Using OLS is a Tempting (but bad) Idea

```{r}
#| echo: true
library(plotly)
ggplotly(ggplot(aes(x=Income,y=YachtNum),data=DataYachts)+
  geom_point(size=2.7, color="magenta")+
  scale_x_continuous(limits = c(-50,500), breaks = seq(0,500,50))+ 
  scale_y_continuous(breaks = seq(0,1.25,0.25))+ 
  labs(y="Probability of Yacht Ownership", x="Income in $1,000"))
```
## Using OLS is a Tempting (but bad) Idea

```{r}
#| echo: true
library(plotly)
ggplotly(ggplot(aes(x=Income,y=YachtNum),data=DataYachts)+
  geom_hline(yintercept = 0.5)+
  geom_point(size=2.7, color=ifelse(DataYachts$Income==45,"red","cyan"))+
  geom_smooth(method="lm",se=FALSE, size=1.7)+
  scale_x_continuous(limits = c(-50,500), breaks = seq(0,500,50))+ 
  scale_y_continuous(breaks = seq(0,1.25,0.25))+ 
  labs(y="Probability of Yacht Ownership", x="Income in $1,000"))
```

## Quick Way to Find a Decision Boundary {.smaller}

::: columns
::: {.column width="50%"}
```{r}
ggplot(aes(x=Income,y=YachtNum),data=DataYachts)+
  geom_hline(yintercept = 0.5)+
  geom_vline(xintercept = 158, linetype="dashed", color="blue")+
  geom_point(size=2.7, color=ifelse(DataYachts$Income==45,"red","cyan"))+
  geom_smooth(method="lm",se=FALSE, size=1.7)+
  scale_x_continuous(limits = c(-50,500), breaks = seq(0,500,50))+ 
  scale_y_continuous(breaks = seq(0,1.25,0.25))+ 
  labs(y="Probability of Yacht Ownership", x="Income in $1,000")
```
:::

::: {.column width="50%"}
1.  Find the intersection point between the prediction line and the horizontal 0.5 probability line.

2.  Draw a vertical line through the intersection point. This line is called a **decision boundary**.

3.  All incomes left of the *decision boundary* (income smaller than 158) are predicted as "no". All incomes right of the *decision boundary* (income greater than 158) are predicted as "yes".»
:::
:::

## Why OLS for Classification is a Bad Idea {.smaller}

```{r}
library(plotly)
ggplotly(ggplot(aes(x=Income,y=YachtNum),data=DataYachts)+
  geom_hline(yintercept = 0.5)+
  geom_vline(xintercept = 158, linetype="dashed", color="blue")+
  geom_point(size=2.7, color=ifelse(DataYachts$Income==45,"red","cyan"))+
  geom_smooth(method="lm",se=FALSE, size=1.7)+
  scale_x_continuous(limits = c(-50,500), breaks = seq(0,500,50))+ 
  scale_y_continuous(breaks = seq(0,1.25,0.25))+ 
  labs(y="Probability of Yacht Ownership", x="Income in $1,000"))
```

Note, incomes \> \$370,000 are predicted with a probability \> 100% to be yacht owners(?)<br> 
E.g. probability of owning a yacht for an income of \$500,000 is 125% (?)

A similar problem can occur with negative probabilities!

## A Step-Function as an Alternative to OLS

```{r}
ModelDesignLogRegr=logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

RecipeYachts=recipe(Yacht~Income, data = DataYachts)

set.seed(123)
WFYachts=workflow() %>% 
  add_model(ModelDesignLogRegr) %>% 
  add_recipe(RecipeYachts) %>% 
  fit(DataYachts)

b1=tidy(WFYachts)[[2,2]]
b2=tidy(WFYachts)[[1,2]]
ggplot(aes(x=Income,y=YachtNum),data=DataYachts)+
  geom_hline(yintercept = 0.5)+
  geom_vline(xintercept = 158, linetype="dashed", color="blue")+
  geom_vline(xintercept = 110, linetype="dashed", color="magenta")+
  geom_point(size=2.7, color=ifelse(DataYachts$Income==45,"red","cyan"))+
  geom_function(fun=function(x) 1/(1+exp(-b1*x-b2)), color="magenta", size=1.7)+
  geom_smooth(method="lm",se=FALSE, size=1.7)+
  scale_x_continuous(limits = c(-50,500), breaks = seq(0,500,50))+
  scale_y_continuous(breaks = seq(0,1.25,0.25))+
  labs(y="Probability of Yacht Ownership", x="Income in $1,000")
```

## Popular Step-Functions (Sigmoid Function)


- The **Hyperbolic Tangent function**.

- The **Arc Tangent function**.

- The **Logistic function** (confusingly sometimes also called the *sigmoid function*). 

. . .

$$
y_i=\frac{1}{1+e^{-x_i}}
$$ »

## The Logistic Function

::: nonincremental
- The **Logistic function** (confusingly sometimes also called the *sigmoid function*):

:::

$$
y_i=\frac{1}{1+e^{-x_i}}
$$

. . .


We use: $y_i=P^{rob}_{yes,i}$ and $x_i=\beta_1 Inc_i+\beta_2$ which gives us:

. . .

$$
P^{rob}_{yes,i}=\frac{1}{1+e^{-(\beta_1 Inc_i+\beta_2)}}
$$
**$\beta_1$ and $\beta_2$ change slope and position <br />
$\beta_1=1$ and $\beta_2=0$ gives the org. logistic function.**
🤓

::: footer
See: [Try it yourself](https://econ.lange-analytics.com/calcat/logisticregr)
:::

## What Makes the Logistic Function so Special? {.scrollable}
### --- compared to other sigmoid (step) functions ---

**Time for some mathematical magic:**

Logistic function $P^{rob}_{yes,i}$ := probability for  positive event (e.g.  yacht ownership: yes):

$$P^{rob}_{yes,i}=\frac{1}{1+e^{-(\beta_1\cdot x_i+\beta_2)}} $$

. . . 



Take the inverse on both sides of the equation:

$$\frac{1}{P^{rob}_{yes,i}}=1+e^{-(\beta_1\cdot x_i+\beta_2)}$$

. . .

Subtract 1 on both sides: 

$$\frac{1}{P^{rob}_{yes,i}}-1=e^{-(\beta_1\cdot x_i+\beta_2)}$$
 
. . .

Consider that $-1=-\frac{P^{rob}_{yes,i}}{P^{rob}_{yes,i}}$ and substitute $-1$  accordingly, we get after simplification:

$$\frac{1-P^{rob}_{yes,i}}{P^{rob}_{yes,i}}=e^{-(\beta_1\cdot x_i+\beta_2)}$$

. . .

$1-P^{rob}_{yes,i}$ equals by definition $P^{rob}_{no,i}$:

$$\frac{P^{rob}_{no,i}}{P^{rob}_{yes,i}}=e^{-(\beta_1\cdot x_i+\beta_2)}$$

. . .

Take again the inverse on both sides:

$$\frac{P^{rob}_{yes,i}}{P^{rob}_{no,i}}=e^{\beta_1\cdot x_i+\beta_2}$$

. . .

Take the logarithm on both sides:

$$\ln\left (\frac{P^{rob}_{yes,i}}{P^{rob}_{no,i}}\right )=\beta_1\cdot x_i+\beta_2$$


## One More Step --- Odds vs Probabilties 

- The fraction of the yes/no probabilities can be interpreted as 
  $Odds$ as they are often used in betting. 

- Example: The probability of getting two heads when flipping two coins is  is $P^{rob}_{yes,i}=0.25$. 

- Consequently, the probability of **not** getting two heads when flipping two coins is  $P^{rob}_{no,i}=0.75$.

- $Odds$ for 2 Heads compared to **not** 2 heads is 1 to 3 or 33%: 

. . .

$$O^{dds}=\frac{P^{rob}_{yes,i}}{P^{rob}_{no,i}}=\frac{0.25}{0.75}=\frac{1}{3}=0.33$$




## Interpretation of the $\beta s$: Yacht Ownership

$$\ln(O^{dds})=\ln\left (\frac{P^{rob}_{yes,i}}{P^{rob}_{no,i}}\right )=0.02\cdot Inc_i+(-2.7)$$

Model results after running and printing the workflow():

```{r}
#| echo: true

# The workflow WFYachts had been created before and
# DataTest were fitted to the workflow.
print(WFYachts)
```

## Interpretation of the $\beta s$: Yacht Ownership

$$\ln(O^{dds})=\ln\left (\frac{P^{rob}_{yes,i}}{P^{rob}_{no,i}}\right )=0.02\cdot Inc_i+(-2.7)$$

- If income increases by 1 ($1,000) the logarithm of the odds increases by 0.02.

- Since change of a logarithm is a relative change (**percentage**):

. . .

  If income increases by 1 ($1,000) the odds increases by 2% (0.02). (careful with the results because data were made up and N is too small!)


## Confusion Matrix

Note, in the mockup we did not create training and testing data. Therefore, we use DataYachts (the data we used to fit/train the workflow) here. This is not a proper methodology but good enough for the mock-up: 
```{r}
#| echo: true
DataYachtsWithPred=augment(WFYachts, new_data=DataYachts)
conf_mat(DataYachtsWithPred, truth=Yacht, estimate=.pred_class)
```

## Real World Churn Analysis with Logistic Regression --- the Data  


We use data (7,043 customers) of the  fictional telecommunication  company *TELCO*, generated by *IBM* for training purposes:

- The outcome variable $Churn$ indicates, if a customer departed within the last month ($Churn=Yes$) or not ($Churn=No$). 
- Predictor variables contain:  
  - Customers' $Gender$ ($Female$ or $Male$), 
  - Customers' $SeniorCitizen$ status ($0$ for no or $1$ for yes), 
  - Customers' $Tenure$ with *TELCO* (month of membership), 
  - Customers' $MonthlyCharges$ (in US-\$), as well as
  - Customers' $TotalCharges$ (in US-\$). 

## Real World Churn Analysis with Logistic Regression --- the Data

```{r}
#| echo: true
DataChurn=import("https://lange-analytics.com/AIBook/Data/TelcoChurnData.csv") %>%
  clean_names("upper_camel") %>%
  select(Churn,Gender,SeniorCitizen,Tenure,MonthlyCharges) %>%
  mutate(Churn=factor(Churn, levels=c("Yes","No"))) 
head(DataChurn)
```


## Real World Churn Analysis with Logistic Regression
### --- Do it yourself ---

Create the Churn analysis with logistic regression. Click on the link in the footer to get an R-script with a skeleton for the analysis.🤓

::: footer
See: [LogRegrChurnScript.R](https://lange-analytics.com/AIBook/Scripts/LogRegrChurnScript.R)
:::


## Results from Churn Analysis with Logistic Regression{.smaller}

```{r}
set.seed(789)
Split3070=initial_split(DataChurn, prop=0.7,strata =Churn)

DataTrain=training(Split3070)
DataTest=testing(Split3070)

set.seed(789)
RecipeChurn=recipe(Churn ~ ., data=DataTrain) %>% 
  step_naomit() %>% 
  step_dummy(Gender)

ModelDesignLogisticRegr= logistic_reg() %>% 
  set_engine("glm") %>% 
  set_mode("classification")

WFModelChurn=workflow() %>% 
  add_recipe(RecipeChurn) %>% 
  add_model(ModelDesignLogisticRegr) %>% 
  fit(DataTrain)

DataTestWithPred=augment(WFModelChurn, new_data = DataTest)
```

Confusion Matrix:

```{r}
kbl(conf_mat(DataTestWithPred, truth = Churn, estimate = .pred_class)$table)
```

Accuracy:

```{r}
kbl(accuracy(DataTestWithPred, truth = Churn, estimate = .pred_class))
```

Sensitivity:

```{r}
kbl(sensitivity(DataTestWithPred, truth = Churn, estimate = .pred_class))
```

Specificity:

```{r}
kbl(specificity(DataTestWithPred, truth = Churn, estimate = .pred_class))
```

**Hint:** What do the column sums of the confusion matrix tell you?

## Problem: Unballanced Training Data


```{r}
kbl(count(DataTrain, Churn))
```

**Majority Class:** $Churn=No$ has 3621  observations in the training dataset.

**Minority class** $Churn=Yes$ has 1308 observations in the training dataset. 

## What can we do?

```{r}
kbl(count(DataTrain, Churn))
```

- **Downsampling:** Randomly **delete observations from majority class** until ratio of the observations from the majority and the minority class reaches the desired ratio (e.g., 1:1).

- **Upsampling:** In simplest version, **creates new observations for the minority class** by copying randomly chosen observations from the minority class until the ratio of the observations from the majority and the minority class reaches the desired ratio (e.g., 1:1).

- Often, a combination of *downsampling* and *upsampling* is performed.

## Performing Down-Sampling with `step_upsample()`

You need to add the R package `themis`. Then in your script, you can add `step_downsample(Churn)` to the recipe (don't forget to execute the following command lines again). As a reminder our original DataTrain had 4,929 observations, $Churn_{Yes}=1308$, $Churn_{No}=3621$:

```{r}
#| echo: true
library(themis)
RecipeChurn=recipe(Churn ~ ., data=DataTrain) %>% 
  step_naomit() %>% 
  step_dummy(Gender) %>% 
  step_downsample(Churn)

 # you do not need to do the following steps.
# They just allow to display the count() for the training data
ExtractedDataTrain=juice(RecipeChurn %>% prep())
kbl(count(ExtractedDataTrain, Churn))
```

Note, the number of observations has decreased by 2313. This is an information loss!

## Performing Up-Sampling with `step_upsample()`

You need to add the R package `themis`. Then in your script, you can add `step_upsample(Churn)` to the recipe (don't forget to execute the following command lines again). As a reminder our original DataTrain had 4,929 observations, $Churn_{Yes}=1308$, $Churn_{No}=3621$:

```{r}
#| echo: true
library(themis)
RecipeChurn=recipe(Churn ~ ., data=DataTrain) %>% 
  step_naomit() %>% 
  step_dummy(Gender) %>% 
  step_upsample(Churn)

 # you do not need to do the following steps.
# They just allow to display the count() for the training data
ExtractedDataTrain=juice(RecipeChurn %>% prep())
kbl(count(ExtractedDataTrain, Churn))
```

Note, the number of observations has increased by 2313. The information in the dataset has not increased!

## Performing Up-Sampling with `step_smote()`. What is the advantage

As a reminder our original DataTrain had 4,929 observations, $Churn_{Yes}=1308$, $Churn_{No}=3621$:

```{r}
#| echo: true
library(themis)
RecipeChurn=recipe(Churn ~ ., data=DataTrain) %>% 
  step_naomit() %>% 
  step_dummy(Gender) %>% 
  step_upsample(Churn)

 # you do not need to do the following steps.
# They just allow to display the count() for the training data
ExtractedDataTrain=juice(RecipeChurn %>% prep())
kbl(count(ExtractedDataTrain, Churn))
```

Instead of copying a record from the training dataset, `step_smote()` finds the Nearest Neighbor to that record and creates a new record that has features generated as a weighted average between the Nearest Neighbor and the original record.