---
title: "Linear Regression"
subtitle: "Review"
format: 
  revealjs:
    code-fold: true
    scrollable: true
    echo: true
---


## What Will You Learn/Review {.scrollable .smaller}

-   Reviewing the basic idea behind linear regression

-   Learning how how to measure predictive quality with Mean Square Error ($MSE$).

-   Understanding the role of in a linear regression model (applies also to other models)

-   Calculating optimal regression parameters using OLS

-   Finding optimal regression parameters by trial and error

-   Finding optimal regression parameters with `lm` and `tidymodels`.

-   Distinguish between unfitted and fitted models

-   Using `tidymodels` to split observations randomly into a training and testing dataset.

-   Using `tidymodels` to analyze housing prices.

## Loading the Libraries and the Data

```{r}
library(tidymodels)
library(rio)
library(janitor)
Seed=123
set.seed(Seed)
DataHouses=import("../Data/DataHousingBig.rds") |> 
           clean_names("upper_camel") |> 
           select(Price,Sqft=SqftLiving,Bedrooms,Condition) |> 
           sample_n(144) |> 
           mutate(Price=228*Sqft+32543+rnorm(144,0,50000))
head(DataHouses)

```
## Splitting in Training and Testing Data:

```{r}
set.seed(Seed)
Split7030=initial_split(DataHouses,prop=0.7, strata = Price)
DataTrain=training(Split7030)
DataTest=testing(Split7030)
```
## How much is a House Worth in King County?

. . .

**A house with average properties should be predicted with an average price!**

```{r}
MeanSqft=mean(DataHouses$Sqft)
cat("The mean square footage of a house in King county is:", MeanSqft)

MeanPrice=mean(DataHouses$Price)
cat("The mean price of a house in King county is:", MeanPrice)
```

## Predicting the Price of an Average Sized House as the Average of all House Prices

```{r}
#| echo: false
ggplot(aes(x=Sqft, y=Price), data=DataHouses)+
  geom_point(color="red")+
  geom_hline(yintercept=mean(DataHouses$Price))+
  geom_vline(xintercept=mean(DataHouses$Sqft))
```
# How to Measure Prediction Quality with the Mean Squared Error (MSE) {.smaller}

```{=tex}
\begin{eqnarray*}
MSE & = & \frac{1}{N} \sum_{i=1}^{N}(\widehat{y}_{i}-y_{i})^{2} \\
 & \Longleftrightarrow& \\
MSE & =  & \frac{1}{N} \sum_{i=1}^{N}(\underbrace{\overbrace{\beta_{1}x_{i}+\beta_2}^{\mbox{Prediction $i$}}-y_i}_{\mbox{Error $i$}})^2
\end{eqnarray*}
```
**Note, when the data are given (i.e.,** $x_i$ and $y_i$ are given), the $MSE$ depends only on the choice of $\beta_1$ and $\beta_2$ »


## Including Sqft as Determiant of Price
### Preparing the Data

Blueprint for the data:
```{r}
#| code-fold: false
RecipeHouses=recipe(Price~Sqft, data=DataTrain)
```

## Choosing the Model Blueprint

Blueprint for the model:
```{r}
#| code-fold: false
ModelDesignOLS=linear_reg() |> 
            set_engine("lm") |> 
            set_mode("regression")
```

## How does the Unfitted Model Looks Like?

$$
\underbrace{Price}_y=\underbrace{\beta_1}_m \underbrace{Sqft}_x + \underbrace{\beta_0}_b
$$
## Using a Workflow to Fit the Model to the Data (finding the optimal $\beta_1$ and $\beta_2$ values

$$
\underbrace{Price}_y=\underbrace{\beta_1}_m \underbrace{Sqft}_x + \underbrace{\beta_0}_b
$$
```{r}
#| code-fold: false
WFModelHouses=workflow() |> 
              add_recipe(RecipeHouses) |> 
              add_model(ModelDesignOLS) |> 
              fit(DataTrain)
```

## Unfitted Model vs Fitted Workflow Model {.smaller}

Unfitted Model:
$$
\underbrace{Price}_y=\underbrace{\beta_1}_m \underbrace{Sqft}_x + \underbrace{\beta_0}_b
$$

```{r}
tidy(WFModelHouses)
```

Fitted Model:
$$
\underbrace{Price}_y=\underbrace{228}_m \cdot\underbrace{Sqft}_x + \underbrace{32390}_b
$$

Predict the price for a house with 1,000 sqft and send it to me in a private chat!

## How Does the fitted model that considers SQFT improves the prediction as a simple average

```{r}
#| echo: false
ggplot(aes(x=Sqft, y=Price), data=DataTrain)+
  geom_point(color="red")+
  geom_vline(xintercept=mean(DataHouses$Sqft))+
  geom_hline(yintercept=mean(DataHouses$Price))+
  geom_smooth(method="lm", se=FALSE, color="blue")
```

## Interpretation of the Results

## Unfitted Model vs Fitted Workflow Model

Unfitted Model:
$$
Price=\beta_1 Sqft + \beta_0
$$

```{r}
tidy(WFModelHouses)
```

Fitted Model:

$$
\begin{aligned}
Price&=&177 \cdot &Sqft + &131134\\
+177&= &177  \cdot &(+1)+ &(+0)
\end{aligned}
$$






