---
title: "Ggplot2"
subtitle: "Use the Grammar of Graphics to Visualize Data"
format: 
  revealjs:
    code-fold: true
    incremental: false
    scrollable: true
    echo: true
---

## What Will You Learn {.scrollable .smaller}




-   You will r

## Loading the Titanic Dataset and Performing Basic DataEngeneering {.smaller}

"The purchasing power of a British pound from the year 1912 was about 90.4 British pounds at the beginning of 2023 (see Bank of England (2022)). Multiplying this with the exchange rate for the British pound at the beginning of 2023 ($1.2/Brit. pound) (see Federal Reserve Bank of St. Louis (2023)) gives us the multiplier of 108.5" (Lange, C., Practical Machine Learning)
```{r}
#| output-location: fragment
library(rio)
library(tidyverse)
DataTitanic= import(
        "https://ai.lange-analytics.com/data/Titanic.csv") |> 
             select(Survived,Sex,Age,FareInPounds, Class=Pclass) |> 
             mutate(FareIn2023Dollars=108.5*FareInPounds) |> 
             mutate(FareInPounds=NULL) |> 
             mutate(Survived=as.factor(Survived))
head(DataTitanic)
```


## The Grammar of Graphics: maping variables as dimensions to be displayed)
### Example Age and Class

```{r}
#| output-location: fragment
ggplot(DataTitanic, mapping=aes(x=FareIn2023Dollars, y=Age))
```

## The Grammar of Graphics: adding a geom_ to a Graph
### Example Age and Fare

```{r}
#| output-location: fragment
ggplot(DataTitanic, mapping=aes(x=FareIn2023Dollars, y=Age))+
  geom_point(color="red", size=1)
```

## The Grammar of Graphics: adding a geom_ to a Graph
### Example Age,  Fare, and Survived (three dimensions)

```{r}
#| output-location: fragment
ggplot(DataTitanic, mapping=aes(x=FareIn2023Dollars, y=Age, color=Survived))+
  geom_point(size=1)
```

## The Grammar of Graphics: adding labels and changing scale
### Example Age,  Fare, and Survived (three dimensions)

```{r}
#| output-location: fragment
ggplot(DataTitanic, mapping=aes(x=FareIn2023Dollars, y=Age, color=Survived))+
  geom_point(size=1) +
  labs(x="Fare", y="Passenger Age", 
       title="Survival on the Titanic")+
  scale_x_continuous(breaks=seq(0,60000,5000))
```

## The Grammar of Graphics: adding labels and changing scale
### Example Age,  Class, and Survived (three dimensions)

```{r}
#| output-location: fragment
ggplot(DataTitanic, mapping=aes(x=Class, y=Age, color=Survived))+
  geom_point(size=1) +
  labs(x="Fare", y="Passenger Age", 
       title="Survival on the Titanic")
```

## The Grammar of Graphics: jitter plot
### Example Age,  Class, and Survived (three dimensions)

```{r}
#| output-location: fragment
ggplot(DataTitanic, mapping=aes(x=Class, y=Age, color=Survived))+
  geom_jitter(size=1) +
  labs(x="Fare", y="Passenger Age", 
       title="Survival on the Titanic")
```

## Time Series for Apple Stock

```{r}
#| output-location: fragment
DataStockTimeSeries=
  data.frame(EndOfYear=seq(2019,2023,1), 
             Price=c(71.25,129.90,174.91,128.72,191.8))
library(ggthemes)
ggplot(DataStockTimeSeries, mapping=aes(x=EndOfYear, y=Price))+
  geom_line(color="red", linewidth=2 )+
  labs(title="Apple Stock Price",
       subtitle="(end of year)",
       x="Year")+
  theme_economist()
```


## More about Ggplot2

Thomas Lin Pedersen

ggplot workshop part 1 (2,5 hours)

(https://www.youtube.com/watch?v=h29g21z0a68) 

