---
title: " Dunmies, Factors, ANOVA, and Linear Regression"
subtitle: "Review"
format: 
  revealjs:
    code-fold: true
    scrollable: true
editor: 
  markdown: 
    wrap: 72
---

## What Will You Learn 

-   Review: Dummy Variables in Linear Regression
-   One-Way ANOVA
-   AB Tests
-   Two-Way ANOVA

## Libraries and Data

```{r echo=TRUE}
library(rio)
library(tidyverse)
library(janitor)
library(wooldridge)
DataWage=wage1 |>
clean_names("upper_camel") |>
select(Wage, Educ, FemaleYes=Female) |>
filter(Educ>0) |> 
mutate(Wage=Wage*5.51) |>   
mutate(FemaleYes=factor(FemaleYes)) |> 
mutate(Educ=cut(Educ, breaks = c(0,11,12,99), 
                       labels = c("NoHS","HS","Degree")))
head(DataWage)
```

::: footer
Note, $Wage$ from 1976 has been adjusted to 2024 purchasing power using
<https://www.in2013dollars.com/us/inflation/1976?amount=1>.
:::

# ARE WAGE MEANS DIFFERENT FOR FEMALES AND MALES?

## Are Wage Means Different for Females and Males?

#### Graphical Approach (comparing means)

```{r}
DataWageFem=DataWage |> filter(FemaleYes==1)
MeanWageFem=mean(DataWageFem$Wage)
DataWageMale=DataWage |> filter(FemaleYes==0)
MeanWageMale=mean(DataWageMale$Wage)
ggplot(DataWage, aes(x=FemaleYes, y=Wage))+
   geom_point()+
   geom_point(aes(x=1, y=MeanWageMale), size=7, color="Blue")+
   geom_point(aes(x=2, y=MeanWageFem), size=7, color="Red")+
   scale_y_continuous(breaks = c(seq(0,150,10)))
```

## Are Wage Means Different for Females and Males?

#### Graphical Approach (comparing means)

```{r}
DataWageFem=DataWage |> filter(FemaleYes==1)
MeanWageFem=mean(DataWageFem$Wage)
DataWageMale=DataWage |> filter(FemaleYes==0)
MeanWageMale=mean(DataWageMale$Wage)
ggplot(DataWage, aes(x=FemaleYes, y=Wage))+
   geom_jitter(width=0.07)+
   geom_point(aes(x=1, y=MeanWageMale), size=7, color="Blue")+
   geom_point(aes(x=2, y=MeanWageFem), size=7, color="Red")+
   scale_y_continuous(breaks = c(seq(0,150,10)))
```

## Are Wage Means Different for Females and Males?

#### t-test (comparing means)

```{r echo=TRUE}
DataWageFem= filter(DataWage, FemaleYes==1)
DataWageMale= filter(DataWage, FemaleYes==0)

t.test(DataWageFem$Wage,DataWageMale$Wage, var.equal = TRUE)
```

The difference in means is: `r mean(DataWageFem$Wage)-mean(DataWageMale$Wage)`

## Are Wage Means Different for Females and Males?

#### Graphical Approach (OLS Regression)

::: r-stack
::: fragment
```{r  }
Plot1=ggplot(DataWage, aes(x=FemaleYes, y=Wage))+
   geom_point()+
   geom_point(aes(x=1, y=MeanWageMale), size=7, color="Blue")+
   geom_point(aes(x=2, y=MeanWageFem), size=7, color="Red")+
  scale_y_continuous(breaks = c(seq(0,150,10)))
plot(Plot1)
```
:::

::: fragment
```{r }
Plot1=Plot1+
    geom_segment(aes(x = 1,y = MeanWageMale, 
                     xend = 2,yend = MeanWageFem),                                     color="magenta", linewidth=2.7)
plot(Plot1)
```
:::
:::

## Are Wage Means Different for Females and Males?

#### OLS Regression Approach

```{r echo=TRUE}

ModelSexDis=lm(Wage~FemaleYes, data=DataWage)
summary(ModelSexDis)

```

## Are Wage Means Different for Females and Males? {.smaller}
### Interpretation of Dummy Variables 

```{r echo=TRUE}
print(ModelSexDis)
```

$$
\begin{aligned}
Wage & = \beta_1& \cdot &FemaleYes& + \beta_0 \\
Wage & = (-13.8)& \cdot &FemaleYes& + 39.1 \\
[-13.8] & = (-13.8)& \cdot &[+1]& + [+0] \\
\end{aligned}
$$
$FemaleYes$ can only increase by +1 from 0 to 1 --- when a observation with $FemaleYes=0$ (male) switches to  $FemaleYes=1$ (female). The consequence is that wage changes with -13.8. 

## Are Wage Means Different for Females and Males?

#### ANOVA

::: r-stack
::: fragment
```{r  }
ggplot(DataWage, aes(x=FemaleYes, y=Wage))+
   geom_jitter(width=0.07)+
   geom_point(aes(x=1, y=MeanWageMale), size=7, color="Blue")+
   geom_point(aes(x=2, y=MeanWageFem), size=7, color="Red")+
   geom_segment(aes(x = 1,y = MeanWageMale, 
                     xend = 2,yend = MeanWageFem),                                     color="magenta", linewidth=2.7)+
   scale_y_continuous(breaks = c(seq(0,150,10)))
```
:::

::: fragment
```{r  }
ggplot(DataWage, aes(x=FemaleYes, y=Wage))+
   geom_jitter(width=0.07)+
   geom_point(aes(x=1, y=MeanWageMale), size=7, color="Blue")+
   geom_point(aes(x=2, y=MeanWageFem), size=7, color="Red")+
   geom_segment(aes(x = 1,y = MeanWageMale, 
                     xend = 2,yend = MeanWageFem),                                     color="magenta", linewidth=2.7)+
   scale_y_continuous(breaks = c(seq(0,90,5)), limits=c(10,90))
```
:::

::: fragment
```{r  }
ggplot(DataWage, aes(x=FemaleYes, y=Wage))+
   geom_jitter(width=0.07)+
   geom_point(aes(x=1, y=MeanWageMale), size=7, color="Blue")+
   geom_point(aes(x=2, y=MeanWageFem), size=7, color="Red")+
   geom_segment(aes(x = 1,y = MeanWageMale, 
                     xend = 2,yend = MeanWageFem),                                     color="magenta", linewidth=2.7)+
   geom_hline(yintercept=mean(DataWage$Wage),color="red",size=2)+
   scale_y_continuous(breaks = c(seq(0,90,5)), limits=c(10,90))
```
:::
:::

## Are Wage Means Different for Females and Males? 
#### ANOVA

$$
\begin{aligned}
\mbox{Mean Total Error (Variance): } & MTE \approx \frac{1}{N}
                             \sum_{i=1}^N (y_i - \overline{y})^2 \\
\mbox{Mean Residual Error: } & MSE \approx \frac{1}{N}
                             \sum_{i=1}^N (y_i - \widehat{y})^2
\end{aligned}
$$

F-Value:

$$
F=\frac{MTE}{MSE}= \frac{\mbox{Mean Total Error (Restricted Model)}}{\mbox{Mean Residual Error (Full Model)}}
$$

## Are Wage Means Different for Females and Males?  {.smaller}

#### ANOVA 

$$
F=\frac{MTE}{MSE}= \frac{\mbox{Mean Sum of Errors: Restricted Model}}{\mbox{Mean Sum of Errors: Full Model}}
$$

```{r echo=TRUE}
ModelSexDis=lm(Wage~FemaleYes, data=DataWage)
ANOVATable=anova(ModelSexDis)
print(ANOVATable)
```

$MTE$= `r format(ANOVATable[[1,3]], scientific=FALSE)` +
`r format(ANOVATable[[2,3]], scientific=FALSE)`=
`r format(ANOVATable[[1,3]]+ ANOVATable[[2,3]], scientific=FALSE)`

If F\>1. I.e., full model mean squared error is smaller than the one of
the restricted model. And if this is not by chance ($P$ is very small),
then the variables from the full model must be significant.

## Are Education Differences Relevant for Wages?

#### Data

```{r echo=TRUE}
head(DataWage)
levels(DataWage$Educ)
```

## Are Education Differences Relevant for Wages?

#### (pair wise) t-test

Pair-wise t-test is problematic to indicate if factor is relevant
because of multi-testing problem.

## Are Education Differences Relevant for Wages?

#### One-Way ANOVA (one factor (eduction) with 3 groups)

```{r echo=TRUE}
ModelEdu=lm(Wage~Educ, data=DataWage )
summary(ModelEdu)

ANOVATable=anova(ModelEdu)
print(ANOVATable)
```

$MTE$= `r format(ANOVATable[[1,3]], scientific=FALSE)` +
`r format(ANOVATable[[2,3]], scientific=FALSE)`=
`r format(ANOVATable[[1,3]]+ ANOVATable[[2,3]], scientific=FALSE)`


## Are Education Differences Relevant for Wages?

#### Adjusted Paired t-test


```{r echo=TRUE}
pairwise.t.test(DataWage$Wage,DataWage$Educ, pool.sd=FALSE, p.adjust.method="bonferroni")
```

## AB-Test {.smaller}

A very powerful and easy to use methodology to compare means of one or more *Treatment* groups to a *Control Group*. 


**Goal:** Determine if a treatment(s) (e.g., conversion rates for one or more new websites (*Treatment*)) are significant compared to the old website (*Control Group*).

. . .

**Methodology:** To evaluate significant differences between groups *One-Way ANOVA* possibly followed by a *post-hoc* pairwise t-test, if *ANOVA* was successful and the dependent variable is continuous. For binary dependent variables alternatives to ANOVA exist.For experiments with binary outcome and one control group and one treatment group a t-test can be used.

. . .

**Requirement:** Participants must be randomly assigned to the groups (no self-selection!).

. . .

**Problem:** While it is easy and technically straightforward to randomly assign website visitors to different webpages, it can be difficult or impossible in other cases. 

For example, it is not feasible to assign participants of a marketing event into a treatment group (drank a glass of champagne before the talk) and a control group (did not drink a glass of champagne before the talk)

## Two-Way Anova (not relevant for midterm)
#### With Interaction Term 
```{r echo=TRUE}
Model2WayWage=lm(Wage~Educ*FemaleYes, data=DataWage)
anova(Model2WayWage)
```

## Two-Way Anova (not relevant for midterm)
#### With Interaction Term (again with different order)
```{r echo=TRUE}
Model2WayWage=lm(Wage~FemaleYes*Educ, data=DataWage)
anova(Model2WayWage)
```

## Two-Way Anova (not relevant for midterm)
#### Without Interaction Term 

```{r echo=TRUE}
Model2WayWage=lm(Wage~FemaleYes+Educ, data=DataWage)
anova(Model2WayWage)
```

## Tukey Pairwise t-Test (not relevant for midterm)
#### Without Interaction Term 

```{r echo=TRUE}
Model2WayWageNoInt=lm(Wage~FemaleYes+Educ, data=DataWage)
TukeyHSD(aov(Model2WayWageNoInt))
```