---
title: "Intro to Machine Learning"
subtitle: "Types, Tasks, Terminology"
format: 
  revealjs:
    code-fold: true
---

## Topics {.smaller .scrollable}

```{r}
#| echo: false
#| warning: false
library(tidyverse)
library(rio)
library(janitor)
library(plotly)
```

-   Economic Activity in the United States
    -   Current State
    -   Future Expectations
    -   Discussion
-   Employment
    -   Current and Historic States
    -   Future Expectations
    -   Discussion
-   Inflation, Monetary Policy, and Government Debt in the United States
    -   Current and Historic States
    -   Future Expectations
    -   Discussion
-   Investment Climate --- Interest Rate, Regulations, Uncertainty
    -   Current and Historic States
    -   Future Expectations
    -   Discussion
-   International Trade --- Trade Balance, Exchange Rate, Tariffs, Globalization
    -   Current and Historic States
    -   Future Expectations
    -   Discussion
-   Opportunities
    -   Start Ups, Entrepreneurship, Innovations, AI
-   What does this mean for China and Korea
    -   Discussion

## GDP Growth

```{r}
#| echo: false
DataGDP=import("data/GDPUS.xlsx") |> 
        select(Date, GDP=GDPC1) |> 
        mutate(Date=as_date(Date)) |> 
        mutate(GDPGrowth=((GDP-lag(GDP))/lag(GDP)+1)^4-1) |>          filter(Date>as_date("2000-01-01"))

AvgGDPGrowth=mean(DataGDP$GDPGrowth, na.rm = TRUE)

PlotGDP=ggplot(DataGDP, aes(y=GDPGrowth, x=Date))+
geom_line(aes(color="Growth Rate"))+
scale_y_continuous(labels = scales::percent, 
                     breaks = seq(-0.3, 0.3, 0.05)) +
scale_x_date(date_breaks = "2 years", date_labels = "%Y") + 
labs(title = "GDP Growth in the USA", 
     x = "Date", y = "Real GDP Growth(in %)")  

ggplotly(PlotGDP)
```

## GDP Growth

```{r}
PlotGDPAdj=PlotGDP+ 
  scale_y_continuous(labels = scales::percent,
                     breaks = seq(-0.1, 0.10, 0.01))+ 
geom_hline(aes(yintercept = AvgGDPGrowth, 
               color = "Average"))+
geom_smooth(aes(color = "Smoothed Growth"),method = "loess",              span=0.2, se=FALSE)+
scale_color_manual(name = "Legend", 
                   values = c("Growth Rate" = "red",   
                              "Average" = "blue", 
                              "Smoothed Growth" = "green"))+
coord_cartesian(ylim = c(-0.07,0.07))

ggplotly(PlotGDPAdj)
```

## Employment

```{r}
#| echo: false
#| warning: false
DataUnempl=import("data/UnemploymentUS.xlsx") |>
           select(Date,UnRate=UNRATE) |> 
           mutate(Date=as_date(Date))|>                 
           filter(Date>as_date("2000-01-01"))

PlotUN=ggplot(DataUnempl,aes(x=Date, y=UnRate)) +
       geom_line(color="red")+
scale_x_date(date_breaks = "3 years", date_labels = "%Y") + 
labs(title = "Unemployment in the USA", 
     x = "Date", y = "Unemployment Rate(in %)")  

ggplotly(PlotUN)
```

## Furture Outlook for the U.S Economy

???

. . .

![](PapaSmurf.mp4){width="480" height="360" }

## Your questions