---
title: Scientific Notation
subtitle: "How are Very Small and Very Big Numbers Displayed in R and Other Statistics Apps?" 
author:
  name: "Carsten Lange"
  email: "clange@cpp.edu"
  affiliation: "Cal Poly, Pomona"

format: 
  revealjs:
    echo: true
    code-fold: true
    output-location: fragment
    incremental: false
    warning: false
    message: false
    navigation-mode: vertical
    scrollable: true
    smaller: true
    footer: "<a href='https://ai.lange-analytics.com/htmlbook/RAndRStudio.html'>Textbook </a> "
    
include-in-header:
  text: |
    <style>
    .reveal .slides section:not(:first-of-type) h1,
    .reveal .slides section:not(:first-of-type) h2,
    .reveal .slides section:not(:first-of-type) h3,
    .reveal .slides section:not(:first-of-type) h4,
    .reveal .slides section:not(:first-of-type) h5,
    .reveal .slides section:not(:first-of-type) h6 {
      font-size: 142% !important;
    }
    </style>
---

## Learning Outcomes

-  

## How are Very Big Numbers Presented {.smaller}


The GDP for 2021 in the US was $ 22,996,086,000,000 (rounded to millions)

\begin{eqnarray*}
GDP&=&2.2996086 \cdot 10000000000000\\
&\Longleftrightarrow&\\
GDP&=&2.2996086 \cdot 10\cdot 10\cdot 10\cdot 10\cdot 10\cdot 10\cdot 10\cdot 10\cdot 10\cdot 10\cdot 10\cdot 10\cdot 10 \\
&\Longleftrightarrow&\\
GDP&=&2.2996086 \cdot 10^{13}
\end{eqnarray*}

Let us see what R does:

```{r}
#| echo: true
GDPUS=22996086000000
print(GDPUS)

```




## How are Very Small Numbers Presented 🤓 {.smaller}

::: {.panel-tabset}

### Main 

The probability of getting struck by lightning in the US is about $0.000000000365$ on any randomly chosen day.

\begin{eqnarray*}
ProbLight&=&\frac{3.65}{10000000000} \\
&\Longleftrightarrow&\\
ProbLight&=&\frac{3.65}{10\cdot 10\cdot 10\cdot 10\cdot 10\cdot                              10\cdot 10\cdot 10\cdot 10\cdot 10} \\
&\Longleftrightarrow&\\
ProbLight&=&\frac{3.65}{10^{10}}\\
&\Longleftrightarrow&\\
ProbLight&=&3.65 \cdot 10^{-10}
\end{eqnarray*}


### Why?

In the U.S., a person has a 1:10,000-lifetime risk of being struck by lightning. Assuming a life span of 75 years and 365.25 days per year, the probability per day is:

$$\frac{1}{10,000 \cdot 365.25\cdot 75)}=0.000000000365$$
### What R Does  

Let us see what R does:

```{r}
#| echo: true
ProbStruck=0.000000000365
cat("Probabilty to get stuck by a lighning on an avg. day:", ProbStruck)
```

:::

::: footer
See: [RandRStudioScriptPart1.R script400](https://econ.lange-analytics.com/RScripts/RAndRStudioScriptPart1.R){target="_blank"}
:::